### 2024-03-27 paper/paper.qmd
Update gitignore, update readme, get rid of extra examples on paper, update scripts, delete replication file

---
title: "From Theory to Practice: Investigating Treatment Responses in Breast Cancer"
subtitle: "My subtitle if needed"
author: 
  - First author
thanks: "Code and data are available at: LINK."
date: today
date-format: long
abstract: "First sentence. Second sentence. Third sentence. Fourth sentence."
format: pdf
number-sections: true
bibliography: references.bib
---

```{r}
#| include: false
#| warning: false
#| message: false

library(tidyverse)
library(palmerpenguins)
```

# Introduction

You can and should cross-reference sections and sub-sections. We use @citeR and @rohan.

The remainder of this paper is structured as follows. @sec-data....

# Data {#sec-data}

Some of our data is of penguins (@fig-bills), from @palmerpenguins.

```{r}
#| label: fig-bills
#| fig-cap: Bills of penguins
#| echo: false

```

Talk more about it.

And also planes (@fig-planes). (You can change the height and width, but don't worry about doing that until you have finished every other aspect of the paper - Quarto will try to make it look nice and the defaults usually work well once you have enough text.)

```{r}
#| label: fig-planes
#| fig-cap: Relationship between wing length and width
#| echo: false
#| warning: false
#| message: false

```

Talk way more about it.

# Model

The goal of our modelling strategy is twofold. Firstly,...

Here we briefly describe the Bayesian analysis model used to investigate... Background details and diagnostics are included in [Appendix -@sec-model-details].

## Model set-up

Define $y_i$ as the number of seconds that the plane remained aloft. Then $\beta_i$ is the wing width and $\gamma_i$ is the wing length, both measured in millimeters.

\begin{align} 
y_i|\mu_i, \sigma &\sim \mbox{Normal}(\mu_i, \sigma) \\
\mu_i &= \alpha + \beta_i + \gamma_i\\
\alpha &\sim \mbox{Normal}(0, 2.5) \\
\beta &\sim \mbox{Normal}(0, 2.5) \\
\gamma &\sim \mbox{Normal}(0, 2.5) \\
\sigma &\sim \mbox{Exponential}(1)
\end{align}

We run the model in R [@citeR] using the `rstanarm` package of @rstanarm. We use the default priors from `rstanarm`.

### Model justification

We expect a positive relationship between the size of the wings and time spent aloft. In particular...

We can use maths by including latex between dollar signs, for instance $\theta$.

# Results

Our results are summarized in @tbl-modelresults.

```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false

```

```{r}
#| echo: false
#| eval: true
#| label: tbl-modelresults
#| tbl-cap: "Explanatory models of flight time based on wing width and wing length"
#| warning: false

```

# Discussion

## First discussion point {#sec-first-point}

If my paper were 10 pages, then should be be at least 2.5 pages. The discussion is a chance to show off what you know and what you learnt from all this.

## Second discussion point

## Third discussion point

## Weaknesses and next steps

Weaknesses and next steps should also be included.

\newpage

\appendix

# Appendix {-}

# Additional data details

# Model details {#sec-model-details}
rior-2 we compare the posterior with the prior. This shows...

```{r}
#| eval: true
#| echo: false
#| message: false
#| warning: false
#| label: fig-ppcheckandposteriorvsprior
#| layout-ncol: 2
#| fig-cap: "Examining how the model fits, and is affected by, the data"
#| fig-subcap: ["Posterior prediction check", "Comparing the posterior with the prior"]

```

## Diagnostics

```{r}
#| echo: false
#| eval: true
#| message: false
#| warning: false
#| label: fig-stanareyouokay
#| fig-cap: "Checking the convergence of the MCMC algorithm"
#| fig-subcap: ["Trace plot", "Rhat plot"]
#| layout-ncol: 2

```

\newpage

# References

### 2024-03-27 scripts/00-simulate_data.R

@@ -1,19 +1,33 @@
 #### Preamble ####
-# Purpose: Simulates... [...UPDATE THIS...]
-# Author: Rohan Alexander [...UPDATE THIS...]
-# Date: 11 February 2023 [...UPDATE THIS...]
-# Contact: rohan.alexander@utoronto.ca [...UPDATE THIS...]
+# Purpose:Stimulates the clean dataset used for the model
+# Author: Lexi Knight
+# Date: 26 March 2024
+# Contact: lexi.knight@mail.utoronto.ca
 # License: MIT
-# Pre-requisites: [...UPDATE THIS...]
-# Any other information needed? [...UPDATE THIS...]
+# Pre-requisites: -
+# Any other information needed? -
 
 
 #### Workspace setup ####
+# install packages 
+# install.packages("tidyverse")
+
+# load packages
 library(tidyverse)
-# [...UPDATE THIS...]
+library(testthat)
+
+
 
 #### Simulate data ####
-# [...ADD CODE HERE...]
 
 
-
+data <- 
+  tibble( 
+    level = rnorm(n = 1000, mean = 100, sd = 10) |> floor()
+    weakness = sample(x = c("fire", "not fire"), size = 1000, replace = TRUE)
+    
+    
+    
+    
+    
+    

### 2024-03-28 scripts/00-simulate_data.R
simulated the data and started writing some tests

@@ -10,24 +10,69 @@
 
 #### Workspace setup ####
 # install packages 
-# install.packages("tidyverse")
+install.packages("tidyverse")
+install.packages("conflicted")
+install.packages("dplyr")
+install.packages("magrittr")
 
 # load packages
 library(tidyverse)
+library(conflicted)
 library(testthat)
-
+library(dplyr)
+library(magrittr)
 
 
 #### Simulate data ####
 
+# Set seed for reproducibility
+set.seed(0)
 
-data <- 
-  tibble( 
-    level = rnorm(n = 1000, mean = 100, sd = 10) |> floor()
-    weakness = sample(x = c("fire", "not fire"), size = 1000, replace = TRUE)
-    
-    
-    
-    
-    
-    
+# Number of entries
+num_entries <- 981
+
+# Simulating the data 
+simulated_data <- data.frame(
+  ajcc_pathologic_n = sample(c("N0", "N1", "N2", "NX"), num_entries, replace = TRUE),
+  ajcc_pathologic_m = sample(c("M0", "MX", "M1a", "M1b"), num_entries, replace = TRUE),
+  ajcc_pathologic_t = sample(c("T1", "T1a", "T1b", "T2", "T2a", "T2b", "T3", "T4", "TX"), num_entries, replace = TRUE),
+  age_at_diagnosis = round(runif(num_entries, min = 14681, max = 32872)),
+  days_to_death = round(runif(num_entries, min = 0, max = 6236)),
+  race = sample(c("not reported", "white", "american indian or alaska native", "asian", "black or african american"), num_entries, replace = TRUE),
+  ajcc_pathologic_stage = sample(c("Stage IA", "Stage IIA", "Stage IIIA", "Stage IB", "Stage IIB", "Stage IV", "Stage I"), num_entries, replace = TRUE),
+  treatment_type = sample(c("Pharmaceutical Therapy, NOS", "Radiation Therapy, NOS"), num_entries, replace = TRUE),
+  treatment = sample(c("yes", "no", "not reported"), num_entries, replace = TRUE)
+) 
+
+
+# Renaming variables for clarity - not sure if i should do this here or later ****
+simulated_data <- simulated_data %>%
+  rename(
+    Pathogenic_Stage = ajcc_pathologic_stage,
+    Lymph_Node_Involvement = ajcc_pathologic_n,
+    Presence_of_Metastasis = ajcc_pathologic_m,
[…]

### 2024-03-28 scripts/03-test_data.R

@@ -1,8 +1,8 @@
 #### Preamble ####
 # Purpose: Tests... [...UPDATE THIS...]
-# Author: Rohan Alexander [...UPDATE THIS...]
-# Date: 11 February 2023 [...UPDATE THIS...]
-# Contact: rohan.alexander@utoronto.ca [...UPDATE THIS...]
+# Author: Lexi Knight
+# Date: 28 March 2024
+# Contact: lexi.knight@mail.utoronto.ca
 # License: MIT
 # Pre-requisites: [...UPDATE THIS...]
 # Any other information needed? [...UPDATE THIS...]
@@ -10,6 +10,59 @@
 
 #### Workspace setup ####
 library(tidyverse)
-# [...UPDATE THIS...]
+
 
 #### Test data ####
+
+
+
+# Testing the simulated table - * i think i should just move this to the testing file ****
+
+# Test if the dataset has 981 entries 
+test_that("Dataset has 981 entries", {
+  expect_equal(nrow(simualted_data), 981)
+})
+
+# Test if 'Pathogenic_Stage' contains the correct categories 
+test_that("Pathogenic stage variable is correct", {
+  expect_true(all(simulated_data$Pathogenic_Stage %in% c('Stage IA', 'Stage IIA', 'Stage IIIA', 'Stage IB', 'Stage IIB', 'Stage IV', 'Stage I')))
+})
+
+# Test if 'Lymph_Node_Involvement' contains the correct categories 
+test_that("Lymph node involvement variable is correct", {
+  expect_true(all(simulated_data$Lymph_Node_Involvement %in% c('N0', 'N1', 'N2', 'NX')))
+})
+
+# Test if 'Presence_of_Metastasis' contains the correct categories 
+test_that("Presence of metastasis variable is correct", {
+  expect_true(all(simulated_data$Presence_of_Metastasis %in% c('M0', 'MX', 'M1a', 'M1b')))
+})
+
+# Test if 'Tumor_Size' contains the correct categories 
+test_that("Tumor size variable is correct", {
+  expect_true(all(simulated_data$Tumor_Size %in% c('T1', 'T1a', 'T1b', 'T2', 'T2', 'T2b', 'T3', 'T4', 'TX')))
+})
+
+# Test if 'Age_at_Diagnosis_in_Days is numeric 
+class(simulated_data$Age_at_Diagnosis_in_Days) == "numeric"
+
+# Test if 'Days_to_Death' is numeric 
+class(simulated_data$Days_to_Death) == "numeric"
+
[…]

### 2024-03-28 scripts/00-simulate_data.R
updated simulation

@@ -4,23 +4,16 @@
 # Date: 26 March 2024
 # Contact: lexi.knight@mail.utoronto.ca
 # License: MIT
-# Pre-requisites: -
-# Any other information needed? -
+# Pre-requisites: N/A
 
 
 #### Workspace setup ####
+
 # install packages 
-install.packages("tidyverse")
-install.packages("conflicted")
-install.packages("dplyr")
-install.packages("magrittr")
+# install.packages("tidyverse")
 
 # load packages
 library(tidyverse)
-library(conflicted)
-library(testthat)
-library(dplyr)
-library(magrittr)
 
 
 #### Simulate data ####
@@ -45,7 +38,7 @@ simulated_data <- data.frame(
 ) 
 
 
-# Renaming variables for clarity - not sure if i should do this here or later ****
+# Renaming variables for clarity 
 simulated_data <- simulated_data %>%
   rename(
     Pathogenic_Stage = ajcc_pathologic_stage,
@@ -58,7 +51,6 @@ simulated_data <- simulated_data %>%
     Treatment_Type = treatment_type,
     Treatment_Decision = treatment
   )
-
 
 
 simulated_data <- simulated_data |> select( Pathogenic_Stage,

### 2024-03-28 scripts/01-download_data.R
update packages

@@ -1,26 +1,39 @@
 #### Preamble ####
-# Purpose: Downloads and saves the data from [...UPDATE THIS...]
-# Author: Rohan Alexander [...UPDATE THIS...]
-# Date: 11 February 2023 [...UPDATE THIS...]
-# Contact: rohan.alexander@utoronto.ca [...UPDATE THIS...]
+# Purpose: Downloads and saves the lung cancer dataset from the OSF dataverse
+# database with the variables of interest. 
+# Author: Lexi Knight
+# Date: 28 March 2024
+# Contact: Lexi.Knight@mail.utoronto.ca
 # License: MIT
-# Pre-requisites: [...UPDATE THIS...]
-# Any other information needed? [...UPDATE THIS...]
+# Pre-requisites: -
+
 
 
 #### Workspace setup ####
-library(opendatatoronto)
+# install packages 
+
+#load packages 
+
 library(tidyverse)
-# [...UPDATE THIS...]
+library(dataverse)
+library(gutenbergr)
+library(janitor)
+library(knitr)
+library(readr)
 
 #### Download data ####
-# [...ADD CODE HERE TO DOWNLOAD...]
 
+lung_cancer_data <- get_dataframe_by_name(
+    filename = "Lung_Cancer_Meta.csv"
+    dataset = "https://osf.io/qk34b/"
+    server = "open.science.framework.io"
+    .f = read_csv
+  ) 
 
 
 #### Save data ####
-# [...UPDATE THIS...]
-# change the_raw_data to whatever name you assigned when you downloaded it.
-write_csv(the_raw_data, "inputs/data/raw_data.csv") 
 
-         
+# change the_raw_data to Lung_Cancer_raw_dataset.csv
+write_csv(lung_cancer_data, "inputs/data/raw_data/Lung_Cancer_raw_dataset.csv") 
+
+      

### 2024-03-28 scripts/03-test_data.R

@@ -1,26 +1,29 @@
 #### Preamble ####
-# Purpose: Tests... [...UPDATE THIS...]
+# Purpose: Tests the cleaned dataset
 # Author: Lexi Knight
 # Date: 28 March 2024
 # Contact: lexi.knight@mail.utoronto.ca
 # License: MIT
-# Pre-requisites: [...UPDATE THIS...]
-# Any other information needed? [...UPDATE THIS...]
+# Pre-requisites: run 01-download_data.R as well as 02-data_cleaning.R in scripts in order to obtain raw and cleaned data. 
 
 
 #### Workspace setup ####
+
+#install packages 
+install.packages("tidyverse")
+install.packages("testthat")
+
+# load packages
 library(tidyverse)
+library(testthat)
 
 
 #### Test data ####
 
 
-
-# Testing the simulated table - * i think i should just move this to the testing file ****
-
 # Test if the dataset has 981 entries 
 test_that("Dataset has 981 entries", {
-  expect_equal(nrow(simualted_data), 981)
+  expect_equal(nrow(simulated_data), 981)
 })
 
 # Test if 'Pathogenic_Stage' contains the correct categories 
@@ -44,19 +47,21 @@ test_that("Tumor size variable is correct", {
 })
 
 # Test if 'Age_at_Diagnosis_in_Days is numeric 
-class(simulated_data$Age_at_Diagnosis_in_Days) == "numeric"
+test_that("Age at diagnosis variable is numeric", {
+  expect_true(is.numeric(simulated_data$Age_at_Diagnosis_in_Days))
+})
 
 # Test if 'Days_to_Death' is numeric 
 class(simulated_data$Days_to_Death) == "numeric"
 
 # Test if 'Race' contains the correct categories 
 test_that("Race variable is correct", {
-  expect_true(all(simulated_data$Race %in% c('not reported", "white", "american indian or alaska native", "asian", "black or african american')))
+  expect_true(all(simulated_data$Race %in% c('not reported', 'white','american indian or alaska native', 'asian', 'black or african american')))
 })
 
 # Test if 'Treatment_Type' only contains 'Pharmaceutical Therapy, NOS' and 'Radiation Therapy, NOS'
 test_that("Treatment type variable is correct", {
-  expect_true(all(simulated_dat$Treatment_Type %in% c('Pharmaceutical Therapy, NOS', 'Radiation Therapy, NOS')))
[…]

### 2024-03-30 paper/paper.qmd
Used ChatGPT to create interesting hook in the introduction and to come up with a creative and unique title. Thus updated llm/usage

@@
 ---
-title: "From Theory to Practice: Investigating Treatment Responses in Breast Cancer"
+title: "Survival Compass: Statistical Insights into Lung Cancer Patients' Journey Post-Diagnosis"
 subtitle: "My subtitle if needed"

### 2024-04-01 paper/paper.qmd
Edited author, code and added an abstract. Added llm for abstract.

--- 2024-03-27 paper/paper.qmd
@@
 ---
-title: "From Theory to Practice: Investigating Treatment Responses in Breast Cancer"
-subtitle: "My subtitle if needed"
+title: "Survival Compass: Statistical Insights into Lung Cancer Patients' Journey Post-Diagnosis"
 author: 
-  - First author
-thanks: "Code and data are available at: LINK."
+  - Lexi Knight
+thanks: "Code and data are available at: https://github.com/LexiKnight/Lung_Cancer/tree/main"
 date: today
 date-format: long
-abstract: "First sentence. Second sentence. Third sentence. Fourth sentence."
+abstract: "This study investigates the impact of pathogenic stage and treatment modalities on lung cancer survival post-diagnosis.
+Analysis of patient data reveals significant correlation between pathogenic stage, seeking treatment and survival outcomes.
+Notably, patients at advanced stages with metastases in distant sites beyond the lung, extensive lymph node involvement and tumors with extensive growth, invading nearby structures demonstrate lower survival rates.
+These findings underscore the critical importance of early detection, tailored treatment strategies and ongoing research efforts to enhance lung cancer survival rates globally."
 format: pdf
@@
 
+# install packages
+install.packages("knitr")
+install.packages("kableExtra")
+
+# load packages
 library(tidyverse)
-library(palmerpenguins)
+library(knitr)
+library(kableExtra)
+
 ```

### 2024-04-02 scripts/00-simulate_data.R
remove race from simulation

@@ -4,16 +4,17 @@
 # Date: 26 March 2024
 # Contact: lexi.knight@mail.utoronto.ca
 # License: MIT
-# Pre-requisites: N/A
+# Pre-requisites: -
+# Any other information needed? -
 
 
 #### Workspace setup ####
-
 # install packages 
 # install.packages("tidyverse")
 
 # load packages
 library(tidyverse)
+
 
 
 #### Simulate data ####
@@ -31,7 +32,6 @@ simulated_data <- data.frame(
   ajcc_pathologic_t = sample(c("T1", "T1a", "T1b", "T2", "T2a", "T2b", "T3", "T4", "TX"), num_entries, replace = TRUE),
   age_at_diagnosis = round(runif(num_entries, min = 14681, max = 32872)),
   days_to_death = round(runif(num_entries, min = 0, max = 6236)),
-  race = sample(c("not reported", "white", "american indian or alaska native", "asian", "black or african american"), num_entries, replace = TRUE),
   ajcc_pathologic_stage = sample(c("Stage IA", "Stage IIA", "Stage IIIA", "Stage IB", "Stage IIB", "Stage IV", "Stage I"), num_entries, replace = TRUE),
   treatment_type = sample(c("Pharmaceutical Therapy, NOS", "Radiation Therapy, NOS"), num_entries, replace = TRUE),
   treatment = sample(c("yes", "no", "not reported"), num_entries, replace = TRUE)
@@ -47,7 +47,6 @@ simulated_data <- simulated_data %>%
     Tumor_Size = ajcc_pathologic_t,
     Age_at_Diagnosis_in_Days = age_at_diagnosis,
     Days_to_Death = days_to_death,
-    Race = race,
     Treatment_Type = treatment_type,
     Treatment_Decision = treatment
   )
@@ -57,7 +56,7 @@ simulated_data <- simulated_data |> select( Pathogenic_Stage,
                                             Lymph_Node_Involvement,
                                             Presence_of_Metastasis, Tumor_Size,
                                             Age_at_Diagnosis_in_Days,
-                                            Days_to_Death, Race, Treatment_Type,
+                                            Days_to_Death, Treatment_Type,
                                             Treatment_Decision)
 
 
@@ -67,4 +66,3 @@ head(simulated_data)
 
 # Show a summary of the data
 summary(simulated_data)
-

### 2024-04-02 paper/paper.qmd
Updated readme - the overview section. Added introduction and a bit about the model

@@
 # install packages
-install.packages("knitr")
-install.packages("kableExtra")
+# install.packages("knitr")
+# install.packages("kableExtra")
 
 # load packages
-library(tidyverse)
-library(knitr)
-library(kableExtra)
+# library(tidyverse)
+# library(knitr)
+# library(kableExtra)
 
@@
 
-You can and should cross-reference sections and sub-sections. We use @citeR and @rohan.
+Clinging to life amidst the shadows of lung cancer, where every breath becomes a battleground.
+Survival becomes not just a statistic but an interplay between several individual characteristics.
+We explore the hidden keys to defying the odds and emerging victorious against one of the deadliest adversaries of our time.
+Lung cancer is the leading cause of cancer-related deaths in the world (Park, 2017).
+It is a disease that develops in the lining of the airways in lung tissues.
+Non-small cell lung cancer (NSCLC) is the most common type, accounting for  80-85% of all lung cancers according to the American Cancer society (Markman, 2023).
+Staging is important for prognosis and making treatment decisions.
+Common treatments include surgery, radiation therapy and chemotherapy (Kai, 2021).
+Pathogenic stage is determined by presence of nearby metastasis, lymph node involvement as well as tumor spread and size (Markman, 2023).
+This paper investigates the relationship between lung cancer patients' survival and pathogenic stage.
+The estimand is the median survival time in days post-diagnosis.
+We also look at whether patients decided to have treatment and if so, which method; radiation therapy or chemotherapy.
+Through analysis of a dataset made up of  981 patients in Sydney, Australia, we offer insight into the prognostic markers.
 
-The remainder of this paper is structured as follows. @sec-data....
+Tumor size is often the main determinant of stage and treatment.
+As tumor categories increase, the tumor expands, invading nearby structures (Zhang, 2015).
+A study involving 52,287 patients diagnosed between the years 1998 and 2003 found tumor size to be an independent prognostic factor in estimating overall survival.
+The authors found that patients presenting with larger tumors predicted a worse prognosis and thus are associated with a decrease in survival.
+There is a similar relationship between extensive lymph node involvement and patient survival (Zhang, 2015).
+Initial spread of cancer cells are localized, then become regional, involving nearby lymph nodes and the most severe cases comprises expansion to other organs such as the brain, liver and bones (Markman, 2023).
+A study looked at five year survival rates based on the severity of spread.
+62.8% of patients with localized spread, 34.8% of patients with regional and 8% of patients with distant, advanced spread were found to survive for 5 years post diagnosis.
+More than half of these lung cancer patients have advanced spread to other organs when diagnosed(Markman, 2023) .
+Overall, it is found that patients with no regional lymph node metastases, and smaller tumors are easier to be treated and thus are associated with improved survival rates (Zhang, 2015).
+
+Presence of metastatic LN is one of the most important determinants of prognosis of NSCLC cases (Kai, 2021).
+In the early stage, cancer has not spread to lymph nodes.
+As severity increases, lymph node metastasis sequentially spreads to more distant lymph nodes such as mediastinal and there is severe lymph node involvement  (Park, 2017).
+Lymph node involvement, also termed lymph node ratio, is a crucial factor in guiding treatment options (Kai, 2021).
+A study made up of 97 patients with a mean age of 63 who have undergone surgery between the years 2009 and 2015 in Korea find that increased lymph node involvement is associated with a more advanced disease status and hence affiliated with prognosis (Park, 2017).
+Another study looked at 11,341 NSCLC patients between the years 2004 to 2015, from 18 geographically diverse populations, covering approximately 28% of the population of the United States.
+These patients were treatment naive and underwent surgical resection of the tumor.
+Although 5757 patients died, the rest showed great results, with a median survival of 22 months (Kai, 2021).
+The authors found that patients with low lymph node involvement lead to higher survival compared to patients with high lymph node ratios.
+A regression analysis revealed that lymph node ratio is an independent and significant predictor of patient survival.
+The authors also observed that disease burden and anatomical location of the lymph nodes involved may influence the patients survival (Kai, 2021).
+
+Presence of metastatic LN is one of the most important determinants of prognosis of NSCLC cases (Kai, 2021).
+In the early stage, cancer has not spread to lymph nodes.
+As severity increases, lymph node metastasis sequentially spreads to more distant lymph nodes such as mediastinal and there is severe lymph node involvement  (Park, 2017).
+Lymph node involvement, also termed lymph node ratio, is a crucial factor in guiding treatment options (Kai, 2021).
+A study made up of 97 patients with a mean age of 63 who have undergone surgery between the years 2009 and 2015 in Korea find that increased lymph node involvement is associated with a more advanced disease status and hence affiliated with prognosis (Park, 2017).
+
+Another study looked at 11,341 NSCLC patients between the years 2004 to 2015, from 18 geographically diverse populations, covering approximately 28% of the population of the United States.
+These patients were treatment naive and underwent surgical resection of the tumor.
+Although 5757 patients died, the rest showed great results, with a median survival of 22 months (Kai, 2021).
+The authors found that patients with low lymph node involvement lead to higher survival compared to patients with high lymph node ratios.
+A regression analysis revealed that lymph node ratio is an independent and significant predictor of patient survival.
+The authors also observed that disease burden and anatomical location of the lymph nodes involved may influence the patients survival (Kai, 2021).
+
+After tumor size, LN involvement and presence of distant metastasis are categorized, the pathogenic stage of the cancer is then determined (Eldridge, 2022).
+The most valuable prognostic factor in non-small cell lung cancer is the pathogenic stage (Park, 2017) .Stage is determined by tumor size, number of tumors  and where the cancer has spread.
+Stage 1 is localized spread, stage 2 and 3 is regional spread while stage 4 is distant spread of the tumor (Eldridge, 2022).
+Cancer stage was determined using the seventh American Joint Committee on Cancer staging system (AJCC) (Park, 2017).
+A study done in Australia including 2119 lung cancer patients illustrated those with stage IV disease, the most advanced stage,  showed shorter survival than those at lower stages (Denton, 2016).
+The earlier the cancer is found, that is the lower the pathogenic stage, the greater the likelihood curative radiation therapy is an effective treatment (Eldridge, 2022).
+
+However, there is minimal literature looking at post-diagnosis survival rates based on pathogenic stage and method of treatment.
+The extent of this disease illustrates the importance of living a healthy lifestyle, undergoing regular screening and development of improved treatment methods.
+Over the past decade, there has been great improvement of lymph node assessment in cancer patients (Kai, 2021).
+Experts hope survival rates continue to improve with  new therapies and treatment approaches  (Markman, 2023).
+
+The remainder of this paper is structured as follows.
+In @sec-data, we visualize the exploration of variables constituting the pathogenic stage and treatment types.
+@sec-model, outlines the model employed to analyze the relationship between these variables and the duration of survival post-diagnosis.
+Moreover, @sec-results offers visual depictions of the study's outcomes.
+Finally, in @sec-discussion, we summarize the primary findings, propose avenues for enhancement, and identify potential areas for future research.
 
@@
 
-Some of our data is of penguins (@fig-bills), from @palmerpenguins.
+Our data is
 
@@
 Talk more about it.
-
-And also planes (@fig-planes).
-(You can change the height and width, but don't worry about doing that until you have finished every other aspect of the paper - Quarto will try to make it look nice and the defaults usually work well once you have enough text.)
 
@@
 
-# Model
+# Model {#sec-model}
 
-The goal of our modelling strategy is twofold. Firstly,...
+The goal of our modeling strategy is twofold.
+Firstly, we aim to investigate the post-diagnosis survival outcomes of lung cancer patients, considering the pathogenic stage.
+Secondly, we seek to assess the efficacy of two common treatment modalities, radiation therapy and chemotherapy, in improving survival chances across different pathogenic stages.
 
-Here we briefly describe the Bayesian analysis model used to investigate... Background details and diagnostics are included in [Appendix -@sec-model-details].
+Here, we briefly describe the Bayesian analysis model used to investigate survival outcomes in lung cancer patients based on pathogenic stage and treatment modality.
+Detailed background information and diagnostic assessments are provided in the [Appendix -@].
+
+Our Bayesian analysis model is tailored to capture the complex relationships between pathogenic stage, treatment modality, and post-diagnosis survival outcomes in lung cancer patients.
+The model accounts for the multifaceted nature of cancer progression, incorporating tumor size, metastatic spread, and lymph node involvement to characterize pathogenic stage.
 
@@
 
-# Results
+The Bayesian analysis model serves as a powerful tool for investigating post-diagnosis survival outcomes in lung cancer patients.
+By incorporating pathogenic stage information and treatment modalities into the analysis, we aim to elucidate the factors influencing patient prognosis and provide insights to guide personalized treatment decisions.
 
-Our results are summarized in @tbl-modelresults.
+# Results {#sec-results}
+
+Our results are summarized in @.
 
@@
 
-# Discussion
+# Discussion {#sec-discussion}
 
 ## First discussion point {#sec-first-point}
-
-If my paper were 10 pages, then should be be at least 2.5 pages. The discussion is a chance to show off what you know and what you learnt from all this.
 
@@
 
-Weaknesses and next steps should also be included.
-
 \newpage
@@
 
-# Appendix {-}
+# Appendix {#sec-appendix}
 
@@
 # Model details {#sec-model-details}
-rior-2 we compare the posterior with the prior. This shows...
+we compare the posterior with the prior. This shows...
 
@@
 ## Diagnostics
+Is this needed?
 

### 2024-04-03 scripts/01-download_data.R
Update steps for accesssing data in README. Write download steps in scripts.

@@ -8,32 +8,12 @@
 # Pre-requisites: -
 
 
+### Steps to Download data ###
 
-#### Workspace setup ####
-# install packages 
-
-#load packages 
-
-library(tidyverse)
-library(dataverse)
-library(gutenbergr)
-library(janitor)
-library(knitr)
-library(readr)
-
-#### Download data ####
-
-lung_cancer_data <- get_dataframe_by_name(
-    filename = "Lung_Cancer_Meta.csv"
-    dataset = "https://osf.io/qk34b/"
-    server = "open.science.framework.io"
-    .f = read_csv
-  ) 
-
-
-#### Save data ####
-
-# change the_raw_data to Lung_Cancer_raw_dataset.csv
-write_csv(lung_cancer_data, "inputs/data/raw_data/Lung_Cancer_raw_dataset.csv") 
-
-      
+# 1. Go to https://osf.io/search?activeFilters=%5B%5D&sort=-relevance&view_only=
+# 2. Type in search bar "Lung Cancer"
+# 3. Select the project "Lung cancer Predictions" by Sicong Chen(Magic)
+# 4. Under files, click  plus sign on "OSF Storage (Australia - Sydney) to expand files 
+# 5. Click "Lung_Cancer_Meta.xlsx" 
+# 6. There are two tabs containing two differnt datasets, select "clinical_info" 
+# and click download the dataset as an excel file. 

### 2024-04-03 scripts/02-data_cleaning.R
completed data cleaning and associated llm usage documented

@@ -8,37 +8,57 @@
 # Any other information needed? [...UPDATE THIS...]
 
 #### Workspace setup ####
+# install packages
+install.packages("readr")
+
+# load libraries 
 library(tidyverse)
+library(readr)
 
 #### Clean data ####
-raw_data <- read_csv("inputs/data/plane_data.csv")
 
-cleaned_data <-
-  raw_data |>
-  janitor::clean_names() |>
-  select(wing_width_mm, wing_length_mm, flying_time_sec_first_timer) |>
-  filter(wing_width_mm != "caw") |>
-  mutate(
-    flying_time_sec_first_timer = if_else(flying_time_sec_first_timer == "1,35",
-                                   "1.35",
-                                   flying_time_sec_first_timer)
-  ) |>
-  mutate(wing_width_mm = if_else(wing_width_mm == "490",
-                                 "49",
-                                 wing_width_mm)) |>
-  mutate(wing_width_mm = if_else(wing_width_mm == "6",
-                                 "60",
-                                 wing_width_mm)) |>
-  mutate(
-    wing_width_mm = as.numeric(wing_width_mm),
-    wing_length_mm = as.numeric(wing_length_mm),
-    flying_time_sec_first_timer = as.numeric(flying_time_sec_first_timer)
-  ) |>
-  rename(flying_time = flying_time_sec_first_timer,
-         width = wing_width_mm,
-         length = wing_length_mm
-         ) |> 
-  tidyr::drop_na()
+# read the csv file 
+raw_data <- read_csv("data/raw_data/lung_cancer_raw.csv")
+
+# check column names of raw_data
+colnames(raw_data)
+
+# select columns of interest 
+cleaned_data <- raw_data |>
+  select("days_to_death", "age_at_diagnosis", "ajcc_pathologic_m",
+         "ajcc_pathologic_n", "ajcc_pathologic_stage", "ajcc_pathologic_t",
+         "treatment_or_therapy", "treatment_type")
+
+# check data types of columns 
+str(cleaned_data)
+
+# handle missing or non-numeric values
+cleaned_data <- cleaned_data %>%
+  filter(!is.na(days_to_death), !is.na(age_at_diagnosis)) %>%
+  mutate_at(vars(days_to_death, age_at_diagnosis), as.numeric)
+
+# create new column for survival_post_diagnosis
+cleaned_data <- cleaned_data %>%
[…]

### 2024-04-03 scripts/00-simulate_data.R
tidied simulation, download and testing . Fixed data cleaning. Added to llm usage for data cleaning code. And saved the cleaned dataset!

@@ -4,8 +4,6 @@
 # Date: 26 March 2024
 # Contact: lexi.knight@mail.utoronto.ca
 # License: MIT
-# Pre-requisites: -
-# Any other information needed? -
 
 
 #### Workspace setup ####
@@ -14,7 +12,6 @@
 
 # load packages
 library(tidyverse)
-
 
 
 #### Simulate data ####
@@ -51,7 +48,6 @@ simulated_data <- simulated_data %>%
     Treatment_Decision = treatment
   )
 
-
 simulated_data <- simulated_data |> select( Pathogenic_Stage,
                                             Lymph_Node_Involvement,
                                             Presence_of_Metastasis, Tumor_Size,
@@ -59,10 +55,6 @@ simulated_data <- simulated_data |> select( Pathogenic_Stage,
                                             Days_to_Death, Treatment_Type,
                                             Treatment_Decision)
 
-
-
-# View first few rows of the simulated data
-head(simulated_data)
 
 # Show a summary of the data
 summary(simulated_data)

### 2024-04-03 scripts/01-download_data.R

@@ -1,11 +1,10 @@
 #### Preamble ####
-# Purpose: Downloads and saves the lung cancer dataset from the OSF dataverse
-# database with the variables of interest. 
+# Purpose: Gives step by step instructions on how to access and download the 
+# lung cancer dataset from the OSF dataverse database. 
 # Author: Lexi Knight
 # Date: 28 March 2024
 # Contact: Lexi.Knight@mail.utoronto.ca
 # License: MIT
-# Pre-requisites: -
 
 
 ### Steps to Download data ###

### 2024-04-03 scripts/02-data_cleaning.R

@@ -1,15 +1,14 @@
 #### Preamble ####
-# Purpose: Cleans the raw plane data recorded by two observers..... [...UPDATE THIS...]
-# Author: Rohan Alexander [...UPDATE THIS...]
-# Date: 6 April 2023 [...UPDATE THIS...]
-# Contact: rohan.alexander@utoronto.ca [...UPDATE THIS...]
+# Purpose: Cleans the raw lung cancer data to prepare it for analysis.
+# Author: Lexi Knight
+# Date: 2 April 2024
+# Contact: lexi.knight@mail.utoronto.ca
 # License: MIT
-# Pre-requisites: [...UPDATE THIS...]
-# Any other information needed? [...UPDATE THIS...]
+# Pre-requisites: follow 01-download_data.R in scripts folder in order to access raw data
 
 #### Workspace setup ####
 # install packages
-install.packages("readr")
+# install.packages("readr")
 
 # load libraries 
 library(tidyverse)
@@ -25,25 +24,18 @@ colnames(raw_data)
 
 # select columns of interest 
 cleaned_data <- raw_data |>
-  select("days_to_death", "age_at_diagnosis", "ajcc_pathologic_m",
+  select("days_to_death", "ajcc_pathologic_m",
          "ajcc_pathologic_n", "ajcc_pathologic_stage", "ajcc_pathologic_t",
          "treatment_or_therapy", "treatment_type")
 
-# check data types of columns 
-str(cleaned_data)
-
-# handle missing or non-numeric values
+# Handle missing or non-numeric values in days_to_death column
 cleaned_data <- cleaned_data %>%
-  filter(!is.na(days_to_death), !is.na(age_at_diagnosis)) %>%
-  mutate_at(vars(days_to_death, age_at_diagnosis), as.numeric)
+  filter(!is.na(days_to_death), !is.na(as.numeric(days_to_death)))
 
-# create new column for survival_post_diagnosis
+# Exclude rows where any column contains "'--" values
 cleaned_data <- cleaned_data %>%
-  mutate(survival_post_diagnosis = days_to_death - age_at_diagnosis)
+  filter(!if_any(everything(), ~ . == "'--"))
 
-# Delete columns "days_to_death" and "age_at_diagnosis"
-cleaned_data <- cleaned_data %>%
-  select(-c(days_to_death, age_at_diagnosis))
 
 # clean the column names
 cleaned_data <- cleaned_data %>%
[…]

### 2024-04-03 scripts/03-test_data.R

@@ -4,14 +4,14 @@
 # Date: 28 March 2024
 # Contact: lexi.knight@mail.utoronto.ca
 # License: MIT
-# Pre-requisites: run 01-download_data.R as well as 02-data_cleaning.R in scripts in order to obtain raw and cleaned data. 
+# Pre-requisites: follow 01-download_data.R and run 02-data_cleaning.R in scripts in order to obtain raw and cleaned data. 
 
 
 #### Workspace setup ####
 
 #install packages 
-install.packages("tidyverse")
-install.packages("testthat")
+#install.packages("tidyverse")
+#install.packages("testthat")
 
 # load packages
 library(tidyverse)
@@ -19,7 +19,6 @@ library(testthat)
 
 
 #### Test data ####
-
 
 # Test if the dataset has 981 entries 
 test_that("Dataset has 981 entries", {
@@ -53,11 +52,6 @@ test_that("Age at diagnosis variable is numeric", {
 
 # Test if 'Days_to_Death' is numeric 
 class(simulated_data$Days_to_Death) == "numeric"
-
-# Test if 'Race' contains the correct categories 
-test_that("Race variable is correct", {
-  expect_true(all(simulated_data$Race %in% c('not reported', 'white','american indian or alaska native', 'asian', 'black or african american')))
-})
 
 # Test if 'Treatment_Type' only contains 'Pharmaceutical Therapy, NOS' and 'Radiation Therapy, NOS'
 test_that("Treatment type variable is correct", {

### 2024-04-03 scripts/02-data_cleaning.R
fixed issue with exlcuding "not reported" value in treatment_decision column. updated llm and downloaded new cleaned dataset

@@ -36,7 +36,6 @@ cleaned_data <- cleaned_data %>%
 cleaned_data <- cleaned_data %>%
   filter(!if_any(everything(), ~ . == "'--"))
 
-
 # clean the column names
 cleaned_data <- cleaned_data %>%
   rename(
@@ -46,6 +45,10 @@ cleaned_data <- cleaned_data %>%
     tumor_size = ajcc_pathologic_t,
     treatment_decision = treatment_or_therapy
   )
+
+# Exclude rows where treatment_decision contains "not reported" values
+cleaned_data <- cleaned_data %>%
+  filter(treatment_decision != "not reported")
 
 # view entire dataset
 print(cleaned_data, n = nrow(cleaned_data))

### 2024-04-03 scripts/03-test_data.R
updated and passed all 8 tests :) yay!

@@ -12,56 +12,44 @@
 #install packages 
 #install.packages("tidyverse")
 #install.packages("testthat")
-
-# load packages
-library(tidyverse)
-library(testthat)
+#install.packages("here")
 
 
-#### Test data ####
 
-# Test if the dataset has 981 entries 
-test_that("Dataset has 981 entries", {
-  expect_equal(nrow(simulated_data), 981)
-})
+#### Load data ####
 
-# Test if 'Pathogenic_Stage' contains the correct categories 
-test_that("Pathogenic stage variable is correct", {
-  expect_true(all(simulated_data$Pathogenic_Stage %in% c('Stage IA', 'Stage IIA', 'Stage IIIA', 'Stage IB', 'Stage IIB', 'Stage IV', 'Stage I')))
-})
-
-# Test if 'Lymph_Node_Involvement' contains the correct categories 
-test_that("Lymph node involvement variable is correct", {
-  expect_true(all(simulated_data$Lymph_Node_Involvement %in% c('N0', 'N1', 'N2', 'NX')))
-})
-
-# Test if 'Presence_of_Metastasis' contains the correct categories 
-test_that("Presence of metastasis variable is correct", {
-  expect_true(all(simulated_data$Presence_of_Metastasis %in% c('M0', 'MX', 'M1a', 'M1b')))
-})
-
-# Test if 'Tumor_Size' contains the correct categories 
-test_that("Tumor size variable is correct", {
-  expect_true(all(simulated_data$Tumor_Size %in% c('T1', 'T1a', 'T1b', 'T2', 'T2', 'T2b', 'T3', 'T4', 'TX')))
-})
-
-# Test if 'Age_at_Diagnosis_in_Days is numeric 
-test_that("Age at diagnosis variable is numeric", {
-  expect_true(is.numeric(simulated_data$Age_at_Diagnosis_in_Days))
-})
-
-# Test if 'Days_to_Death' is numeric 
-class(simulated_data$Days_to_Death) == "numeric"
-
-# Test if 'Treatment_Type' only contains 'Pharmaceutical Therapy, NOS' and 'Radiation Therapy, NOS'
-test_that("Treatment type variable is correct", {
-  expect_true(all(simulated_data$Treatment_Type %in% c('Pharmaceutical Therapy, NOS', 'Radiation Therapy, NOS')))
-})
-
-# Test if 'Treatment_Decision" contains the correct categories 
[…]

### 2024-04-03 paper/paper.qmd
update intro
playing around with different figures for data section

@@
 62.8% of patients with localized spread, 34.8% of patients with regional and 8% of patients with distant, advanced spread were found to survive for 5 years post diagnosis.
-More than half of these lung cancer patients have advanced spread to other organs when diagnosed(Markman, 2023) .
+More than half of these lung cancer patients have advanced spread to other organs when diagnosed(Markman, 2023).
 Overall, it is found that patients with no regional lymph node metastases, and smaller tumors are easier to be treated and thus are associated with improved survival rates (Zhang, 2015).
@@
 
-Presence of metastatic LN is one of the most important determinants of prognosis of NSCLC cases (Kai, 2021).
-In the early stage, cancer has not spread to lymph nodes.
-As severity increases, lymph node metastasis sequentially spreads to more distant lymph nodes such as mediastinal and there is severe lymph node involvement  (Park, 2017).
-Lymph node involvement, also termed lymph node ratio, is a crucial factor in guiding treatment options (Kai, 2021).
-A study made up of 97 patients with a mean age of 63 who have undergone surgery between the years 2009 and 2015 in Korea find that increased lymph node involvement is associated with a more advanced disease status and hence affiliated with prognosis (Park, 2017).
-
-Another study looked at 11,341 NSCLC patients between the years 2004 to 2015, from 18 geographically diverse populations, covering approximately 28% of the population of the United States.
-These patients were treatment naive and underwent surgical resection of the tumor.
-Although 5757 patients died, the rest showed great results, with a median survival of 22 months (Kai, 2021).
-The authors found that patients with low lymph node involvement lead to higher survival compared to patients with high lymph node ratios.
-A regression analysis revealed that lymph node ratio is an independent and significant predictor of patient survival.
-The authors also observed that disease burden and anatomical location of the lymph nodes involved may influence the patients survival (Kai, 2021).
-
 After tumor size, LN involvement and presence of distant metastasis are categorized, the pathogenic stage of the cancer is then determined (Eldridge, 2022).
@@
 The earlier the cancer is found, that is the lower the pathogenic stage, the greater the likelihood curative radiation therapy is an effective treatment (Eldridge, 2022).
-
 However, there is minimal literature looking at post-diagnosis survival rates based on pathogenic stage and method of treatment.
@@
 #| label: fig-bills
-#| fig-cap: Bills of penguins
+#| fig-cap: bb
 #| echo: false
+
+# install.packages
+# install.packages("ggplot2")
+# install.packages("dplyr")
+
+# load packaged
+library(ggplot2)
+library(dplyr)
+
+# Histogram of lung cancer patients survival time in days, post-diagnosis
+ggplot(analysis_data, aes(x = days_to_death)) +
+  geom_histogram(binwidth = 50, fill = "forestgreen", color = "black") +
+  labs(title = "Distribution of Survial Time in Lung Cancer Patients Post-Diagnosis", x = "Survival Time (days)", y = "Frequency") + theme_minimal()
+
+# create bar chart for pathogenic stage
+pathogenic_stage_counts <- analysis_data %>%
+  group_by(pathogenic_stage) %>%
+  summarise(count = n ()) %>%
+  mutate(percentage = count / sum(count) * 100) %>%
+  arrange(desc(pathogenic_stage))
+
+ggplot(pathogenic_stage_counts, aes(x = pathogenic_stage, y = percentage)) +
+  geom_bar(stat = "identity", fill = "forestgreen") +
+  labs(title = "Percentage of Patients by Pathogenic Stage",
+       x = "Pathogenic Stage",
+       y = "Percentage") +
+  theme_minimal()
+
+# Create bar chart for treatment type
+treatment_type_counts <- analysis_data %>%
+  group_by(treatment_type) %>%
+  summarise(count = n()) %>%
+  mutate(percentage = count / sum(count) * 100) %>%
+  arrange(desc(treatment_type))
+
+ggplot(treatment_type_counts, aes(x = treatment_type, y = count)) +
+  geom_bar(stat = "identity", fill = "forestgreen") +
+  labs(title = "Distribution of Patients by Treatment Type",
+       x = "Treatment Type",
+       y = "Count") +
+  theme_minimal()
+
+ggplot(treatment_type_counts, aes(x = treatment_type, y = percentage)) +
+  geom_bar(stat = "identity", fill = "forestgreen") +
+  labs(title = "Percentage of Patients by Treatment Type",
+       x = "Treatment Type",
+       y = "Percentage") +
+  theme_minimal()
+
+# trying to represent percentage of pts by treatment type in a different way
+
+# Calculate counts and percentages of patients by treatment type
+treatment_type_counts <- analysis_data %>%
+  group_by(treatment_type) %>%
+  summarise(count = n()) %>%
+  mutate(percentage = count / sum(count) * 100)
+
+# Pie chart for percentage of patients by treatment type
+ggplot(treatment_type_counts, aes(x = "", y = percentage, fill = treatment_type)) +
+  geom_bar(stat = "identity", width = 1, color = "white") +
+  coord_polar("y", start = 0) +
+  labs(title = "Percentage of Patients by Treatment Type",
+       fill = "Treatment Type",
+       x = NULL,
+       y = NULL) +
+  theme_void()
+
+# Stacked bar chart for percentage of patients by treatment type
+ggplot(treatment_type_counts, aes(x = "", y = percentage, fill = treatment_type)) +
+  geom_bar(stat = "identity") +
+  labs(title = "Percentage of Patients by Treatment Type",
+       x = NULL,
+       y = "Percentage") +
+  theme_minimal() +
+  theme(legend.position = "bottom")
+
+# dont like those either
+
+# Donut chart for percentage of patients by treatment type
+ggplot(treatment_type_counts, aes(x = "", y = percentage, fill = treatment_type)) +
+  geom_bar(stat = "identity", width = 1, color = "white") +
+  coord_polar("y", start = 0) +
+  labs(title = "Percentage of Patients by Treatment Type",
+       fill = "Treatment Type",
+       x = NULL,
+       y = NULL) +
+  theme_void() +
+  theme(legend.position = "right") +  # Adjust legend position if necessary
+  annotate("text", x = 0, y = 0, label = "Total", size = 10, color = "black")  # Add label for total in the center
+
+# hmm maybe donut was cuter
+
+# Waffle chart for percentage of patients by treatment type
+# install.packages("waffle")
+library(waffle)
+
+# Prepare data for waffle chart
+waffle_data <- treatment_type_counts %>%
+  mutate(squares = percentage / 2)  # Adjust scale if needed for better visualization
+
+# Create waffle chart
+waffle(waffle_data$squares, rows = 10, size = 0.6, colors = c("lightblue", "lightgreen"),
+       title = "Percentage of Patients by Treatment Type", 
+       xlab = "1 square = 2%", 
+       pad = 1)  # Adjust padding between squares if necessary
+
+# idea for the morning - sub in presence of distant metastasis, lymph node involvement and tumor size for pathogenic stage (frist graph)
 

### 2024-04-04 paper/paper.qmd
finished making figures for data section. messing around to make them cuter :)

--- 2024-04-02 paper/paper.qmd
@@
 #| label: fig-bills
-#| fig-cap: Bills of penguins
+#| fig-cap: bb
 #| echo: false
+
+# install.packages
+# install.packages("ggplot2")
+# install.packages("dplyr")
+# install.packages("showtext")
+
+# load packaged
+library(ggplot2)
+library(dplyr)
+library(showtext)
+
+# 1. Add the desired font (e.g., Patrick Hand)
+font_add_google("Patrick Hand", "patrick")
+
+# 2. Set the font for ggplot
+showtext_auto()
+
+# Histogram of survival time in days (days_to_death)
+ggplot(analysis_data, aes(x = days_to_death)) +
+  geom_histogram(binwidth = 50, fill = "forestgreen", color = "black") +
+  labs(title = "Distribution of Survival Time in Lung Cancer Patients Post-Diagnosis",
+       x = "Survival Time (days)", y = "Frequency") +
+  theme_minimal() +
+  theme(
+    text = element_text(family = "patrick") # Set font to Patrick Hand
+  )
+
+# create bar chart for pathogenic_stage
+pathogenic_stage_counts <- analysis_data %>%
+  group_by(pathogenic_stage) %>%
+  summarise(count = n ()) %>%
+  mutate(percentage = count / sum(count) * 100) %>%
+  arrange(desc(pathogenic_stage))
+
+ggplot(pathogenic_stage_counts, aes(x = pathogenic_stage, y = percentage)) +
+  geom_bar(stat = "identity", fill = "lightpink") +
+  labs(title = "Percentage of Patients by Pathogenic Stage",
+       x = "Pathogenic Stage",
+       y = "Percentage") +
+  theme_minimal() +
+  theme(
+    text = element_text(family = "patrick") # Set font to Patrick Hand
+  )
+
+###########################
+
+# Define color palette for different stages
+stage_colors <- c("Early" = "#FFB6C1", "Intermediate" = "#FF69B4", "Advanced" = "#FF1493")
+
+# Modify data to ensure correct ordering of stages
+analysis_data$pathogenic_stage <- factor(analysis_data$pathogenic_stage, levels = c("Early", "Intermediate", "Advanced"))
+
+# Create bar chart for pathogenic_stage
+pathogenic_stage_counts <- analysis_data %>%
+  group_by(pathogenic_stage) %>%
+  summarise(count = n ()) %>%
+  mutate(percentage = count / sum(count) * 100) %>%
+  arrange(desc(pathogenic_stage))
+
+# Plot the data
+ggplot(pathogenic_stage_counts, aes(x = pathogenic_stage, y = percentage, fill = pathogenic_stage)) +
+  geom_bar(stat = "identity") +
+  labs(title = "Percentage of Patients by Pathogenic Stage",
+       x = "Pathogenic Stage",
+       y = "Percentage") +
+  scale_fill_manual(values = stage_colors) +  # Apply color palette
+  theme_minimal() +
+  theme(
+    text = element_text(family = "patrick")  # Set font to Patrick Hand
+  )
+
+# create bar chart for presence_of_distant metastasis 
+presence_of_distant_metastasis_counts <- analysis_data %>%
+  group_by(presence_of_distant_metastasis) %>%
+  summarise(count = n ()) %>%
+  mutate(percentage = count / sum(count) * 100) %>%
+  arrange(desc(presence_of_distant_metastasis))
+
+ggplot(presence_of_distant_metastasis_counts, aes(x = presence_of_distant_metastasis, y = percentage)) +
+  geom_bar(stat = "identity", fill = "gold") +
+  labs(title = "Percentage of Patients by Presence of Distant Metastasis",
+       x = "Presence of Distant Metastasis",
+       y = "Percentage") +
+  theme_minimal() +
+  theme(
+    text = element_text(family = "patrick") # Set font to Patrick Hand
+  )
+
+# create bar chart for lymph node involvement 
+lymph_node_involvement_counts <- analysis_data %>%
+  group_by(lymph_node_involvement) %>%
+  summarise(count = n ()) %>%
+  mutate(percentage = count / sum(count) * 100) %>%
+  arrange(desc(lymph_node_involvement))
+
+ggplot(lymph_node_involvement_counts, aes(x = lymph_node_involvement, y = percentage)) +
+  geom_bar(stat = "identity", fill = "cyan") +
+  labs(title = "Percentage of Patients by Lymph Node Involvement",
+       x = "Lymph Node Involvement",
+       y = "Percentage") +
+  theme_minimal() +
+  theme(
+    text = element_text(family = "patrick") # Set font to Patrick Hand
+  )
+
+# create bar chart for tumor size
+tumor_size_counts <- analysis_data %>%
+  group_by(tumor_size) %>%
+  summarise(count = n ()) %>%
+  mutate(percentage = count / sum(count) * 100) %>%
+  arrange(desc(tumor_size))
+
+ggplot(tumor_size_counts, aes(x = tumor_size, y = percentage)) +
+  geom_bar(stat = "identity", fill = "purple") +
+  labs(title = "Percentage of Patients by Tumor Size",
+       x = "Tumor Size",
+       y = "Percentage") +
+  theme_minimal() +
+  theme(
+    text = element_text(family = "patrick") # Set font to Patrick Hand
+  )
+
+# create bar chart for treatment decision
+ggplot(treatment_decision_counts, aes(x = reorder(treatment_decision, -percentage), y = percentage)) +
+  geom_bar(stat = "identity", fill = "blue") +
+  geom_text(aes(label = sprintf("%.1f%%", percentage)), hjust = -0.2, color = "black", size = 3) +
+  labs(title = "Percentage of Patients by Treatment Decision",
+       x = "Treatment Decision",
+       y = "Percentage") +
+  theme_minimal() +
+  coord_flip() +
+  theme(
+    text = element_text(family = "patrick") # Set font to Patrick Hand
+  )
+
+# Create bar chart for treatment type
+ggplot(treatment_type_counts, aes(x = reorder(treatment_type, -percentage), y = percentage)) +
+  geom_bar(stat = "identity", fill = "orange") +
+  geom_text(aes(label = sprintf("%.1f%%", percentage)), hjust = -0.2, color = "black", size = 3) +
+  geom_text(data = treatment_type_counts, aes(label = treatment_type), hjust = 1.2, color = "black", size = 3) +
+  labs(title = "Percentage of Patients by Treatment Type",
+       x = "Treatment Type",
+       y = "Percentage") +
+  theme_minimal() +
+  coord_flip() +
+  scale_x_discrete(breaks = seq(0, max(treatment_type_counts$percentage), by = 5)) +
+  theme(
+    text = element_text(family = "patrick") # Set font to Patrick Hand
+  )
 

### 2024-04-04 paper/paper.qmd
woking on data section - tyring to fix pathogenic stage bar chart

@@
 
-# 1. Add the desired font (e.g., Patrick Hand)
+# Add the desired font (e.g., Patrick Hand)
 font_add_google("Patrick Hand", "patrick")
 
-# 2. Set the font for ggplot
+# Set the font for ggplot
 showtext_auto()
@@
   )
+
+# Replace NA values with your desired stages
+analysis_data <- analysis_data %>%
+  mutate(pathogenic_stage = ifelse(is.na(pathogenic_stage), 
+                                   c('Stage I', 'Stage IA', 'Stage IB', 
+                                     'Stage II', 'Stage IIA', 'Stage IIB', 
+                                     'Stage III', 'Stage IIIA', 'Stage IIIB', 
+                                     'Stage IV'), 
+                                   pathogenic_stage))
 
@@
        y = "Percentage") +
-  theme_minimal() +
-  theme(
-    text = element_text(family = "patrick") # Set font to Patrick Hand
-  )
-
-###########################
-
-# Define color palette for different stages
-stage_colors <- c("Early" = "#FFB6C1", "Intermediate" = "#FF69B4", "Advanced" = "#FF1493")
-
-# Modify data to ensure correct ordering of stages
-analysis_data$pathogenic_stage <- factor(analysis_data$pathogenic_stage, levels = c("Early", "Intermediate", "Advanced"))
-
-# Create bar chart for pathogenic_stage
-pathogenic_stage_counts <- analysis_data %>%
-  group_by(pathogenic_stage) %>%
-  summarise(count = n ()) %>%
-  mutate(percentage = count / sum(count) * 100) %>%
-  arrange(desc(pathogenic_stage))
-
-# Plot the data
-ggplot(pathogenic_stage_counts, aes(x = pathogenic_stage, y = percentage, fill = pathogenic_stage)) +
-  geom_bar(stat = "identity") +
-  labs(title = "Percentage of Patients by Pathogenic Stage",
-       x = "Pathogenic Stage",
-       y = "Percentage") +
-  scale_fill_manual(values = stage_colors) +  # Apply color palette
   theme_minimal() +

### 2024-04-04 paper/paper1.qmd
fixed and finished data section figures

--- 2024-04-04 paper/paper.qmd
@@
 
-# load packaged
+# load packages
 library(ggplot2)
@@
 library(showtext)
+
+# read in data
+analysis_data <- read.csv(file = here::here("data/analysis_data/lung_cancer_analysis_data.csv"))
 
@@
   )
-
-# Replace NA values with your desired stages
-analysis_data <- analysis_data %>%
-  mutate(pathogenic_stage = ifelse(is.na(pathogenic_stage), 
-                                   c('Stage I', 'Stage IA', 'Stage IB', 
-                                     'Stage II', 'Stage IIA', 'Stage IIB', 
-                                     'Stage III', 'Stage IIIA', 'Stage IIIB', 
-                                     'Stage IV'), 
-                                   pathogenic_stage))
 
@@
 # create bar chart for treatment decision
+treatment_decision_counts <- analysis_data %>%
+  group_by(treatment_decision) %>%
+  summarise(count = n ()) %>%
+  mutate(percentage = count / sum(count) * 100) %>%
+  arrange(desc(treatment_decision))
+
 ggplot(treatment_decision_counts, aes(x = reorder(treatment_decision, -percentage), y = percentage)) +
@@
 # Create bar chart for treatment type
+treatment_type_counts <- analysis_data %>%
+  group_by(treatment_type) %>%
+  summarise(count = n ()) %>%
+  mutate(percentage = count / sum(count) * 100) %>%
+  arrange(desc(treatment_type))
+
 ggplot(treatment_type_counts, aes(x = reorder(treatment_type, -percentage), y = percentage)) +

### 2024-04-04 paper/paper.qmd
messing around with different figures for results section.

@@
 
+# Load required libraries
+library(ggplot2)
+
+# Scatter plot survival time vs. pathogenic stage - to compare survival probabilities
+ggplot(analysis_data, aes(x = pathogenic_stage, y = days_to_death)) +
+  geom_point(size = 3, color = "lightpink") +
+  geom_smooth(method = "lm", se = FALSE, color = "red") +  
+  # Add a linear trend line
+  labs(title = "Survival Time vs. Pathogenic Stage Post-Diagnosis",
+       x = "Pathogenic Stage",
+       y = "Survival Time (days)") +
+  theme_minimal() +
+  theme(axis.text.x = element_text(angle = 45, hjust = 1)) + 
+  # Rotate x-axis labels for better readability
+  theme(text = element_text(family = "patrick") 
+  # Set font to Patrick Hand
+  )
+
+# Plotting
+ggplot(analysis_data, aes(x = pathogenic_stage, y = days_to_death)) +
+  geom_boxplot(fill = "lightblue") +
+  labs(title = "Days to Death Distribution by Pathogenic Stage for Lung Cancer Patients",
+       x = "Pathogenic Stage",
+       y = "Days to Death") +
+  theme_minimal() +
+  theme(axis.text.x = element_text(angle = 45, hjust = 1))  # Rotate x-axis labels for better readability
+
+# Scatter plot survival time vs. Presence of Distant Metastasis
+ggplot(analysis_data, aes(x = presence_of_distant_metastasis, y = days_to_death)) +
+  geom_point(size = 3, color = "gold") +
+  geom_smooth(method = "lm", se = FALSE, color = "red") +  
+  # Add a linear trend line
+  labs(title = "Survival Time vs. Presence of Distant Metastasis Post-Diagnosis",
+       x = "Presence of Distant Metastasis",
+       y = "Survival Time (days)") +
+  theme_minimal() +
+  theme(axis.text.x = element_text(angle = 45, hjust = 1)) + 
+  # Rotate x-axis labels for better readability
+  theme(text = element_text(family = "patrick") 
+  # Set font to Patrick Hand
+  )
+
+# Scatter plot survival time vs. Lymph node involvement
+ggplot(analysis_data, aes(x = lymph_node_involvement, y = days_to_death)) +
+  geom_point(size = 3, color = "cyan") +
+  geom_smooth(method = "lm", se = FALSE, color = "red") +  
+  # Add a linear trend line
+  labs(title = "Survival Time vs. Lymph Node Involvement Post-Diagnosis",
+       x = "Lymph Node Involvement",
+       y = "Survival Time (days)") +
+  theme_minimal() +
+  theme(axis.text.x = element_text(angle = 45, hjust = 1)) + 
+  # Rotate x-axis labels for better readability
+  theme(text = element_text(family = "patrick") 
+  # Set font to Patrick Hand
+  )
+
+# Scatter plot survival time vs. Tumor Size
+ggplot(analysis_data, aes(x = tumor_size, y = days_to_death)) +
+  geom_point(size = 3, color = "purple") +
+  geom_smooth(method = "lm", se = FALSE, color = "red") +  
+  # Add a linear trend line
+  labs(title = "Survival Time vs. Tumor Size Post-Diagnosis",
+       x = "Tumor Size",
+       y = "Survival Time (days)") +
+  theme_minimal() +
+  theme(axis.text.x = element_text(angle = 45, hjust = 1)) + 
+  # Rotate x-axis labels for better readability
+  theme(text = element_text(family = "patrick") 
+  # Set font to Patrick Hand
+  )
+
+# Scatter plot survival time vs. Treatment Decision
+ggplot(analysis_data, aes(x = treatment_decision, y = days_to_death)) +
+  geom_point(size = 3, color = "blue") +
+  geom_smooth(method = "lm", se = FALSE, color = "red") +  
+  # Add a linear trend line
+  labs(title = "Survival Time vs. Treatment Decision Post-Diagnosis",
+       x = "Treatment Decision",
+       y = "Survival Time (days)") +
+  theme_minimal() +
+  theme(axis.text.x = element_text(angle = 45, hjust = 1)) + 
+  # Rotate x-axis labels for better readability
+  theme(text = element_text(family = "patrick") 
+  # Set font to Patrick Hand
+  )
+
+# Scatter plot survival time vs. Treatment Type
+ggplot(analysis_data, aes(x = treatment_type, y = days_to_death)) +
+  geom_point(size = 3, color = "cyan") +
+  geom_smooth(method = "lm", se = FALSE, color = "red") +  
+  # Add a linear trend line
+  labs(title = "Survival Time vs. Treatment Type Post-Diagnosis",
+       x = "Treatment Type",
+       y = "Survival Time (days)") +
+  theme_minimal() +
+  theme(axis.text.x = element_text(angle = 45, hjust = 1)) + 
+  # Rotate x-axis labels for better readability
+  theme(text = element_text(family = "patrick") 
+  # Set font to Patrick Hand
+  )
+
 ```

### 2024-04-04 paper/paper.qmd
messing around with figures in results section. kinda chaotic but im getting there, slowly :)

@@
 
-# Plotting
+# Line graph with trend lines 
 ggplot(analysis_data, aes(x = pathogenic_stage, y = days_to_death)) +
-  geom_boxplot(fill = "lightblue") +
-  labs(title = "Days to Death Distribution by Pathogenic Stage for Lung Cancer Patients",
+  geom_line(color = "blue", group = 1) +
+  geom_smooth(method = "lm", se = FALSE, color = "red") +  
+  # Add a linear trend line
+  labs(title = "Days to Death vs. Pathogenic Stage for Lung Cancer Patients",
        x = "Pathogenic Stage",
@@
   theme(axis.text.x = element_text(angle = 45, hjust = 1))  # Rotate x-axis labels for better readability
+
+# not sure what this is 
+# Plotting
+ggplot(analysis_data, aes(x = days_to_death, fill = pathogenic_stage)) +
+  geom_density(alpha = 0.5) +
+  geom_density(color = "black", size = 1.5) +
+  labs(title = "Density Plot of Days to Death by Pathogenic Stage for Lung Cancer Patients",
+       x = "Days to Death",
+       y = "Density") +
+  scale_fill_manual(values = c("lightgreen", "lightgreen", "lightgreen","forestgreen", "forestgreen", "forestgreen", "darkgreen","darkgreen", "darkgreen", "red")) +
+  theme_minimal()
+
+# not sure .2
+ggplot(analysis_data, aes(x = days_to_death, fill = pathogenic_stage)) +
+  geom_density(alpha = 0.5) +
+  geom_density(color = "black", size = 1.5) +
+  labs(title = "Density Plot of Days to Death by Pathogenic Stage for Lung Cancer Patients",
+       x = "Days to Death",
+       y = "Density") +
+  scale_fill_manual(values = c("lightgreen", "lightgreen", "lightgreen","forestgreen", "forestgreen", "forestgreen", "darkgreen","darkgreen", "darkgreen", "red")) +
+  theme_minimal() +
+  facet_wrap(~pathogenic_stage, scales = "free")
+
+# Boxplot 
+ggplot(analysis_data, aes(x = pathogenic_stage, y = days_to_death)) +
+  geom_boxplot(fill = "lightblue") +
+  labs(title = "Survial Distribution Post-Diagnosis by Pathogenic Stage",
+       x = "Pathogenic Stage",
+       y = "Days to Death") +
+  theme_minimal() +
+  theme(axis.text.x = element_text(angle = 45, hjust = 1))  # Rotate x-axis labels for better readability
+
+######################################
 
@@
   )
+#########################
+
+# create bar chart for treatment decision
+treatment_decision_counts <- analysis_data %>%
+  group_by(treatment_decision) %>%
+  summarise(count = n ()) %>%
+  mutate(percentage = count / sum(count) * 100) %>%
+  arrange(desc(treatment_decision))
+
+ggplot(treatment_decision_counts, aes(x = reorder(treatment_decision, -percentage), y = percentage)) +
+  geom_bar(stat = "identity", fill = "blue") +
+  geom_text(aes(label = sprintf("%.1f%%", percentage)), hjust = -0.2, color = "black", size = 3) +
+  labs(title = "Percentage of Patients by Treatment Decision",
+       x = "Treatment Decision",
+       y = "Percentage") +
+  theme_minimal() +
+  coord_flip() +
+  theme(
+    text = element_text(family = "patrick") # Set font to Patrick Hand
+  )
 
@@
   )
+
+#######################
+
+# Grouped bar chart for survival time vs. Treatment Type
+ggplot(analysis_data, aes(x = treatment_type, y = days_to_death, fill = treatment_type)) +
+  geom_bar(stat = "summary", fun = "mean", position = "dodge") +
+  labs(title = "Average Survival Time by Treatment Type Post-Diagnosis",
+       x = "Treatment Type",
+       y = "Average Survival Time (days)") +
+  theme_minimal() +
+  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
+  theme(text = element_text(family = "patrick")) 
+  # Set font to Patrick Hand
+
+# Horizontal grouped bar chart for survival time vs. Treatment Type
+ggplot(analysis_data, aes(y = treatment_type, x = days_to_death, fill = treatment_type)) +
+  geom_bar(stat = "summary", fun = "mean", position = "dodge", orientation = "horizontal") +
+  labs(title = "Average Survival Time by Treatment Type Post-Diagnosis",
+       y = "Treatment Type",
+       x = "Average Survival Time (days)") +
+  theme_minimal() +
+  theme(text = element_text(family = "patrick")) 
+  # Set font to Patrick Hand
+
+# Grouped dot plot for survival time vs. Treatment Type
+ggplot(analysis_data, aes(x = treatment_type, y = days_to_death, color = treatment_type)) +
+  geom_jitter(position = position_dodge(width = 0.5), size = 3, alpha = 0.7) +
+  labs(title = "Survival Time by Treatment Type Post-Diagnosis",
+       x = "Treatment Type",
+       y = "Survival Time (days)") +
+  theme_minimal() +
+  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
+  theme(text = element_text(family = "patrick")) 
+  # Set font to Patrick Hand
+
+# Grouped dot plot with flipped axes for survival time vs. Treatment Type
+ggplot(analysis_data, aes(y = treatment_type, x = days_to_death, color = treatment_type)) +
+  geom_jitter(position = position_dodge(width = 0.5), size = 3, alpha = 0.7) +
+  labs(title = "Survival Time by Treatment Type Post-Diagnosis",
+       y = "Treatment Type",
+       x = "Survival Time (days)") +
+  theme_minimal() +
+  theme(axis.text.y = element_text(angle = 0, hjust = 1)) +
+  theme(text = element_text(family = "patrick")) 
+  # Set font to Patrick Hand
+
+# Stacked bar chart for treatment method vs. Average Survival Time
+ggplot(analysis_data, aes(x = days_to_death, fill = treatment_type)) +
+  geom_bar(position = "fill") +
+  labs(title = "Distribution of Treatment Methods by Average Survival Time Post-Diagnosis",
+       x = "Average Survival Time (days)",
+       y = "Proportion of Patients") +
+  scale_fill_manual(values = c("cyan", "magenta")) + # Adjust fill colors
+  theme_minimal() +
+  theme(text = element_text(family = "patrick")) 
+  # Set font to Patrick Hand
+
+# Stacked bar chart for treatment method vs. Average Survival Time
+ggplot(analysis_data, aes(x = days_to_death, fill = treatment_type)) +
+  geom_bar(position = "fill") +
+  labs(title = "Distribution of Treatment Methods by Average Survival Time Post-Diagnosis",
+       x = "Average Survival Time (days)",
+       y = "Proportion of Patients") +
+  scale_fill_manual(values = c("cyan", "magenta")) + # Adjust fill colors
+  scale_x_continuous(expand = c(0, 0)) + # Adjust x-axis limits
+  theme_minimal() +
+  theme(text = element_text(family = "patrick")) 
+  # Set font to Patrick Hand
+
+# Stacked bar chart for treatment method vs. Average Survival Time
+ggplot(analysis_data, aes(x = days_to_death, fill = treatment_type)) +
+  geom_bar(position = "fill") +
+  labs(title = "Distribution of Treatment Methods by Average Survival Time Post-Diagnosis",
+       x = "Average Survival Time (days)",
+       y = "Proportion of Patients") +
+  scale_fill_manual(values = c("cyan", "magenta")) + # Adjust fill colors
+  scale_x_continuous(expand = c(0, 0), limits = c(0, max(analysis_data$days_to_death))) + # Adjust x-axis limits
+  theme_minimal() +
+  theme(text = element_text(family = "patrick")) 
+  # Set font to Patrick Hand
+
+# Boxplot 
+ggplot(analysis_data, aes(x = treatment_type, y = days_to_death)) +
+  geom_boxplot(fill = "lightblue") +
+  labs(title = "Survial Distribution Post-Diagnosis by Treatment Type",
+       x = "Treatment Type",
+       y = "Days to Death") +
+  theme_minimal() +
+  theme(axis.text.x = element_text(angle = 45, hjust = 1))  # Rotate x-axis labels for better readability
 

### 2024-04-05 scripts/04-model.R
attempt at modeling, inlcuded llm and saving model

@@ -1,37 +1,60 @@
 #### Preamble ####
-# Purpose: Models... [...UPDATE THIS...]
-# Author: Rohan Alexander [...UPDATE THIS...]
-# Date: 11 February 2023 [...UPDATE THIS...]
-# Contact: rohan.alexander@utoronto.ca [...UPDATE THIS...]
+# Purpose: To create the logistic regression model.
+# Author: Lexi Knight
+# Date: 4 April 2024
+# Contact: lexi.knight@mail.utoronto.ca
 # License: MIT
-# Pre-requisites: [...UPDATE THIS...]
-# Any other information needed? [...UPDATE THIS...]
+# Pre-requisites: follow 01-download_data.R and run 02-data_cleaning.R in scripts in order to obtain raw and cleaned data. Run 03-test_data.R to test cleaned dataset.  
 
+# note about approach: linear model (ch. 13.2, logistic regression model )
+# note, must delete treatment_decision from entire analysis. 
 
 #### Workspace setup ####
-library(tidyverse)
+# install packages
+# install.packages("broom.mixed")
+# install.packages("dplyr")
+# install.packages("rstanarm")
+# install.packages("tidyverse")
+
+# load packages
+library(broom.mixed)
+library(dplyr)
 library(rstanarm)
+library(tidyverse)
 
 #### Read data ####
-analysis_data <- read_csv("data/analysis_data/analysis_data.csv")
+analysis_data <- read.csv(file = here::here("data/analysis_data/lung_cancer_analysis_data.csv"))
 
 ### Model data ####
-first_model <-
-  stan_glm(
-    formula = flying_time ~ length + width,
-    data = analysis_data,
-    family = gaussian(),
-    prior = normal(location = 0, scale = 2.5, autoscale = TRUE),
-    prior_intercept = normal(location = 0, scale = 2.5, autoscale = TRUE),
-    prior_aux = exponential(rate = 1, autoscale = TRUE),
-    seed = 853
-  )
+set.seed(123) # setting seed for reproducibility
+
+## Linear Regression Model 
+# interested in predicting days_to_death, a continuous outcome variable based on
+# multiple predictor variables  (pathogenic_stage, lymph_node_involvement, 
+# presence_of_distant_metastasis, tumor_size, treatment_decision, treatment_type)
+# Linear regression model are suitable for predicting continuous outcomes
+# quantifying relationship between predictor variables and outcome variable. 
+
[…]

### 2024-04-05 scripts/00-simulate_data.R
fixed issue with pathogenic vs. pathologic stage. removed treatment_decision from dataset and thus regenerated analysis_data (cleaned dataset)

@@ -27,33 +27,26 @@ simulated_data <- data.frame(
   ajcc_pathologic_n = sample(c("N0", "N1", "N2", "NX"), num_entries, replace = TRUE),
   ajcc_pathologic_m = sample(c("M0", "MX", "M1a", "M1b"), num_entries, replace = TRUE),
   ajcc_pathologic_t = sample(c("T1", "T1a", "T1b", "T2", "T2a", "T2b", "T3", "T4", "TX"), num_entries, replace = TRUE),
-  age_at_diagnosis = round(runif(num_entries, min = 14681, max = 32872)),
   days_to_death = round(runif(num_entries, min = 0, max = 6236)),
   ajcc_pathologic_stage = sample(c("Stage IA", "Stage IIA", "Stage IIIA", "Stage IB", "Stage IIB", "Stage IV", "Stage I"), num_entries, replace = TRUE),
-  treatment_type = sample(c("Pharmaceutical Therapy, NOS", "Radiation Therapy, NOS"), num_entries, replace = TRUE),
-  treatment = sample(c("yes", "no", "not reported"), num_entries, replace = TRUE)
-) 
+  treatment_type = sample(c("Pharmaceutical Therapy, NOS", "Radiation Therapy, NOS"), num_entries, replace = TRUE)
+)
 
 
 # Renaming variables for clarity 
 simulated_data <- simulated_data %>%
   rename(
-    Pathogenic_Stage = ajcc_pathologic_stage,
-    Lymph_Node_Involvement = ajcc_pathologic_n,
-    Presence_of_Metastasis = ajcc_pathologic_m,
-    Tumor_Size = ajcc_pathologic_t,
-    Age_at_Diagnosis_in_Days = age_at_diagnosis,
-    Days_to_Death = days_to_death,
-    Treatment_Type = treatment_type,
-    Treatment_Decision = treatment
+    pathologic_stage = ajcc_pathologic_stage,
+    lymph_node_involvement = ajcc_pathologic_n,
+    presence_of_distant_metastasis = ajcc_pathologic_m,
+    tumor_size = ajcc_pathologic_t,
   )
 
-simulated_data <- simulated_data |> select( Pathogenic_Stage,
-                                            Lymph_Node_Involvement,
-                                            Presence_of_Metastasis, Tumor_Size,
-                                            Age_at_Diagnosis_in_Days,
-                                            Days_to_Death, Treatment_Type,
-                                            Treatment_Decision)
[…]

### 2024-04-05 scripts/02-data_cleaning.R

@@ -26,7 +26,7 @@ colnames(raw_data)
 cleaned_data <- raw_data |>
   select("days_to_death", "ajcc_pathologic_m",
          "ajcc_pathologic_n", "ajcc_pathologic_stage", "ajcc_pathologic_t",
-         "treatment_or_therapy", "treatment_type")
+         "treatment_type")
 
 # Handle missing or non-numeric values in days_to_death column
 cleaned_data <- cleaned_data %>%
@@ -41,14 +41,9 @@ cleaned_data <- cleaned_data %>%
   rename(
     lymph_node_involvement = ajcc_pathologic_n,
     presence_of_distant_metastasis = ajcc_pathologic_m,
-    pathogenic_stage = ajcc_pathologic_stage,
+    pathologic_stage = ajcc_pathologic_stage,
     tumor_size = ajcc_pathologic_t,
-    treatment_decision = treatment_or_therapy
   )
-
-# Exclude rows where treatment_decision contains "not reported" values
-cleaned_data <- cleaned_data %>%
-  filter(treatment_decision != "not reported")
 
 # view entire dataset
 print(cleaned_data, n = nrow(cleaned_data))

### 2024-04-05 scripts/02-data_cleaning.R
saved analysis_data as parquet. trying to solve issue that days_to_death is showing up as character instead of numeric

@@ -9,10 +9,12 @@
 #### Workspace setup ####
 # install packages
 # install.packages("readr")
+# install.packages("arrow")
 
 # load libraries 
 library(tidyverse)
 library(readr)
+library(arrow)
 
 #### Clean data ####
 
@@ -45,8 +47,15 @@ cleaned_data <- cleaned_data %>%
     tumor_size = ajcc_pathologic_t,
   )
 
+# Convert the days_to_death column to numeric
+analysis_data$days_to_death <- as.numeric(analysis_data$days_to_death)
+
+# Check the data types again
+#print(class(analysis_data$days_to_death))
+
+
 # view entire dataset
 print(cleaned_data, n = nrow(cleaned_data))
 
 #### Save data ####
-write_csv(cleaned_data, "data/analysis_data/lung_cancer_analysis_data.csv")
+write_parquet(cleaned_data, "data/analysis_data/lung_cancer_analysis_data.parquet")

### 2024-04-05 scripts/03-test_data.R

@@ -9,47 +9,52 @@
 
 #### Workspace setup ####
 
-#install packages 
-#install.packages("tidyverse")
-#install.packages("testthat")
-#install.packages("here")
+# install packages 
+# install.packages("tidyverse")
+# install.packages("testthat")
+# install.packages("here")
+# install.packages("arrow")
+# install.packages("dplyr")
 
+# load libraries
+# library(tidyverse) # maybe dont need
+library(testthat)
+library(here)
+library(arrow) # needed to read parquet
+# library(dplyr) # maybe dont need
 
-
-#### Load data ####
+#### Test data ####
 
 test_that("Data Cleaning Tests", {
-  # load data
-  analysis_data <- read.csv(file = here::here("data/analysis_data/lung_cancer_analysis_data.csv"))
-  
   # load packages
-  library(tidyverse)
-  library(testthat)
-  library(here)
+  suppressWarnings(library(tidyverse))
+  suppressWarnings(library(dplyr))
   
-  # Test if the dataset has 346 entries
-  expect_equal(nrow(analysis_data), 346)
+  # load data
+  analysis_data <- read_parquet(file = here::here("data/analysis_data/lung_cancer_analysis_data.parquet"))
   
-  # Test if 'Pathogenic_Stage' contains the correct categories 
-  expect_true(all(analysis_data$pathogenic_stage %in% c('Stage I', 'Stage IA', 'Stage IB', 'Stage II', 'Stage IIA', 'Stage IIB', 'Stage III', 'Stage IIIA', 'Stage IIIB', 'Stage IV')))
+  # Test if the dataset has 382 entries
+  expect_equal(nrow(analysis_data), 382)
   
-  # Test if 'Lymph_Node_Involvement' contains the correct categories 
-  expect_true(all(analysis_data$lymph_node_involvement %in% c('N0', 'N1', 'N2', 'NX')))
+  # Test if 'pathologic_stage' contains the correct categories 
+  expect_true(all(analysis_data$pathologic_stage %in% c('Stage I', 'Stage IA', 'Stage IB', 'Stage II', 'Stage IIA', 'Stage IIB', 'Stage III', 'Stage IIIA', 'Stage IIIB', 'Stage IV')))
   
-  # Test if 'Presence_of_Metastasis' contains the correct categories
+  # Test if 'lymph_node_involvement' contains the correct categories 
[…]

### 2024-04-05 scripts/04-model.R

@@ -21,9 +21,10 @@ library(broom.mixed)
 library(dplyr)
 library(rstanarm)
 library(tidyverse)
+library(arrow) # needed to read parquet
 
 #### Read data ####
-analysis_data <- read.csv(file = here::here("data/analysis_data/lung_cancer_analysis_data.csv"))
+analysis_data <- read_parquet(file = here::here("data/analysis_data/lung_cancer_analysis_data.parquet"))
 
 ### Model data ####
 set.seed(123) # setting seed for reproducibility
@@ -38,7 +39,7 @@ set.seed(123) # setting seed for reproducibility
 # Fit a linear regression model to predict continuous outcome variable 
 # 'days_to_death', using stan_glm
 linear_model <- stan_glm( # Bayesian generalized linear models via Stan
-  formula = days_to_death ~  pathogenic_stage + lymph_node_involvement +
+  formula = days_to_death ~  pathologic_stage + lymph_node_involvement +
     presence_of_distant_metastasis + tumor_size + treatment_type,
   data = analysis_data,
   family = gaussian(),

### 2024-04-05 scripts/02-data_cleaning.R
fixed days_to_death. It is now numeric and thus all tests run successfully :) Updated analysis_data which is saved as a parquet. Scripts is officially done! WHoop WHoop!

@@ -47,11 +47,12 @@ cleaned_data <- cleaned_data %>%
     tumor_size = ajcc_pathologic_t,
   )
 
+
 # Convert the days_to_death column to numeric
-analysis_data$days_to_death <- as.numeric(analysis_data$days_to_death)
+cleaned_data$days_to_death <- as.numeric(analysis_data$days_to_death)
 
 # Check the data types again
-#print(class(analysis_data$days_to_death))
+# print(class(analysis_data$days_to_death))
 
 
 # view entire dataset

### 2024-04-05 paper/paper.qmd
updating paper - working on modeling

@@
 library(showtext)
+library(arrow)
 
-# read in data
-analysis_data <- read.csv(file = here::here("data/analysis_data/lung_cancer_analysis_data.csv"))
+### read in data ###
+analysis_data <- read_parquet(file = here::here("data/analysis_data/lung_cancer_analysis_data.parquet"))
 
@@
 
-# create bar chart for pathogenic_stage
-pathogenic_stage_counts <- analysis_data %>%
-  group_by(pathogenic_stage) %>%
+# create bar chart for pathologic_stage
+pathologic_stage_counts <- analysis_data %>%
+  group_by(pathologic_stage) %>%
   summarise(count = n ()) %>%
   mutate(percentage = count / sum(count) * 100) %>%
-  arrange(desc(pathogenic_stage))
+  arrange(desc(pathologic_stage))
 
-ggplot(pathogenic_stage_counts, aes(x = pathogenic_stage, y = percentage)) +
+ggplot(pathologic_stage_counts, aes(x = pathologic_stage, y = percentage)) +
   geom_bar(stat = "identity", fill = "lightpink") +
-  labs(title = "Percentage of Patients by Pathogenic Stage",
-       x = "Pathogenic Stage",
+  labs(title = "Percentage of Patients by Pathologic Stage",
+       x = "Pathologic Stage",
        y = "Percentage") +
@@
 
-# create bar chart for treatment decision
-treatment_decision_counts <- analysis_data %>%
-  group_by(treatment_decision) %>%
-  summarise(count = n ()) %>%
-  mutate(percentage = count / sum(count) * 100) %>%
-  arrange(desc(treatment_decision))
-
-ggplot(treatment_decision_counts, aes(x = reorder(treatment_decision, -percentage), y = percentage)) +
-  geom_bar(stat = "identity", fill = "blue") +
-  geom_text(aes(label = sprintf("%.1f%%", percentage)), hjust = -0.2, color = "black", size = 3) +
-  labs(title = "Percentage of Patients by Treatment Decision",
-       x = "Treatment Decision",
-       y = "Percentage") +
-  theme_minimal() +
-  coord_flip() +
-  theme(
-    text = element_text(family = "patrick") # Set font to Patrick Hand
-  )
-
 # Create bar chart for treatment type
@@
 
-Talk way more about it.
-
 # Model {#sec-model}
 
-The goal of our modeling strategy is twofold.
-Firstly, we aim to investigate the post-diagnosis survival outcomes of lung cancer patients, considering the pathogenic stage.
-Secondly, we seek to assess the efficacy of two common treatment modalities, radiation therapy and chemotherapy, in improving survival chances across different pathogenic stages.
+## Linear Regression Model 
+# interested in predicting days_to_death, a continuous outcome variable based on
+# multiple predictor variables  (pathogenic_stage, lymph_node_involvement, 
+# presence_of_distant_metastasis, tumor_size, treatment_decision, treatment_type)
+# Linear regression model are suitable for predicting continuous outcomes
+# quantifying relationship between predictor variables and outcome variable.
 
-Here, we briefly describe the Bayesian analysis model used to investigate survival outcomes in lung cancer patients based on pathogenic stage and treatment modality.
-Detailed background information and diagnostic assessments are provided in the [Appendix -@].
-
-Our Bayesian analysis model is tailored to capture the complex relationships between pathogenic stage, treatment modality, and post-diagnosis survival outcomes in lung cancer patients.
-The model accounts for the multifaceted nature of cancer progression, incorporating tumor size, metastatic spread, and lymph node involvement to characterize pathogenic stage.
+# Fit a linear regression model to predict continuous outcome variable 
+# 'days_to_death'
 
 ## Model set-up
+In this study, we aim to investigate the survival outcomes of lung cancer patients post-diagnosis.
+Let $y_i$ denote the number of days from diagnosis to death for patient $i$.
+We consider several predictors including pathogenic stage, lymph node involvement, presence of distant metastasis, tumor size, and treatment type.
 
+We model the survival time $y_i$ using a linear regression framework where $mi_i$ is the expected survival time for patient $i$, and $σ$ represents the standard deviation of the survival times.
+The linear predictor $mu_i$ is specified as:
+
+\begin{align*}
+y_i | \mu_i, \sigma &\sim \text{Normal}(\mu_i, \sigma) \\
+\mu_i &= \alpha + \beta_{\text{pathologic\_stage}} \times \text{pathologic\_stage}_i + \beta_{\text{lymph\_node}} \times \text{lymph\_node\_involvement}_i \\
+&\quad + \beta_{\text{metastasis}} \times \text{presence\_of\_distant\_metastasis}_i + \beta_{\text{tumor\_size}} \times \text{tumor\_size}_i \\
+&\quad + \beta_{\text{treatment\_type}} \times \text{treatment\_type}_i
+\end{align*}
+
+Here, \alpha represents the intercept term, \beta coefficients denote the effect of respective predictors, and subscripts $i$ denote individual patients.
+
+Priors are specified for the model parameters as follows:
+
+\begin{align*}
+\alpha &\sim \text{Normal}(0, 2.5) \\
+\beta_{\text{pathologic_stage}}, \beta_{\text{lymph_node}}, \beta_{\text{metastasis}}, \beta_{\text{tumor_size}}, \beta_{\text{treatment_type}} &\sim \text{Normal}(0, 2.5) \\
+\sigma &\sim \text{Exponential}(1)
+\end{align*}
+
+TEMPLATE:
 Define $y_i$ as the number of seconds that the plane remained aloft. Then $\beta_i$ is the wing width and $\gamma_i$ is the wing length, both measured in millimeters.
@@
 \begin{align}
-y_i|\mu_i, \sigma &\sim \mbox{Normal}(\mu_i, \sigma) \\
+
 \mu_i &= \alpha + \beta_i + \gamma_i\\
@@
 
-We expect a positive relationship between the size of the wings and time spent aloft. In particular...
+We anticipate that the survival time of lung cancer patients post-diagnosis will be influenced by various clinical factors such as pathologic stage, extent of lymph node involvement, presence of distant metastasis, tumor size, and treatment type.
+Specifically, we expect that advanced pathologic stages, increased lymph node involvement, presence of distant metastasis, larger tumor sizes, and certain treatment types will be associated with shorter survival times.
 
-We can use maths by including latex between dollar signs, for instance $\theta$.
-
-The Bayesian analysis model serves as a powerful tool for investigating post-diagnosis survival outcomes in lung cancer patients.
-By incorporating pathogenic stage information and treatment modalities into the analysis, we aim to elucidate the factors influencing patient prognosis and provide insights to guide personalized treatment decisions.
+The linear regression model allows us to quantify the relationships between these predictors and survival outcomes, providing valuable insights into the factors influencing the prognosis of lung cancer patients.
 
@@
 library(ggplot2)
+library(arrow)
 
-# Scatter plot survival time vs. pathogenic stage - to compare survival probabilities
-ggplot(analysis_data, aes(x = pathogenic_stage, y = days_to_death)) +
+### read in data ###
+analysis_data <- read_parquet(file = here::here("data/analysis_data/lung_cancer_analysis_data.parquet"))
+
+# Scatter plot survival time vs. pathologic stage - to compare survival probabilities
+ggplot(analysis_data, aes(x = pathologic_stage, y = days_to_death)) +
   geom_point(size = 3, color = "lightpink") +
@@
   # Add a linear trend line
-  labs(title = "Survival Time vs. Pathogenic Stage Post-Diagnosis",
-       x = "Pathogenic Stage",
+  labs(title = "Survival Time vs. Pathologic Stage Post-Diagnosis",
+       x = "Pathologic Stage",
        y = "Survival Time (days)") +
@@
 # Line graph with trend lines 
-ggplot(analysis_data, aes(x = pathogenic_stage, y = days_to_death)) +
+ggplot(analysis_data, aes(x = pathologic_stage, y = days_to_death)) +
   geom_line(color = "blue", group = 1) +
@@
   # Add a linear trend line
-  labs(title = "Days to Death vs. Pathogenic Stage for Lung Cancer Patients",
-       x = "Pathogenic Stage",
+  labs(title = "Days to Death vs. Pathologic Stage for Lung Cancer Patients",
+       x = "Pathologic Stage",
        y = "Days to Death") +
@@
 # Plotting
-ggplot(analysis_data, aes(x = days_to_death, fill = pathogenic_stage)) +
+ggplot(analysis_data, aes(x = days_to_death, fill = pathologic_stage)) +
   geom_density(alpha = 0.5) +
   geom_density(color = "black", size = 1.5) +
-  labs(title = "Density Plot of Days to Death by Pathogenic Stage for Lung Cancer Patients",
+  labs(title = "Density Plot of Days to Death by Pathologic Stage for Lung Cancer Patients",
        x = "Days to Death",
@@
 # not sure .2
-ggplot(analysis_data, aes(x = days_to_death, fill = pathogenic_stage)) +
+ggplot(analysis_data, aes(x = days_to_death, fill = pathologic_stage)) +
   geom_density(alpha = 0.5) +
   geom_density(color = "black", size = 1.5) +
-  labs(title = "Density Plot of Days to Death by Pathogenic Stage for Lung Cancer Patients",
+  labs(title = "Density Plot of Days to Death by Pathologic Stage for Lung Cancer Patients",
        x = "Days to Death",
@@
   theme_minimal() +
-  facet_wrap(~pathogenic_stage, scales = "free")
+  facet_wrap(~pathologic_stage, scales = "free")
 
 # Boxplot 
-ggplot(analysis_data, aes(x = pathogenic_stage, y = days_to_death)) +
+ggplot(analysis_data, aes(x = pathologic_stage, y = days_to_death)) +
   geom_boxplot(fill = "lightblue") +
-  labs(title = "Survial Distribution Post-Diagnosis by Pathogenic Stage",
-       x = "Pathogenic Stage",
+  labs(title = "Survial Distribution Post-Diagnosis by Pathologic Stage",
+       x = "Pathologic Stage",
        y = "Days to Death") +
@@
 
-# Scatter plot survival time vs. Treatment Decision
-ggplot(analysis_data, aes(x = treatment_decision, y = days_to_death)) +
-  geom_point(size = 3, color = "blue") +
-  geom_smooth(method = "lm", se = FALSE, color = "red") +  
-  # Add a linear trend line
-  labs(title = "Survival Time vs. Treatment Decision Post-Diagnosis",
-       x = "Treatment Decision",
-       y = "Survival Time (days)") +
-  theme_minimal() +
-  theme(axis.text.x = element_text(angle = 45, hjust = 1)) + 
-  # Rotate x-axis labels for better readability
-  theme(text = element_text(family = "patrick") 
-  # Set font to Patrick Hand
-  )
 #########################
-
-# create bar chart for treatment decision
-treatment_decision_counts <- analysis_data %>%
-  group_by(treatment_decision) %>%
-  summarise(count = n ()) %>%
-  mutate(percentage = count / sum(count) * 100) %>%
-  arrange(desc(treatment_decision))
-
-ggplot(treatment_decision_counts, aes(x = reorder(treatment_decision, -percentage), y = percentage)) +
-  geom_bar(stat = "identity", fill = "blue") +
-  geom_text(aes(label = sprintf("%.1f%%", percentage)), hjust = -0.2, color = "black", size = 3) +
-  labs(title = "Percentage of Patients by Treatment Decision",
-       x = "Treatment Decision",
-       y = "Percentage") +
-  theme_minimal() +
-  coord_flip() +
-  theme(
-    text = element_text(family = "patrick") # Set font to Patrick Hand
-  )
 
@@
   )
-
-#######################
 

### 2024-04-05 paper/paper.tex
updating paper - working on modeling

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
  letterpaper,
  DIV=11,
  numbers=noendperiod]{scrartcl}

\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else  
    % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{5}
% Make \paragraph and \subparagraph free-standing
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

\KOMAoption{captions}{tableheading}
\makeatletter
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[skins,breakable]{tcolorbox}}
\makeatother
\makeatletter
\@ifundefined{shadecolor}{\definecolor{shadecolor}{rgb}{.97, .97, .97}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Survival Compass: Statistical Insights into Lung Cancer Patients' Journey Post-Diagnosis},
  pdfauthor={Lexi Knight},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\title{Survival Compass: Statistical Insights into Lung Cancer Patients'
Journey Post-Diagnosis\thanks{Code and data are available at:
https://github.com/LexiKnight/Lung\_Cancer/tree/main}}
\author{Lexi Knight}
\date{April 4, 2024}

\begin{document}
\maketitle
\begin{abstract}
This study investigates the impact of pathogenic stage and treatment
modalities on lung cancer survival post-diagnosis. Analysis of patient
data reveals significant correlation between pathogenic stage, seeking
treatment and survival outcomes. Notably, patients at advanced stages
with metastases in distant sites beyond the lung, extensive lymph node
involvement and tumors with extensive growth, invading nearby structures
demonstrate lower survival rates. These findings underscore the critical
importance of early detection, tailored treatment strategies and ongoing
research efforts to enhance lung cancer survival rates globally.
\end{abstract}
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[sharp corners, borderline west={3pt}{0pt}{shadecolor}, boxrule=0pt, interior hidden, breakable, frame hidden, enhanced]}{\end{tcolorbox}}\fi

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

Clinging to life amidst the shadows of lung cancer, where every breath
becomes a battleground. Survival becomes not just a statistic but an
interplay between several individual characteristics. We explore the
hidden keys to defying the odds and emerging victorious against one of
the deadliest adversaries of our time. Lung cancer is the leading cause
of cancer-related deaths in the world (Park, 2017). It is a disease that
develops in the lining of the airways in lung tissues. Non-small cell
lung cancer (NSCLC) is the most common type, accounting for 80-85\% of
all lung cancers according to the American Cancer society (Markman,
2023). Staging is important for prognosis and making treatment
decisions. Common treatments include surgery, radiation therapy and
chemotherapy (Kai, 2021). Pathogenic stage is determined by presence of
nearby metastasis, lymph node involvement as well as tumor spread and
size (Markman, 2023). This paper investigates the relationship between
lung cancer patients' survival and pathogenic stage. The estimand is the
median survival time in days post-diagnosis. We also look at whether
patients decided to have treatment and if so, which method; radiation
therapy or chemotherapy. Through analysis of a dataset made up of 981
patients in Sydney, Australia, we offer insight into the prognostic
markers.

Tumor size is often the main determinant of stage and treatment. As
tumor categories increase, the tumor expands, invading nearby structures
(Zhang, 2015). A study involving 52,287 patients diagnosed between the
years 1998 and 2003 found tumor size to be an independent prognostic
factor in estimating overall survival. The authors found that patients
presenting with larger tumors predicted a worse prognosis and thus are
associated with a decrease in survival. There is a similar relationship
between extensive lymph node involvement and patient survival (Zhang,
2015). Initial spread of cancer cells are localized, then become
regional, involving nearby lymph nodes and the most severe cases
comprises expansion to other organs such as the brain, liver and bones
(Markman, 2023). A study looked at five year survival rates based on the
severity of spread. 62.8\% of patients with localized spread, 34.8\% of
patients with regional and 8\% of patients with distant, advanced spread
were found to survive for 5 years post diagnosis. More than half of
these lung cancer patients have advanced spread to other organs when
diagnosed(Markman, 2023). Overall, it is found that patients with no
regional lymph node metastases, and smaller tumors are easier to be
treated and thus are associated with improved survival rates (Zhang,
2015).

Presence of metastatic LN is one of the most important determinants of
prognosis of NSCLC cases (Kai, 2021). In the early stage, cancer has not
spread to lymph nodes. As severity increases, lymph node metastasis
sequentially spreads to more distant lymph nodes such as mediastinal and
there is severe lymph node involvement (Park, 2017). Lymph node
involvement, also termed lymph node ratio, is a crucial factor in
guiding treatment options (Kai, 2021). A study made up of 97 patients
with a mean age of 63 who have undergone surgery between the years 2009
and 2015 in Korea find that increased lymph node involvement is
associated with a more advanced disease status and hence affiliated with
prognosis (Park, 2017). Another study looked at 11,341 NSCLC patients
between the years 2004 to 2015, from 18 geographically diverse
populations, covering approximately 28\% of the population of the United
States. These patients were treatment naive and underwent surgical
resection of the tumor. Although 5757 patients died, the rest showed
great results, with a median survival of 22 months (Kai, 2021). The
authors found that patients with low lymph node involvement lead to
higher survival compared to patients with high lymph node ratios. A
regression analysis revealed that lymph node ratio is an independent and
significant predictor of patient survival. The authors also observed
that disease burden and anatomical location of the lymph nodes involved
may influence the patients survival (Kai, 2021).

After tumor size, LN involvement and presence of distant metastasis are
categorized, the pathogenic stage of the cancer is then determined
(Eldridge, 2022). The most valuable prognostic factor in non-small cell
lung cancer is the pathogenic stage (Park, 2017) .Stage is determined by
tumor size, number of tumors and where the cancer has spread. Stage 1 is
localized spread, stage 2 and 3 is regional spread while stage 4 is
distant spread of the tumor (Eldridge, 2022). Cancer stage was
determined using the seventh American Joint Committee on Cancer staging
system (AJCC) (Park, 2017). A study done in Australia including 2119
lung cancer patients illustrated those with stage IV disease, the most
advanced stage, showed shorter survival than those at lower stages
(Denton, 2016). The earlier the cancer is found, that is the lower the
pathogenic stage, the greater the likelihood curative radiation therapy
is an effective treatment (Eldridge, 2022). However, there is minimal
literature looking at post-diagnosis survival rates based on pathogenic
stage and method of treatment. The extent of this disease illustrates
the importance of living a healthy lifestyle, undergoing regular
screening and development of improved treatment methods. Over the past
decade, there has been great improvement of lymph node assessment in
cancer patients (Kai, 2021). Experts hope survival rates continue to
improve with new therapies and treatment approaches (Markman, 2023).

The remainder of this paper is structured as follows. In
Section~\ref{sec-data}, we visualize the exploration of variables
constituting the pathogenic stage and treatment types.
Section~\ref{sec-model}, outlines the model employed to analyze the
relationship between these variables and the duration of survival
post-diagnosis. Moreover, Section~\ref{sec-results} offers visual
depictions of the study's outcomes. Finally, in
Section~\ref{sec-discussion}, we summarize the primary findings, propose
avenues for enhancement, and identify potential areas for future
research.

\hypertarget{sec-data}{%
\section{Data}\label{sec-data}}

Our data is

\begin{verbatim}
Warning: package 'ggplot2' was built under R version 4.2.3
\end{verbatim}

\begin{verbatim}
Warning: package 'dplyr' was built under R version 4.2.3
\end{verbatim}

\begin{verbatim}

Attaching package: 'dplyr'
\end{verbatim}

\begin{verbatim}
The following objects are masked from 'package:stats':

    filter, lag
\end{verbatim}

\begin{verbatim}
The following objects are masked from 'package:base':

    intersect, setdiff, setequal, union
\end{verbatim}

\begin{verbatim}
Warning: package 'showtext' was built under R version 4.2.3
\end{verbatim}

\begin{verbatim}
Loading required package: sysfonts
\end{verbatim}

\begin{verbatim}
Warning: package 'sysfonts' was built under R version 4.2.3
\end{verbatim}

\begin{verbatim}
Loading required package: showtextdb
\end{verbatim}

\begin{verbatim}
Warning: package 'showtextdb' was built under R version 4.2.3
\end{verbatim}

\begin{verbatim}
Warning: package 'arrow' was built under R version 4.2.3
\end{verbatim}

\begin{verbatim}

Attaching package: 'arrow'
\end{verbatim}

\begin{verbatim}
The following object is masked from 'package:utils':

    timestamp
\end{verbatim}

\begin{figure}

{\centering \includegraphics{paper_files/figure-pdf/fig-bills-1.pdf}

}

\caption{\label{fig-bills-1}bb}

\end{figure}

\begin{figure}

{\centering \includegraphics{paper_files/figure-pdf/fig-bills-2.pdf}

}

\caption{\label{fig-bills-2}bb}

\end{figure}

\begin{figure}

{\centering \includegraphics{paper_files/figure-pdf/fig-bills-3.pdf}

}

\caption{\label{fig-bills-3}bb}

\end{figure}

\begin{figure}

{\centering \includegraphics{paper_files/figure-pdf/fig-bills-4.pdf}

}

\caption{\label{fig-bills-4}bb}

\end{figure}

\begin{figure}

{\centering \includegraphics{paper_files/figure-pdf/fig-bills-5.pdf}

}

\caption{\label{fig-bills-5}bb}

\end{figure}

\begin{figure}

{\centering \includegraphics{paper_files/figure-pdf/fig-bills-6.pdf}

}

\caption{\label{fig-bills-6}bb}

\end{figure}

Talk more about it.

\hypertarget{sec-model}{%
\section{Model}\label{sec-model}}

\hypertarget{linear-regression-model}{%
\subsection{Linear Regression Model}\label{linear-regression-model}}

\hypertarget{interested-in-predicting-days_to_death-a-continuous-outcome-variable-based-on}{%
\section{interested in predicting days\_to\_death, a continuous outcome
variable based
on}\label{interested-in-predicting-days_to_death-a-continuous-outcome-variable-based-on}}

\hypertarget{multiple-predictor-variables-pathogenic_stage-lymph_node_involvement}{%
\section{multiple predictor variables (pathogenic\_stage,
lymph\_node\_involvement,}\label{multiple-predictor-variables-pathogenic_stage-lymph_node_involvement}}

\hypertarget{presence_of_distant_metastasis-tumor_size-treatment_decision-treatment_type}{%
\section{presence\_of\_distant\_metastasis, tumor\_size,
treatment\_decision,
treatment\_type)}\label{presence_of_distant_metastasis-tumor_size-treatment_decision-treatment_type}}

\hypertarget{linear-regression-model-are-suitable-for-predicting-continuous-outcomes}{%
\section{Linear regression model are suitable for predicting continuous
outcomes}\label{linear-regression-model-are-suitable-for-predicting-continuous-outcomes}}

\hypertarget{quantifying-relationship-between-predictor-variables-and-outcome-variable.}{%
\section{quantifying relationship between predictor variables and
outcome
variable.}\label{quantifying-relationship-between-predictor-variables-and-outcome-variable.}}

\hypertarget{fit-a-linear-regression-model-to-predict-continuous-outcome-variable}{%
\section{Fit a linear regression model to predict continuous outcome
variable}\label{fit-a-linear-regression-model-to-predict-continuous-outcome-variable}}

\hypertarget{days_to_death}{%
\section{`days\_to\_death'}\label{days_to_death}}

\hypertarget{model-set-up}{%
\subsection{Model set-up}\label{model-set-up}}

In this study, we aim to investigate the survival outcomes of lung
cancer patients post-diagnosis. Let \(y_i\) denote the number of days
from diagnosis to death for patient \(i\). We consider several
predictors including pathogenic stage, lymph node involvement, presence
of distant metastasis, tumor size, and treatment type.

We model the survival time \(y_i\) using a linear regression framework:

y\_i\textbar{}\mu\_i, \sigma \&\sim \mbox{Normal}(\mu\_i, \sigma)
\textbackslash{}

where \(mi_i\) is the expected survival time for patient \(i\), and
\(σ\) represents the standard deviation of the survival times. The
linear predictor \(mu_i\) is specified as:

\mu\emph{i = \alpha + \beta}\{\text{pathologic\_stage}\}
\times \text{pathologic\_stage}\emph{i + \beta}\{\text{lymph\_node}\}
\times \text{lymph\_node\_involvement}\emph{i +
\beta}\{\text{metastasis}\}
\times \text{presence\_of\_distant\_metastasis}\emph{i +
\beta}\{\text{tumor\_size}\} \times \text{tumor\_size}\emph{i +
\beta}\{\text{treatment\_type}\} \times \text{treatment\_type}\_i

Here, \alpha represents the intercept term, \beta coefficients denote
the effect of respective predictors, and subscripts \(i\) denote
individual patients.

Priors are specified for the model parameters as follows:

\begin{align*}
\alpha &\sim \text{Normal}(0, 2.5) \\
\beta_{\text{pathologic_stage}}, \beta_{\text{lymph_node}}, \beta_{\text{metastasis}}, \beta_{\text{tumor_size}}, \beta_{\text{treatment_type}} &\sim \text{Normal}(0, 2.5) \\
\sigma &\sim \text{Exponential}(1)
\end{align*}

TEMPLATE: Define \(y_i\) as the number of seconds that the plane
remained aloft. Then \(\beta_i\) is the wing width and \(\gamma_i\) is
the wing length, both measured in millimeters.

\begin{align}

\mu_i &= \alpha + \beta_i + \gamma_i\\
\alpha &\sim \mbox{Normal}(0, 2.5) \\
\beta &\sim \mbox{Normal}(0, 2.5) \\
\gamma &\sim \mbox{Normal}(0, 2.5) \\
\sigma &\sim \mbox{Exponential}(1)
\end{align}

We run the model in R (R Core Team 2023) using the \texttt{rstanarm}
package of Goodrich et al. (2022). We use the default priors from
\texttt{rstanarm}.

\hypertarget{model-justification}{%
\subsubsection{Model justification}\label{model-justification}}

We anticipate that the survival time of lung cancer patients
post-diagnosis will be influenced by various clinical factors such as
pathologic stage, extent of lymph node involvement, presence of distant
metastasis, tumor size, and treatment type. Specifically, we expect that
advanced pathologic stages, increased lymph node involvement, presence
of distant metastasis, larger tumor sizes, and certain treatment types
will be associated with shorter survival times.

The linear regression model allows us to quantify the relationships
between these predictors and survival outcomes, providing valuable
insights into the factors influencing the prognosis of lung cancer
patients.

\hypertarget{sec-results}{%
\section{Results}\label{sec-results}}

Our results are summarized in @.

\includegraphics{paper_files/figure-pdf/unnamed-chunk-4-1.pdf}

\includegraphics{paper_files/figure-pdf/unnamed-chunk-4-2.pdf}

\includegraphics{paper_files/figure-pdf/unnamed-chunk-4-3.pdf}

\includegraphics{paper_files/figure-pdf/unnamed-chunk-4-4.pdf}

\includegraphics{paper_files/figure-pdf/unnamed-chunk-4-5.pdf}

\includegraphics{paper_files/figure-pdf/unnamed-chunk-4-6.pdf}

\includegraphics{paper_files/figure-pdf/unnamed-chunk-4-7.pdf}

\includegraphics{paper_files/figure-pdf/unnamed-chunk-4-8.pdf}

\includegraphics{paper_files/figure-pdf/unnamed-chunk-4-9.pdf}

\includegraphics{paper_files/figure-pdf/unnamed-chunk-4-10.pdf}

\includegraphics{paper_files/figure-pdf/unnamed-chunk-4-11.pdf}

\includegraphics{paper_files/figure-pdf/unnamed-chunk-4-12.pdf}

\includegraphics{paper_files/figure-pdf/unnamed-chunk-4-13.pdf}

\includegraphics{paper_files/figure-pdf/unnamed-chunk-4-14.pdf}

\includegraphics{paper_files/figure-pdf/unnamed-chunk-4-15.pdf}

\includegraphics{paper_files/figure-pdf/unnamed-chunk-4-16.pdf}

\includegraphics{paper_files/figure-pdf/unnamed-chunk-4-17.pdf}

\begin{table}

\end{table}

\hypertarget{sec-discussion}{%
\section{Discussion}\label{sec-discussion}}

\hypertarget{sec-first-point}{%
\subsection{First discussion point}\label{sec-first-point}}

\hypertarget{second-discussion-point}{%
\subsection{Second discussion point}\label{second-discussion-point}}

\hypertarget{third-discussion-point}{%
\subsection{Third discussion point}\label{third-discussion-point}}

\hypertarget{weaknesses-and-next-steps}{%
\subsection{Weaknesses and next steps}\label{weaknesses-and-next-steps}}

\newpage

\appendix

\hypertarget{sec-appendix}{%
\section{Appendix}\label{sec-appendix}}

\hypertarget{additional-data-details}{%
\section{Additional data details}\label{additional-data-details}}

\hypertarget{sec-model-details}{%
\section{Model details}\label{sec-model-details}}

we compare the posterior with the prior. This shows\ldots{}

\begin{figure}

\begin{minipage}[t]{0.50\linewidth}

{\centering

Examining how the model fits, and is affected by, the data

}

\end{minipage}%

\caption{\label{fig-ppcheckandposteriorvsprior}\textbf{?(caption)}}

\end{figure}

\hypertarget{diagnostics}{%
\subsection{Diagnostics}\label{diagnostics}}

Is this needed?

\begin{figure}

\begin{minipage}[t]{0.50\linewidth}

{\centering

Checking the convergence of the MCMC algorithm

}

\end{minipage}%

\caption{\label{fig-stanareyouokay}\textbf{?(caption)}}

\end{figure}

\newpage

\hypertarget{references}{%
\section*{References}\label{references}}
\addcontentsline{toc}{section}{References}

\hypertarget{refs}{}
\begin{CSLReferences}{1}{0}
\leavevmode\vadjust pre{\hypertarget{ref-rstanarm}{}}%
Goodrich, Ben, Jonah Gabry, Imad Ali, and Sam Brilleman. 2022.
{``Rstanarm: {Bayesian} Applied Regression Modeling via {Stan}.''}
\url{https://mc-stan.org/rstanarm/}.

\leavevmode\vadjust pre{\hypertarget{ref-citeR}{}}%
R Core Team. 2023. \emph{R: A Language and Environment for Statistical
Computing}. Vienna, Austria: R Foundation for Statistical Computing.
\url{https://www.R-project.org/}.

\end{CSLReferences}

\end{document}

### 2024-04-05 scripts/04-model.R

@@ -7,7 +7,6 @@
 # Pre-requisites: follow 01-download_data.R and run 02-data_cleaning.R in scripts in order to obtain raw and cleaned data. Run 03-test_data.R to test cleaned dataset.  
 
 # note about approach: linear model (ch. 13.2, logistic regression model )
-# note, must delete treatment_decision from entire analysis. 
 
 #### Workspace setup ####
 # install packages

### 2024-04-15 paper/paper.qmd
Fixed modeling section (formating) and working on data section (started writing)

@@
 ---
-title: "Survival Compass: Statistical Insights into Lung Cancer Patients' Journey Post-Diagnosis"
+title: "Survival Compass"
+subtitle: "Statistical Insights into Lung Cancer Patients Journey Post Diagnosis"
 author: 
@@
 
-## Linear Regression Model 
-# interested in predicting days_to_death, a continuous outcome variable based on
-# multiple predictor variables  (pathogenic_stage, lymph_node_involvement, 
-# presence_of_distant_metastasis, tumor_size, treatment_decision, treatment_type)
-# Linear regression model are suitable for predicting continuous outcomes
-# quantifying relationship between predictor variables and outcome variable.
+## Model set-up
 
-# Fit a linear regression model to predict continuous outcome variable 
-# 'days_to_death'
+In this section, we aim to predict the survival outcomes of lung cancer patients post-diagnosis with a linear regression model framework.
+We consider several predictors including pathogenic stage, lymph node involvement, presence of distant metastasis, tumor size, and treatment type.We specify the model and subsequently justify its appropriateness for our analysis.
 
-## Model set-up
-In this study, we aim to investigate the survival outcomes of lung cancer patients post-diagnosis.
-Let $y_i$ denote the number of days from diagnosis to death for patient $i$.
-We consider several predictors including pathogenic stage, lymph node involvement, presence of distant metastasis, tumor size, and treatment type.
+### Model Specifications ###
 
-We model the survival time $y_i$ using a linear regression framework where $mi_i$ is the expected survival time for patient $i$, and $σ$ represents the standard deviation of the survival times.
+We employ a linear regression model to predict the number of days from diagnosis to death for each lung cancer patient. The model is defined as follows:
+
+\[
+y_i \mid \mu_i, \sigma \sim \text{Normal}(\mu_i, \sigma)
+\]
+
+where:
+\begin{itemize}
+  \item $y_i$ represents the number of days from diagnosis to death for patient $i$.
+  \item $mu_i$ denotes the expected number of days to death for patient $i$.
+  \item $σ$ represents the standard deviation of the survival times.
+\end{itemize}
+
 The linear predictor $mu_i$ is specified as:
@@
 \begin{align*}
-y_i | \mu_i, \sigma &\sim \text{Normal}(\mu_i, \sigma) \\
-\mu_i &= \alpha + \beta_{\text{pathologic\_stage}} \times \text{pathologic\_stage}_i + \beta_{\text{lymph\_node}} \times \text{lymph\_node\_involvement}_i \\
-&\quad + \beta_{\text{metastasis}} \times \text{presence\_of\_distant\_metastasis}_i + \beta_{\text{tumor\_size}} \times \text{tumor\_size}_i \\
+y_i \mid \mu_i, \sigma &\sim \text{Normal}(\mu_i, \sigma) \\
+\mu_i &= \alpha + \beta_{\text{pathologic\_stage}} \times \text{pathologic\_stage}_i \\
+&\quad + \beta_{\text{lymph\_node}} \times \text{lymph\_node\_involvement}_i \\
+&\quad + \beta_{\text{metastasis}} \times \text{presence\_of\_distant\_metastasis}_i \\
+&\quad + \beta_{\text{tumor\_size}} \times \text{tumor\_size}_i \\
 &\quad + \beta_{\text{treatment\_type}} \times \text{treatment\_type}_i
@@
 
-Here, \alpha represents the intercept term, \beta coefficients denote the effect of respective predictors, and subscripts $i$ denote individual patients.
-
-Priors are specified for the model parameters as follows:
-
-\begin{align*}
-\alpha &\sim \text{Normal}(0, 2.5) \\
-\beta_{\text{pathologic_stage}}, \beta_{\text{lymph_node}}, \beta_{\text{metastasis}}, \beta_{\text{tumor_size}}, \beta_{\text{treatment_type}} &\sim \text{Normal}(0, 2.5) \\
-\sigma &\sim \text{Exponential}(1)
-\end{align*}
-
-TEMPLATE:
-Define $y_i$ as the number of seconds that the plane remained aloft. Then $\beta_i$ is the wing width and $\gamma_i$ is the wing length, both measured in millimeters.
-
-\begin{align}
-
-\mu_i &= \alpha + \beta_i + \gamma_i\\
-\alpha &\sim \mbox{Normal}(0, 2.5) \\
-\beta &\sim \mbox{Normal}(0, 2.5) \\
-\gamma &\sim \mbox{Normal}(0, 2.5) \\
-\sigma &\sim \mbox{Exponential}(1)
-\end{align}
-
-We run the model in R [@citeR] using the `rstanarm` package of @rstanarm. We use the default priors from `rstanarm`.
+where: 
+\begin{itemize}
+  \item \alpha represents the intercept term, capturing the baseline number of days to death.
+  \item \beta_{\text{pathologic_stage}}, \beta_{\text{lymph_node}}, \beta_{\text{metastasis}}, \beta_{\text{tumor_size}}, \beta_{\text{treatment_type}} are the coefficients associated with each predictor variable.
+\end{itemize}
 
 ### Model justification
+
+Linear regression models are most appropritate in predicting continuous outcomes.
+As survival time is continuous, this model allows us to quantify the relationships between these predictors and survival outcomes, providing valuable insights into the factors influencing the prognosis of lung cancer patients.
+
+#### Response Variable ####
+
+Out variable of interest is survival time in lung cancer patient afte they have been diagnosed
+
+We model the survival time ($y_i$) as a continuous variable, reflecting the duration from diagnosis to death for each patient.
+This continuous characterization is appropriate for capturing the temporal aspect of survival outcomes in medical contexts.
+
+#### Input Variables ####
+
+We consider several clinically relevant predictors including pathologic stage, lymph node involvement, presence of distant metastasis, tumor size, and treatment type.
+These variables are chosen based on their established associations with lung cancer prognosis, encompassing key aspects of disease severity and treatment strategies.
+
+#### Model Structure #####
+
+The linear regression model relates the expected survival time ($μ_i$) to the linear combination of predictor variables, allowing us to quantify the impact of each predictor on the expected duration of survival.
+This framework facilitates interpretation of the associations between clinical variables and survival outcomes, providing valuable insights for patient prognosis.
+
+#### Parameter Estimation ####
 
@@
 
-The linear regression model allows us to quantify the relationships between these predictors and survival outcomes, providing valuable insights into the factors influencing the prognosis of lung cancer patients.
+We run the model in R [@citeR] estimating the model coefficients ($/alpha$ and $/beta$) using Bayesian inference via the 'stan_glm()' function from the @rstanarm package.
+This approach leverages Markov Chain Monte Carlo (MCMC) algorithms to obtain posterior distributions for the model parameters, enabling robust estimation of parameter uncertainties and inference on the effects of predictor variables.
 

### 2024-04-15 paper/paper.tex
Fixed modeling section (formating) and working on data section (started writing)

@@
 \hypersetup{
-  pdftitle={Survival Compass: Statistical Insights into Lung Cancer Patients' Journey Post-Diagnosis},
+  pdftitle={Survival Compass},
   pdfauthor={Lexi Knight},
@@
 
-\title{Survival Compass: Statistical Insights into Lung Cancer Patients'
-Journey Post-Diagnosis\thanks{Code and data are available at:
+\title{Survival Compass\thanks{Code and data are available at:
 https://github.com/LexiKnight/Lung\_Cancer/tree/main}}
+\usepackage{etoolbox}
+\makeatletter
+\providecommand{\subtitle}[1]{% add subtitle to \maketitle
+  \apptocmd{\@title}{\par {\large #1 \par}}{}{}
+}
+\makeatother
+\subtitle{Statistical Insights into Lung Cancer Patients Journey Post
+Diagnosis}
 \author{Lexi Knight}
-\date{April 4, 2024}
+\date{April 5, 2024}
 
@@
 \end{abstract}
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[sharp corners, borderline west={3pt}{0pt}{shadecolor}, boxrule=0pt, interior hidden, breakable, frame hidden, enhanced]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[frame hidden, borderline west={3pt}{0pt}{shadecolor}, enhanced, interior hidden, breakable, sharp corners, boxrule=0pt]}{\end{tcolorbox}}\fi
 
@@
 
-\hypertarget{linear-regression-model}{%
-\subsection{Linear Regression Model}\label{linear-regression-model}}
-
-\hypertarget{interested-in-predicting-days_to_death-a-continuous-outcome-variable-based-on}{%
-\section{interested in predicting days\_to\_death, a continuous outcome
-variable based
-on}\label{interested-in-predicting-days_to_death-a-continuous-outcome-variable-based-on}}
-
-\hypertarget{multiple-predictor-variables-pathogenic_stage-lymph_node_involvement}{%
-\section{multiple predictor variables (pathogenic\_stage,
-lymph\_node\_involvement,}\label{multiple-predictor-variables-pathogenic_stage-lymph_node_involvement}}
-
-\hypertarget{presence_of_distant_metastasis-tumor_size-treatment_decision-treatment_type}{%
-\section{presence\_of\_distant\_metastasis, tumor\_size,
-treatment\_decision,
-treatment\_type)}\label{presence_of_distant_metastasis-tumor_size-treatment_decision-treatment_type}}
-
-\hypertarget{linear-regression-model-are-suitable-for-predicting-continuous-outcomes}{%
-\section{Linear regression model are suitable for predicting continuous
-outcomes}\label{linear-regression-model-are-suitable-for-predicting-continuous-outcomes}}
-
-\hypertarget{quantifying-relationship-between-predictor-variables-and-outcome-variable.}{%
-\section{quantifying relationship between predictor variables and
-outcome
-variable.}\label{quantifying-relationship-between-predictor-variables-and-outcome-variable.}}
-
-\hypertarget{fit-a-linear-regression-model-to-predict-continuous-outcome-variable}{%
-\section{Fit a linear regression model to predict continuous outcome
-variable}\label{fit-a-linear-regression-model-to-predict-continuous-outcome-variable}}
-
-\hypertarget{days_to_death}{%
-\section{`days\_to\_death'}\label{days_to_death}}
-
 \hypertarget{model-set-up}{%
@@
 
-In this study, we aim to investigate the survival outcomes of lung
-cancer patients post-diagnosis. Let \(y_i\) denote the number of days
-from diagnosis to death for patient \(i\). We consider several
-predictors including pathogenic stage, lymph node involvement, presence
-of distant metastasis, tumor size, and treatment type.
+In this section, we aim to predict the survival outcomes of lung cancer
+patients post-diagnosis with a linear regression model framework. We
+consider several predictors including pathogenic stage, lymph node
+involvement, presence of distant metastasis, tumor size, and treatment
+type.We specify the model and subsequently justify its appropriateness
+for our analysis.
 
-We model the survival time \(y_i\) using a linear regression framework:
+\hypertarget{model-specifications}{%
+\subsubsection{Model Specifications}\label{model-specifications}}
 
-y\_i\textbar{}\mu\_i, \sigma \&\sim \mbox{Normal}(\mu\_i, \sigma)
-\textbackslash{}
+We employ a linear regression model to predict the number of days from
+diagnosis to death for each lung cancer patient. The model is defined as
+follows:
 
-where \(mi_i\) is the expected survival time for patient \(i\), and
-\(σ\) represents the standard deviation of the survival times. The
-linear predictor \(mu_i\) is specified as:
+\(y_i\) \textbar{} \mu\_i, \sigma \&\sim \text{Normal}(\mu\_i, \sigma)
 
-\mu\emph{i = \alpha + \beta}\{\text{pathologic\_stage}\}
-\times \text{pathologic\_stage}\emph{i + \beta}\{\text{lymph\_node}\}
-\times \text{lymph\_node\_involvement}\emph{i +
+where: * \(y_i\) represents the number of days from diagnosis to death
+for patient \(i\). * \(mu_i\) denotes the expected number of days to
+death for patient \(i\). * \(σ\) represesents the standard deviation of
+the survival times.
+
+The linear predictor \(mu_i\) is specified as:
+
+\mu\emph{i = \alpha + \beta}\{\text{pathologic_stage}\}
+\times \text{pathologic_stage}\emph{i + \beta}\{\text{lymph_node}\}
+\times \text{lymph_node_involvement}\emph{i +
 \beta}\{\text{metastasis}\}
-\times \text{presence\_of\_distant\_metastasis}\emph{i +
-\beta}\{\text{tumor\_size}\} \times \text{tumor\_size}\emph{i +
-\beta}\{\text{treatment\_type}\} \times \text{treatment\_type}\_i
+\times \text{presence_of_distant_metastasis}\emph{i +
+\beta}\{\text{tumor_size}\} \times \text{tumor_size}\emph{i +
+\beta}\{\text{treatment_type}\} \times \text{treatment_type}\_i
 
-Here, \alpha represents the intercept term, \beta coefficients denote
-the effect of respective predictors, and subscripts \(i\) denote
-individual patients.
+where:
 
-Priors are specified for the model parameters as follows:
-
-\begin{align*}
-\alpha &\sim \text{Normal}(0, 2.5) \\
-\beta_{\text{pathologic_stage}}, \beta_{\text{lymph_node}}, \beta_{\text{metastasis}}, \beta_{\text{tumor_size}}, \beta_{\text{treatment_type}} &\sim \text{Normal}(0, 2.5) \\
-\sigma &\sim \text{Exponential}(1)
-\end{align*}
-
-TEMPLATE: Define \(y_i\) as the number of seconds that the plane
-remained aloft. Then \(\beta_i\) is the wing width and \(\gamma_i\) is
-the wing length, both measured in millimeters.
-
-\begin{align}
-
-\mu_i &= \alpha + \beta_i + \gamma_i\\
-\alpha &\sim \mbox{Normal}(0, 2.5) \\
-\beta &\sim \mbox{Normal}(0, 2.5) \\
-\gamma &\sim \mbox{Normal}(0, 2.5) \\
-\sigma &\sim \mbox{Exponential}(1)
-\end{align}
-
-We run the model in R (R Core Team 2023) using the \texttt{rstanarm}
-package of Goodrich et al. (2022). We use the default priors from
-\texttt{rstanarm}.
+\begin{itemize}
+\tightlist
+\item
+  \alpha represents the intercept term, capturing the baseline number of
+  days to death.
+\item
+  \beta\emph{\{\text{pathologic_stage}\}, \beta}\{\text{lymph_node}\},
+  \beta\emph{\{\text{metastasis}\}, \beta}\{\text{tumor_size}\},
+  \beta\_\{\text{treatment_type}\} are the coefficients associated with
+  each predictor variable.
+\end{itemize}
 
@@
 \subsubsection{Model justification}\label{model-justification}}
+
+Linear regression models are most appropritate in predicting continuous
+outcomes. As survival time is continuous, this model allows us to
+quantify the relationships between these predictors and survival
+outcomes, providing valuable insights into the factors influencing the
+prognosis of lung cancer patients.
+
+\hypertarget{response-variable}{%
+\paragraph{Response Variable}\label{response-variable}}
+
+Out variable of interest is survival time in lung cancer patient afte
+they have been diagnosed
+
+We model the survival time (\(y_i\)) as a continuous variable,
+reflecting the duration from diagnosis to death for each patient. This
+continuous characterization is appropriate for capturing the temporal
+aspect of survival outcomes in medical contexts.
+
+\hypertarget{input-variables}{%
+\paragraph{Input Variables}\label{input-variables}}
+
+We consider several clinically relevant predictors including pathologic
+stage, lymph node involvement, presence of distant metastasis, tumor
+size, and treatment type. These variables are chosen based on their
+established associations with lung cancer prognosis, encompassing key
+aspects of disease severity and treatment strategies.
+
+\hypertarget{model-structure}{%
+\paragraph{Model Structure}\label{model-structure}}
+
+The linear regression model relates the expected survival time (\(μ_i\))
+to the linear combination of predictor variables, allowing us to
+quantify the impact of each predictor on the expected duration of
+survival. This framework facilitates interpretation of the associations
+between clinical variables and survival outcomes, providing valuable
+insights for patient prognosis.
+
+\hypertarget{parameter-estimation}{%
+\paragraph{Parameter Estimation}\label{parameter-estimation}}
 
@@
 
-The linear regression model allows us to quantify the relationships
-between these predictors and survival outcomes, providing valuable
-insights into the factors influencing the prognosis of lung cancer
-patients.
+We run the model in R (R Core Team 2023) estimating the model
+coefficients (\(/alpha\) and \(/beta\)) using Bayesian inference via the
+`stan\_glm()' function from the Goodrich et al. (2022) package. This
+approach leverages Markov Chain Monte Carlo (MCMC) algorithms to obtain
+posterior distributions for the model parameters, enabling robust
+estimation of parameter uncertainties and inference on the effects of
+predictor variables.
 

### 2024-04-15 paper/paper.qmd
cleaning up data section of paper

@@
 These findings underscore the critical importance of early detection, tailored treatment strategies and ongoing research efforts to enhance lung cancer survival rates globally."
-format: pdf
+format:
+  pdf:
+    toc: true
 number-sections: true
@@
 
-Our data is
+## Software and R-packages
+This project was created using statistical software, `R` [@citeR].
+For data cleaning and manipulation, we used `tidyverse` [@tidy] package.
+For data visualization such as creating the figures, we used `ggplot2` [@ggplot2].
+For converting into Parquet file, we used `arrow` [@arrow], managing file paths `here` [@here], statistical modeling `rstanarm` [@rstanarm].
+Furthermore, `showtext` [@showtext] was used for aesthetic purposes.
+
+## Methodology
+
+The data for this study were collected from a comprehensive database comprising 981 lung cancer patients diagnosed between 1991 to 2013 from `Center for Open Science` [@cfos], a dataset acquired in Sydney Australia.
+The dataset included information on patient demographics, clinical characteristics, treatment modalities, and survival outcomes.
+
+### Data Collection
+We obtained data on lung cancer patients meeting the following criteria: histologically confirmed lung cancer diagnosis, availability of complete clinical data, treatment-naive patients, single malignancy and located in Australia.
+Patients with missing or incomplete information were excluded from the analysis.
+
+### Data Cleaning
+After obtaining the dataset, there were a series of rigorous preprocessing steps undertaken to ensure data quality and consistency.
+This included selecting the columns of interest namely; days to death post diagnosis, presence of distant metastasis, lymph node involvement, pathogenic stage, tumor size and treatment type.
+Next, we cleaned the column names and excluded missing values.
+Additionally, we converted the days to death column to numeric.Tests where included to ensure accuracy, reliability and validity of the datset for subsequent analysis and interpretation.
+
+### Data Analysis
+Descriptive and inferential statistical analyses were conducted to explore the dataset and derive meaningful insights. These included linear regression modeling.
+
+## Features
+The dataset comprised several key features relevant to lung cancer prognosis, including pathogenic stage, presence of distant metastasis, lymph node involvement, tumor size, and treatment type.
+
+### Days to Death
 
 ```{r}
-#| label: fig-bills
-#| fig-cap: bb
+#| warning: false
 #| echo: false
+#| message: false
+#| label: fig-one
+#| fig-cap: Distribution of survival time in patients
 
@@
   )
+
+```
+
+One of the critical features analyzed in this study is the duration between the date of lung cancer diagnosis and the date of death, referred to as "days to death." This metric serves as a key indicator of patient survival and provides valuable insights into the disease trajectory and prognosis.
+By examining the distribution of survival times among lung cancer patients post-diagnosis, we aim to characterize the temporal patterns of disease progression and assess the impact of various clinical and demographic factors on survival outcomes.
+Understanding the time course from diagnosis to death is crucial for guiding treatment decisions, predicting patient outcomes, and identifying opportunities for intervention to improve survival rates.
+Through comprehensive analysis of days to death data, we seek to elucidate the factors influencing patient survival in lung cancer and contribute to the refinement of prognostic models for clinical practice.
+
+### Pathogenic Stage
+Pathogenic stage, a critical determinant of lung cancer prognosis, was classified according to the [mention staging system used, e.g., TNM classification].
+As depicted in Figure 1, the distribution of patients varied across different stages, with [mention any notable trends, e.g., a higher proportion of patients diagnosed at advanced stages].
+
+### Presence of Distant Metastasis
+The presence of distant metastasis, indicative of disease spread beyond the primary tumor site, significantly influences treatment decisions and patient outcomes.
+Figure 2 illustrates the percentage of patients with and without distant metastasis, highlighting the impact of metastatic disease on prognosis.
+
+### Lymph Node Involvement
+Lymph node involvement is a key prognostic factor in lung cancer, reflecting the extent of disease spread to regional lymph nodes.
+Figure 3 presents the distribution of patients based on lymph node involvement, demonstrating its association with survival outcomes.
+
+### Tumor Size
+Tumor size, often measured as the diameter of the primary tumor, is closely linked to disease progression and treatment response in lung cancer patients.
+Figure 4 showcases the distribution of patients across different tumor size categories, elucidating its significance as a prognostic factor.
+
+```{r}
+#| warning: false
+#| echo: false
+#| message: false
+#| label: fig-bills
+#| fig-cap: four
+
+# load packages
+library(ggplot2)
+library(dplyr)
+library(showtext)
+library(arrow)
+
+### read in data ###
+analysis_data <- read_parquet(file = here::here("data/analysis_data/lung_cancer_analysis_data.parquet"))
+
+# Add the desired font (e.g., Patrick Hand)
+font_add_google("Patrick Hand", "patrick")
+
+# Set the font for ggplot
+showtext_auto()
 
@@
 
-# Create bar chart for treatment type
-treatment_type_counts <- analysis_data %>%
-  group_by(treatment_type) %>%
-  summarise(count = n ()) %>%
-  mutate(percentage = count / sum(count) * 100) %>%
-  arrange(desc(treatment_type))
-
-ggplot(treatment_type_counts, aes(x = reorder(treatment_type, -percentage), y = percentage)) +
-  geom_bar(stat = "identity", fill = "orange") +
-  geom_text(aes(label = sprintf("%.1f%%", percentage)), hjust = -0.2, color = "black", size = 3) +
-  geom_text(data = treatment_type_counts, aes(label = treatment_type), hjust = 1.2, color = "black", size = 3) +
-  labs(title = "Percentage of Patients by Treatment Type",
-       x = "Treatment Type",
-       y = "Percentage") +
-  theme_minimal() +
-  coord_flip() +
-  scale_x_discrete(breaks = seq(0, max(treatment_type_counts$percentage), by = 5)) +
-  theme(
-    text = element_text(family = "patrick") # Set font to Patrick Hand
-  )
-
 ```
@@
 Talk more about it.
-
-```{r}
-#| label: fig-planes
-#| fig-cap: Relationship between wing length and width
-#| echo: false
-#| warning: false
-#| message: false
-
-```
 
@@
 We employ a linear regression model to predict the number of days from diagnosis to death for each lung cancer patient. The model is defined as follows:
+\begin{align*}
+y_i \mid \mu_i, \sigma &\sim \text{Normal}(\mu_i, \sigma)
+\end{align*}
+where:
 
-\[
-y_i \mid \mu_i, \sigma \sim \text{Normal}(\mu_i, \sigma)
-\]
+- $y_i$ represents the number of days from diagnosis to death for patient $i$.
+- $\mu_i$ denotes the expected number of days to death for patient $i$.
+- $\sigma$ represents the standard deviation of the survival times.
 
-where:
-\begin{itemize}
-  \item $y_i$ represents the number of days from diagnosis to death for patient $i$.
-  \item $mu_i$ denotes the expected number of days to death for patient $i$.
-  \item $σ$ represents the standard deviation of the survival times.
-\end{itemize}
-
-The linear predictor $mu_i$ is specified as:
-
+The linear predictor $\mu_i$ is specified as:
 \begin{align*}
@@
 \end{align*}
+where:
 
-where: 
-\begin{itemize}
-  \item \alpha represents the intercept term, capturing the baseline number of days to death.
-  \item \beta_{\text{pathologic_stage}}, \beta_{\text{lymph_node}}, \beta_{\text{metastasis}}, \beta_{\text{tumor_size}}, \beta_{\text{treatment_type}} are the coefficients associated with each predictor variable.
-\end{itemize}
+- $\alpha$ represents the intercept term, capturing the baseline number of days to death.
+- $\beta$ {\text{pathologic\_stage}}, $\beta$ {\text{lymph\_node}}, $\beta$ {\text{metastasis}}, $\beta$ {\text{tumor\_size}}, $\beta$ {\text{treatment\_type}} are the coefficients associated with each predictor variable.
 
@@
 
-We run the model in R [@citeR] estimating the model coefficients ($/alpha$ and $/beta$) using Bayesian inference via the 'stan_glm()' function from the @rstanarm package.
+We run the model in R [@citeR] estimating the model coefficients ($\alpha$ and $\beta$) using Bayesian inference via the 'stan_glm()' function from the @rstanarm package.
 This approach leverages Markov Chain Monte Carlo (MCMC) algorithms to obtain posterior distributions for the model parameters, enabling robust estimation of parameter uncertainties and inference on the effects of predictor variables.

### 2024-04-16 paper/paper.qmd
Graphs in data section are all done! Whoop Whoop! Now need to finish writing data section.

@@
 
-### Days to Death
-
 ```{r}
@@
 
+### Days to Death
+
 One of the critical features analyzed in this study is the duration between the date of lung cancer diagnosis and the date of death, referred to as "days to death." This metric serves as a key indicator of patient survival and provides valuable insights into the disease trajectory and prognosis.
@@
 #| message: false
-#| label: fig-bills
-#| fig-cap: four
+#| label: Figure 2
+#| fig-cap: ADD FIGURE 2 CAPTION HERE 
+#| fig-title: "Percentage of Lung Cancer Patients by:"
+
+# install.packages("patchwork")
 
@@
 library(arrow)
+library(patchwork)
 
@@
 
-ggplot(pathologic_stage_counts, aes(x = pathologic_stage, y = percentage)) +
-  geom_bar(stat = "identity", fill = "lightpink") +
-  labs(title = "Percentage of Patients by Pathologic Stage",
-       x = "Pathologic Stage",
-       y = "Percentage") +
-  theme_minimal() +
-  theme(
-    text = element_text(family = "patrick") # Set font to Patrick Hand
-  )
-
 # create bar chart for presence_of_distant metastasis 
@@
 
-ggplot(presence_of_distant_metastasis_counts, aes(x = presence_of_distant_metastasis, y = percentage)) +
-  geom_bar(stat = "identity", fill = "gold") +
-  labs(title = "Percentage of Patients by Presence of Distant Metastasis",
-       x = "Presence of Distant Metastasis",
-       y = "Percentage") +
-  theme_minimal() +
-  theme(
-    text = element_text(family = "patrick") # Set font to Patrick Hand
-  )
+# Define a minimum height for the bars
+min_bar_height <- 0.1  # Adjust as needed based on visibility preference
 
@@
 
-ggplot(lymph_node_involvement_counts, aes(x = lymph_node_involvement, y = percentage)) +
-  geom_bar(stat = "identity", fill = "cyan") +
-  labs(title = "Percentage of Patients by Lymph Node Involvement",
-       x = "Lymph Node Involvement",
-       y = "Percentage") +
-  theme_minimal() +
-  theme(
-    text = element_text(family = "patrick") # Set font to Patrick Hand
-  )
-
 # create bar chart for tumor size
@@
 
-ggplot(tumor_size_counts, aes(x = tumor_size, y = percentage)) +
-  geom_bar(stat = "identity", fill = "purple") +
-  labs(title = "Percentage of Patients by Tumor Size",
-       x = "Tumor Size",
+# Create subplots for all four plots
+plots <- list(
+  ggplot(pathologic_stage_counts, aes(x = pathologic_stage, y = percentage)) +
+    geom_bar(stat = "identity", fill = "lightpink") +
+    labs(title = "Pathologic Stage",
+         x = "",
          y = "Percentage") +
@@
     theme(
+      text = element_text(family = "patrick"), # Set font to Patrick Hand
+      axis.text.x = element_text(angle = 45, hjust = 1) # Rotate x-axis labels by 45 degrees
+    ) +
+    scale_y_continuous(breaks = seq(0, max(pathologic_stage_counts$percentage), by = 5)),
+
+   ggplot(presence_of_distant_metastasis_counts, aes(x = presence_of_distant_metastasis, y = percentage)) +
+    geom_bar(stat = "identity", fill = "gold") +
+    labs(title = "Presence of Distant Metastasis",
+         x = "",
+         y = "") +  # Remove y-axis label 
+    theme_minimal() +
+    theme(
+      text = element_text(family = "patrick"), # Set font to Patrick Hand
+      axis.title.y = element_blank(), # Remove y-axis title
+    ),
+
+   ggplot(lymph_node_involvement_counts, aes(x = lymph_node_involvement, y = percentage)) +
+    geom_bar(stat = "identity", fill = "cyan") +
+    labs(title = "Lymph Node Involvement",
+         x = "",
+         y = "Percentage") +  # Keep y-axis label for this plot
+    theme_minimal() +
+    theme(
       text = element_text(family = "patrick") # Set font to Patrick Hand
+    ) +
+    scale_y_continuous(breaks = seq(0, max(lymph_node_involvement_counts$percentage), by = 10)),
+
+  ggplot(tumor_size_counts, aes(x = tumor_size, y = percentage)) +
+    geom_bar(stat = "identity", fill = "purple") +
+    labs(title = "Tumor Size",
+         x = "",
+         y = "") +  # Remove y-axis label 
+    theme_minimal() +
+    theme(
+      text = element_text(family = "patrick"), # Set font to Patrick Hand
+      axis.title.y = element_blank() # Remove y-axis title
+    )
 )
 
+# Combine plots using patchwork and largen the dimensions
+plots_combined <- wrap_plots(plots, ncol = 2, height = 12, width = 14)
+
+# Print combined plots
+plots_combined
+
 ```
-
-Talk more about it.
 

### 2024-04-16 paper/paper.qmd
Updating data section text and wrote about figure 1. Need to do figure 2 next and citations :)

@@
 ### Data Cleaning
-After obtaining the dataset, there were a series of rigorous preprocessing steps undertaken to ensure data quality and consistency.
-This included selecting the columns of interest namely; days to death post diagnosis, presence of distant metastasis, lymph node involvement, pathogenic stage, tumor size and treatment type.
-Next, we cleaned the column names and excluded missing values.
+After obtaining the dataset, we selected the columns of interest namely; days to death post diagnosis, presence of distant metastasis, lymph node involvement, pathogenic stage, tumor size and treatment type.
+Next, we renamed the columns, giving them meaningful names and excluded the data containing missing values.
+This left us with a sample of 382 lung cancer patients.
 Additionally, we converted the days to death column to numeric.Tests where included to ensure accuracy, reliability and validity of the datset for subsequent analysis and interpretation.
@@
 ### Data Analysis
-Descriptive and inferential statistical analyses were conducted to explore the dataset and derive meaningful insights. These included linear regression modeling.
+Descriptive and inferential statistical analyses were conducted to explore the dataset and derive meaningful insights. These included linear regression modeling, see @sec-model.
 
@@
   labs(title = "Distribution of Survival Time in Lung Cancer Patients Post-Diagnosis",
-       x = "Survival Time (days)", y = "Frequency") +
+       x = "Days to Death", y = "Frequency") +
+  scale_x_continuous(breaks = seq(0, max(analysis_data$days_to_death),
+                                  by = 500)) +  # Set breaks every 500 days
   theme_minimal() +
@@
 
-One of the critical features analyzed in this study is the duration between the date of lung cancer diagnosis and the date of death, referred to as "days to death." This metric serves as a key indicator of patient survival and provides valuable insights into the disease trajectory and prognosis.
-By examining the distribution of survival times among lung cancer patients post-diagnosis, we aim to characterize the temporal patterns of disease progression and assess the impact of various clinical and demographic factors on survival outcomes.
+The main feature analyzed in this study is the duration between the date of lung cancer diagnosis and the date of death, referred to as "days to death." This metric serves as a key indicator of patient survival and provides valuable insights into the disease trajectory and prognosis.
+By examining the distribution of survival times among lung cancer patients post-diagnosis, we aim to characterize the temporal patterns of disease progression and assess the impact of various clinical factors on survival outcomes.
 Understanding the time course from diagnosis to death is crucial for guiding treatment decisions, predicting patient outcomes, and identifying opportunities for intervention to improve survival rates.
 Through comprehensive analysis of days to death data, we seek to elucidate the factors influencing patient survival in lung cancer and contribute to the refinement of prognostic models for clinical practice.
+
+@fig-one illustrates the survival curve for lung cancer patients post-diagnosis where frequency is on the y-axis and survival time in days is on the x-axis.
+There is a clear trend of a decrease in frequency of survival as time increases.
+Mortality is most abundant 500 days after diagnosis, that is equivalent to about a year and four months.
+After about 1500 days, just over four years, the trend line plateus.
+
+@fig-one, the distribution of survival time post-diagnosis suggests that there is a high mortality rate shortly after diagnosis, thus most lung cancer patients die within one year of diagnosis (500 days).
+There is a steady decline in death and hence incline in survial up until four years (1500 days).
+The plateu after this indicates that there is a subgroup of patients who survive beyond four years.
+@sec-results explores which patients had longer survival.
 

### 2024-04-16 paper/paper.qmd
Updated data section - writing about figures 2,3,4,5. Need to fix cross-referencing

@@
 date-format: long 
-abstract: "This study investigates the impact of pathogenic stage and treatment modalities on lung cancer survival post-diagnosis.
-Analysis of patient data reveals significant correlation between pathogenic stage, seeking treatment and survival outcomes.
+abstract: "This study investigates the impact of pathologic stage and treatment modalities on lung cancer survival post-diagnosis.
+Analysis of patient data reveals significant correlation between pathologic stage, seeking treatment and survival outcomes.
 Notably, patients at advanced stages with metastases in distant sites beyond the lung, extensive lymph node involvement and tumors with extensive growth, invading nearby structures demonstrate lower survival rates.
@@
 Common treatments include surgery, radiation therapy and chemotherapy (Kai, 2021).
-Pathogenic stage is determined by presence of nearby metastasis, lymph node involvement as well as tumor spread and size (Markman, 2023).
-This paper investigates the relationship between lung cancer patients' survival and pathogenic stage.
+pathologic stage is determined by presence of nearby metastasis, lymph node involvement as well as tumor spread and size (Markman, 2023).
+This paper investigates the relationship between lung cancer patients' survival and pathologic stage.
 The estimand is the median survival time in days post-diagnosis.
@@
 
-After tumor size, LN involvement and presence of distant metastasis are categorized, the pathogenic stage of the cancer is then determined (Eldridge, 2022).
-The most valuable prognostic factor in non-small cell lung cancer is the pathogenic stage (Park, 2017) .Stage is determined by tumor size, number of tumors  and where the cancer has spread.
+After tumor size, LN involvement and presence of distant metastasis are categorized, the pathologic stage of the cancer is then determined (Eldridge, 2022).
+The most valuable prognostic factor in non-small cell lung cancer is the pathologic stage (Park, 2017) .Stage is determined by tumor size, number of tumors  and where the cancer has spread.
 Stage 1 is localized spread, stage 2 and 3 is regional spread while stage 4 is distant spread of the tumor (Eldridge, 2022).
@@
 A study done in Australia including 2119 lung cancer patients illustrated those with stage IV disease, the most advanced stage,  showed shorter survival than those at lower stages (Denton, 2016).
-The earlier the cancer is found, that is the lower the pathogenic stage, the greater the likelihood curative radiation therapy is an effective treatment (Eldridge, 2022).
-However, there is minimal literature looking at post-diagnosis survival rates based on pathogenic stage and method of treatment.
+The earlier the cancer is found, that is the lower the pathologic stage, the greater the likelihood curative radiation therapy is an effective treatment (Eldridge, 2022).
+However, there is minimal literature looking at post-diagnosis survival rates based on pathologic stage and method of treatment.
 The extent of this disease illustrates the importance of living a healthy lifestyle, undergoing regular screening and development of improved treatment methods.
@@
 
+Radiation therapy is a local treatment, targeting the tumor directly, damaging the DNA within cancer cells with the aim of shrinking the tumor.
+The success rate of radiation therapy treatment is dependent on the location, stage of the cancer as well as individual factors.
+It is the primary treatment for early stage NSCLC and palliative treatment.
+Chemotherapy is a systemic therapy utilizing drugs to kill and inhibit cancer cell growth and is often the primary treatment for stage 3 and 4 cancers.
+NSCLC patients with signs of lymph node metastases have shown great benefit in survival when treated with chemotherapy (Kai, 2021).
+A study looking at patients with pathologic stage 3 and 4 NSCLC showed at those who received chemotherapy survived for an average of 10.5 months whereas those that received RT only survived for 3.7 months (Eldridge, 2022).
+Patients with early pathologic stages such as 1, 2 and 3 undergo curative treatment.
+Stage 4, the advanced stage however, is treated as palliative treatment in that the cancer is far too advanced and thus all that can be done is to minimize symptoms and try to improve quality of life.
+Stage 3b and stage 4 tumors inoperable (Eldridge, 2022).
+Overall, the choice between treatments is made based on assessment of the patient's condition, pathologic stage with tradeoff between providing effective treatment and optimizing quality of life.
+(Eldridge, 2022)
+
 The remainder of this paper is structured as follows.
-In @sec-data, we visualize the exploration of variables constituting the pathogenic stage and treatment types.
+In @sec-data, we visualize the exploration of variables constituting the pathologic stage and treatment types.
 @sec-model, outlines the model employed to analyze the relationship between these variables and the duration of survival post-diagnosis.
@@
 ### Data Cleaning
-After obtaining the dataset, we selected the columns of interest namely; days to death post diagnosis, presence of distant metastasis, lymph node involvement, pathogenic stage, tumor size and treatment type.
+After obtaining the dataset, we selected the columns of interest namely; days to death post diagnosis, presence of distant metastasis, lymph node involvement, pathologic stage, tumor size and treatment type.
 Next, we renamed the columns, giving them meaningful names and excluded the data containing missing values.
@@
 ## Features
-The dataset comprised several key features relevant to lung cancer prognosis, including pathogenic stage, presence of distant metastasis, lymph node involvement, tumor size, and treatment type.
+The dataset comprised several key features relevant to lung cancer prognosis, including pathologic stage, presence of distant metastasis, lymph node involvement, tumor size, and treatment type.
 
@@
 
-### Pathogenic Stage
-Pathogenic stage, a critical determinant of lung cancer prognosis, was classified according to the [mention staging system used, e.g., TNM classification].
-As depicted in Figure 1, the distribution of patients varied across different stages, with [mention any notable trends, e.g., a higher proportion of patients diagnosed at advanced stages].
+### pathologic Stage
+Pathologic stage, a critical determinant of lung cancer prognosis, was classified according to the TNM (tumor, lymph node, metastasis) Cancer Staging.
+As depicted in @fig-two, illustrating the the distribution of pathologic stages with pathologic stage on the x-axis and percentage on the y-axis.
+The distribution of patients varied across different stages, with the highest proportion of patients diagnosed at StageIB.
+At this stage, the cancer is still found early however the tumor has regional spread to nearby lymph nodes.
+Interestingly, at diagnosis, patients had a fairly even spread across the different stages.
 
 ### Presence of Distant Metastasis
-The presence of distant metastasis, indicative of disease spread beyond the primary tumor site, significantly influences treatment decisions and patient outcomes.
-Figure 2 illustrates the percentage of patients with and without distant metastasis, highlighting the impact of metastatic disease on prognosis.
+The presence of distant metastasis, indicative of cancer spread beyond the site of origin to nearby lymph nodes, significantly influences treatment decisions and patient outcomes.
+@fig-3 illustrates the percentage of patients with distant metastasis where the x-axis is the classification and the y-axis is the percentage of patients.
+The figure indicates that when patients are diagnosed, most are found in category 'M0' having no distant metastasis and thus cancer has not spread beyond the site of origin.
+It is important to note that category 'MX', where tumor could not be evaluated has the next largest proprotion of patients.
+Additionally, category 'M1a' appears to show no data however this is because only 1 out of the 382 patients were classified this way.
 
 ### Lymph Node Involvement
-Lymph node involvement is a key prognostic factor in lung cancer, reflecting the extent of disease spread to regional lymph nodes.
-Figure 3 presents the distribution of patients based on lymph node involvement, demonstrating its association with survival outcomes.
+Lymph node involvement is another key prognostic factor in lung cancer, reflecting the extent of disease spread to regional lymph nodes.
+@fig-four presents the distribution of patients based on lymph node involvement where lymph node involvement is on the x-axis and percentage is on the y-axis.
+Here, there is a clear trend that as lymph node involvement becomes more severe, the proportion of patients decline.
+This demonstrates that upon diagnosis, most patients fall under the category 'N0' where there is no regional lymph node metastasis hence cancer has not spread to the lymph nodes.
 
 ### Tumor Size
-Tumor size, often measured as the diameter of the primary tumor, is closely linked to disease progression and treatment response in lung cancer patients.
-Figure 4 showcases the distribution of patients across different tumor size categories, elucidating its significance as a prognostic factor.
+Tumor size is measured based on the diameter of the primary tumor in lung cancer patients.
+@fig-five showcases the distribution of patients across different tumor size categories with tumor size on the x-axis and percentage of patients on the y-axis.
+The highest proportion of patients were categorized as 'T2' where the tumor is between 3 to 5 centimeters in diameter has grown into the inner lining of the lung, possibly leads to swelling and or collapse of the lung.
+This category was overwhelmingly more prominent than all others by twofold.
+All other categories were relatively evenly distributed.
 
@@
 #| message: false
-#| label: Figure 2
+#| label: fig-two
 #| fig-cap: ADD FIGURE 2 CAPTION HERE 
@@
     geom_bar(stat = "identity", fill = "lightpink") +
-    labs(title = "Pathologic Stage",
+    labs(title = "Figure 2: Pathologic Stage",
          x = "",
@@
     geom_bar(stat = "identity", fill = "gold") +
-    labs(title = "Presence of Distant Metastasis",
+    labs(title = "Figure 3: Presence of Distant Metastasis",
          x = "",
@@
     geom_bar(stat = "identity", fill = "cyan") +
-    labs(title = "Lymph Node Involvement",
+    labs(title = "Figure 4: Lymph Node Involvement",
          x = "",
@@
     geom_bar(stat = "identity", fill = "purple") +
-    labs(title = "Tumor Size",
+    labs(title = "Figure 5: Tumor Size",
          x = "",
@@
 In this section, we aim to predict the survival outcomes of lung cancer patients post-diagnosis with a linear regression model framework.
-We consider several predictors including pathogenic stage, lymph node involvement, presence of distant metastasis, tumor size, and treatment type.We specify the model and subsequently justify its appropriateness for our analysis.
+We consider several predictors including pathologic stage, lymph node involvement, presence of distant metastasis, tumor size, and treatment type.We specify the model and subsequently justify its appropriateness for our analysis.
 

### 2024-04-18 paper/paper.qmd
updated cross referencing and did all the citations in the references section. Still working on fixing the figure 2,3,4,5 cross referencing

@@
 
-```{r}
-#| include: false
-#| warning: false
-#| message: false
-
-# install packages
-# install.packages("knitr")
-# install.packages("kableExtra")
-
-# load packages
-# library(tidyverse)
-# library(knitr)
-# library(kableExtra)
-
-```
-
 # Introduction
@@
 We explore the hidden keys to defying the odds and emerging victorious against one of the deadliest adversaries of our time.
-Lung cancer is the leading cause of cancer-related deaths in the world (Park, 2017).
+Lung cancer is the leading cause of cancer-related deaths in the world [@Park].
 It is a disease that develops in the lining of the airways in lung tissues.
-Non-small cell lung cancer (NSCLC) is the most common type, accounting for  80-85% of all lung cancers according to the American Cancer society (Markman, 2023).
+Non-small cell lung cancer (NSCLC) is the most common type, accounting for  80-85% of all lung cancers according to the American Cancer society [@Markman].
 Staging is important for prognosis and making treatment decisions.
-Common treatments include surgery, radiation therapy and chemotherapy (Kai, 2021).
-pathologic stage is determined by presence of nearby metastasis, lymph node involvement as well as tumor spread and size (Markman, 2023).
+Common treatments include surgery, radiation therapy and chemotherapy [@Kai].
+pathologic stage is determined by presence of nearby metastasis, lymph node involvement as well as tumor spread and size [@Markman].
 This paper investigates the relationship between lung cancer patients' survival and pathologic stage.
@@
 Tumor size is often the main determinant of stage and treatment.
-As tumor categories increase, the tumor expands, invading nearby structures (Zhang, 2015).
+As tumor categories increase, the tumor expands, invading nearby structures [@Zhang].
 A study involving 52,287 patients diagnosed between the years 1998 and 2003 found tumor size to be an independent prognostic factor in estimating overall survival.
 The authors found that patients presenting with larger tumors predicted a worse prognosis and thus are associated with a decrease in survival.
-There is a similar relationship between extensive lymph node involvement and patient survival (Zhang, 2015).
-Initial spread of cancer cells are localized, then become regional, involving nearby lymph nodes and the most severe cases comprises expansion to other organs such as the brain, liver and bones (Markman, 2023).
+There is a similar relationship between extensive lymph node involvement and patient survival [@Zhang].
+Initial spread of cancer cells are localized, then become regional, involving nearby lymph nodes and the most severe cases comprises expansion to other organs such as the brain, liver and bones [@Markman].
 A study looked at five year survival rates based on the severity of spread.
 62.8% of patients with localized spread, 34.8% of patients with regional and 8% of patients with distant, advanced spread were found to survive for 5 years post diagnosis.
-More than half of these lung cancer patients have advanced spread to other organs when diagnosed(Markman, 2023).
-Overall, it is found that patients with no regional lymph node metastases, and smaller tumors are easier to be treated and thus are associated with improved survival rates (Zhang, 2015).
+More than half of these lung cancer patients have advanced spread to other organs when diagnosed [@Markman].
+Overall, it is found that patients with no regional lymph node metastases, and smaller tumors are easier to be treated and thus are associated with improved survival rates [@Zhang].
 
-Presence of metastatic LN is one of the most important determinants of prognosis of NSCLC cases (Kai, 2021).
+Presence of metastatic LN is one of the most important determinants of prognosis of NSCLC cases [@Kai].
 In the early stage, cancer has not spread to lymph nodes.
-As severity increases, lymph node metastasis sequentially spreads to more distant lymph nodes such as mediastinal and there is severe lymph node involvement  (Park, 2017).
-Lymph node involvement, also termed lymph node ratio, is a crucial factor in guiding treatment options (Kai, 2021).
-A study made up of 97 patients with a mean age of 63 who have undergone surgery between the years 2009 and 2015 in Korea find that increased lymph node involvement is associated with a more advanced disease status and hence affiliated with prognosis (Park, 2017).
+As severity increases, lymph node metastasis sequentially spreads to more distant lymph nodes such as mediastinal and there is severe lymph node involvement [@Park].
+Lymph node involvement, also termed lymph node ratio, is a crucial factor in guiding treatment options [@Kai].
+A study made up of 97 patients with a mean age of 63 who have undergone surgery between the years 2009 and 2015 in Korea find that increased lymph node involvement is associated with a more advanced disease status and hence affiliated with prognosis [@Park].
 Another study looked at 11,341 NSCLC patients between the years 2004 to 2015, from 18 geographically diverse populations, covering approximately 28% of the population of the United States.
 These patients were treatment naive and underwent surgical resection of the tumor.
-Although 5757 patients died, the rest showed great results, with a median survival of 22 months (Kai, 2021).
+Although 5757 patients died, the rest showed great results, with a median survival of 22 months [@Kai].
 The authors found that patients with low lymph node involvement lead to higher survival compared to patients with high lymph node ratios.
 A regression analysis revealed that lymph node ratio is an independent and significant predictor of patient survival.
-The authors also observed that disease burden and anatomical location of the lymph nodes involved may influence the patients survival (Kai, 2021).
+The authors also observed that disease burden and anatomical location of the lymph nodes involved may influence the patients survival [@Kai].
 
-After tumor size, LN involvement and presence of distant metastasis are categorized, the pathologic stage of the cancer is then determined (Eldridge, 2022).
-The most valuable prognostic factor in non-small cell lung cancer is the pathologic stage (Park, 2017) .Stage is determined by tumor size, number of tumors  and where the cancer has spread.
-Stage 1 is localized spread, stage 2 and 3 is regional spread while stage 4 is distant spread of the tumor (Eldridge, 2022).
-Cancer stage was determined using the seventh American Joint Committee on Cancer staging system (AJCC) (Park, 2017).
-A study done in Australia including 2119 lung cancer patients illustrated those with stage IV disease, the most advanced stage,  showed shorter survival than those at lower stages (Denton, 2016).
-The earlier the cancer is found, that is the lower the pathologic stage, the greater the likelihood curative radiation therapy is an effective treatment (Eldridge, 2022).
+After tumor size, LN involvement and presence of distant metastasis are categorized, the pathologic stage of the cancer is then determined [@Eldridge].
+The most valuable prognostic factor in non-small cell lung cancer is the pathologic stage [@Park] .Stage is determined by tumor size, number of tumors  and where the cancer has spread.
+Stage 1 is localized spread, stage 2 and 3 is regional spread while stage 4 is distant spread of the tumor [@Eldridge].
+Cancer stage was determined using the seventh American Joint Committee on Cancer staging system (AJCC) [@Park].
+A study done in Australia including 2119 lung cancer patients illustrated those with stage IV disease, the most advanced stage,  showed shorter survival than those at lower stages [@Denton].
+The earlier the cancer is found, that is the lower the pathologic stage, the greater the likelihood curative radiation therapy is an effective treatment [@Eldridge].
 However, there is minimal literature looking at post-diagnosis survival rates based on pathologic stage and method of treatment.
 The extent of this disease illustrates the importance of living a healthy lifestyle, undergoing regular screening and development of improved treatment methods.
-Over the past decade, there has been great improvement of lymph node assessment in cancer patients (Kai, 2021).
-Experts hope survival rates continue to improve with  new therapies and treatment approaches  (Markman, 2023).
+Over the past decade, there has been great improvement of lymph node assessment in cancer patients [@Kai].
+Experts hope survival rates continue to improve with  new therapies and treatment approaches [@Markman].
 
@@
 Chemotherapy is a systemic therapy utilizing drugs to kill and inhibit cancer cell growth and is often the primary treatment for stage 3 and 4 cancers.
-NSCLC patients with signs of lymph node metastases have shown great benefit in survival when treated with chemotherapy (Kai, 2021).
-A study looking at patients with pathologic stage 3 and 4 NSCLC showed at those who received chemotherapy survived for an average of 10.5 months whereas those that received RT only survived for 3.7 months (Eldridge, 2022).
+NSCLC patients with signs of lymph node metastases have shown great benefit in survival when treated with chemotherapy [@Kai].
+A study looking at patients with pathologic stage 3 and 4 NSCLC showed at those who received chemotherapy survived for an average of 10.5 months whereas those that received RT only survived for 3.7 months [@Eldridge].
 Patients with early pathologic stages such as 1, 2 and 3 undergo curative treatment.
 Stage 4, the advanced stage however, is treated as palliative treatment in that the cancer is far too advanced and thus all that can be done is to minimize symptoms and try to improve quality of life.
-Stage 3b and stage 4 tumors inoperable (Eldridge, 2022).
+Stage 3b and stage 4 tumors inoperable.
 Overall, the choice between treatments is made based on assessment of the patient's condition, pathologic stage with tradeoff between providing effective treatment and optimizing quality of life.
-(Eldridge, 2022)
+[@Eldridge].
 
@@
 This project was created using statistical software, `R` [@citeR].
-For data cleaning and manipulation, we used `tidyverse` [@tidy] package.
-For data visualization such as creating the figures, we used `ggplot2` [@ggplot2].
-For converting into Parquet file, we used `arrow` [@arrow], managing file paths `here` [@here], statistical modeling `rstanarm` [@rstanarm].
-Furthermore, `showtext` [@showtext] was used for aesthetic purposes.
+For data cleaning and manipulation, we used the `tidyverse` [@tidyverse] package, which includes `dplyr` [@dplyr], `readr` [@readr].
+Specifically, `readr` was employed for efficient reading of rectangular text data, while `dplyr` facilitated data manipulation tasks such as filtering, summarizing, and joining datasets.
+For unit testing dataset, we utilized the `testthat` [@testthat] package. This allowed us to systematically test our functions and ensure that they behaved as expected across various scenarios. 
+In our statistical modeling process, we utilized the `rstanarm` [@rstanarm] package for Bayesian applied regression modeling.
+This package leverages the Stan probabilistic programming language for efficient computation of Bayesian models, allowing us to perform complex regression analyses while incorporating uncertainty.
+To tidy up and summarize mixed effects models, we employed the `broom.mixed` [@broom.mixed] package.
+This package provides functions to visualize the results of mixed effects models, making it easier to interpret the findings from such analyses.
+For arranging and combining plots in our visualizations, we utilized the `patchwork` [@patchwork] package.
+This allowed us to seamlessly arrange multiple plots into a single coherent visual representation, facilitating the communication of complex relationships and patterns in the data.
+For aesthetic purposes in our visualizations, we employed the `showtext` [@showtext] package.
+This enabled us to use a wide range of fonts in our plots, enhancing the visual appeal and customization of our graphical outputs.
 
@@
 
-The data for this study were collected from a comprehensive database comprising 981 lung cancer patients diagnosed between 1991 to 2013 from `Center for Open Science` [@cfos], a dataset acquired in Sydney Australia.
+The data for this study were collected from a comprehensive database comprising 981 lung cancer patients diagnosed between 1991 to 2013 from `Center for Open Science` [@osf], a dataset acquired in Sydney Australia.
 The dataset included information on patient demographics, clinical characteristics, treatment modalities, and survival outcomes.
@@
 # install.packages("showtext")
+# install.packages("arrow")
 
@@
 The presence of distant metastasis, indicative of cancer spread beyond the site of origin to nearby lymph nodes, significantly influences treatment decisions and patient outcomes.
-@fig-3 illustrates the percentage of patients with distant metastasis where the x-axis is the classification and the y-axis is the percentage of patients.
+@fig-three illustrates the percentage of patients with distant metastasis where the x-axis is the classification and the y-axis is the percentage of patients.
 The figure indicates that when patients are diagnosed, most are found in category 'M0' having no distant metastasis and thus cancer has not spread beyond the site of origin.
@@
 #| message: false
-#| label: fig-two
-#| fig-cap: ADD FIGURE 2 CAPTION HERE 
-#| fig-title: "Percentage of Lung Cancer Patients by:"
+#| label: fig-two, fig-three, fig-four, fig-five
+#| fig-cap: |
+#|   fig-two: "Percentage of Lung Cancer Patients by Pathologic Stage"
+#|   fig-three: "Percentage of Lung Cancer Patients by Presence of Distant Metastasis"
+#|   fig-four: "Percentage of Lung Cancer Patients by Lymph Node Involvement"
+#|   fig-five: "Percentage of Lung Cancer Patients by Tumor Size"
 

### 2024-04-18 paper/paper.qmd
Fixed cross referencing ofr the second figure :)

@@
 Pathologic stage, a critical determinant of lung cancer prognosis, was classified according to the TNM (tumor, lymph node, metastasis) Cancer Staging.
-As depicted in @fig-two, illustrating the the distribution of pathologic stages with pathologic stage on the x-axis and percentage on the y-axis.
+As depicted in Graph 1 of @fig-two, illustrating the the distribution of pathologic stages with pathologic stage on the x-axis and percentage on the y-axis.
 The distribution of patients varied across different stages, with the highest proportion of patients diagnosed at StageIB.
@@
 The presence of distant metastasis, indicative of cancer spread beyond the site of origin to nearby lymph nodes, significantly influences treatment decisions and patient outcomes.
-@fig-three illustrates the percentage of patients with distant metastasis where the x-axis is the classification and the y-axis is the percentage of patients.
+Graph 2 of @fig-two illustrates the percentage of patients with distant metastasis where the x-axis is the classification and the y-axis is the percentage of patients.
 The figure indicates that when patients are diagnosed, most are found in category 'M0' having no distant metastasis and thus cancer has not spread beyond the site of origin.
-It is important to note that category 'MX', where tumor could not be evaluated has the next largest proprotion of patients.
+It is important to note that category 'MX', where tumor could not be evaluated has the next largest proportion of patients.
 Additionally, category 'M1a' appears to show no data however this is because only 1 out of the 382 patients were classified this way.
@@
 Lymph node involvement is another key prognostic factor in lung cancer, reflecting the extent of disease spread to regional lymph nodes.
-@fig-four presents the distribution of patients based on lymph node involvement where lymph node involvement is on the x-axis and percentage is on the y-axis.
+Graph 3 of @fig-two presents the distribution of patients based on lymph node involvement where lymph node involvement is on the x-axis and percentage is on the y-axis.
 Here, there is a clear trend that as lymph node involvement becomes more severe, the proportion of patients decline.
@@
 Tumor size is measured based on the diameter of the primary tumor in lung cancer patients.
-@fig-five showcases the distribution of patients across different tumor size categories with tumor size on the x-axis and percentage of patients on the y-axis.
+Graph 4 of @fig-two showcases the distribution of patients across different tumor size categories with tumor size on the x-axis and percentage of patients on the y-axis.
 The highest proportion of patients were categorized as 'T2' where the tumor is between 3 to 5 centimeters in diameter has grown into the inner lining of the lung, possibly leads to swelling and or collapse of the lung.
@@
 #| message: false
-#| label: fig-two, fig-three, fig-four, fig-five
-#| fig-cap: |
-#|   fig-two: "Percentage of Lung Cancer Patients by Pathologic Stage"
-#|   fig-three: "Percentage of Lung Cancer Patients by Presence of Distant Metastasis"
-#|   fig-four: "Percentage of Lung Cancer Patients by Lymph Node Involvement"
-#|   fig-five: "Percentage of Lung Cancer Patients by Tumor Size"
+#| label: fig-two
+#| fig-title: Percentage of Lung Cancer Patients by
+#| fig-cap: Percentage of Lung Cancer Patients by Pathologic Stage, Presence of Distant Metastasis, Lymph Node Involvement, Tumor Size
 
@@
     geom_bar(stat = "identity", fill = "lightpink") +
-    labs(title = "Figure 2: Pathologic Stage",
+    labs(title = "Graph 1: Pathologic Stage",
          x = "",
@@
     geom_bar(stat = "identity", fill = "gold") +
-    labs(title = "Figure 3: Presence of Distant Metastasis",
+    labs(title = "Graph 2: Presence of Distant Metastasis",
          x = "",
@@
     geom_bar(stat = "identity", fill = "cyan") +
-    labs(title = "Figure 4: Lymph Node Involvement",
+    labs(title = "Graph 3: Lymph Node Involvement",
          x = "",
@@
     geom_bar(stat = "identity", fill = "purple") +
-    labs(title = "Figure 5: Tumor Size",
+    labs(title = "Graph 4: Tumor Size",
          x = "",

### 2024-04-18 paper/paper.qmd
playing around with some other figures in the results section

@@
 
+# install.packages
+# install.packages("reshape2")
+# install.packages("vcd")
+
+# Load required libraries
+library(ggplot2)
+library(arrow)
+library(dplyr)
+library(reshape2)
+library(vcd)
+
+### read in data ###
+analysis_data <- read_parquet(file = here::here("data/analysis_data/lung_cancer_analysis_data.parquet"))
+
+# Create a contingency table
+contingency_table <- table(analysis_data$pathologic_stage, 
+                           analysis_data$presence_of_distant_metastasis, 
+                           analysis_data$lymph_node_involvement, 
+                           analysis_data$tumor_size, 
+                           analysis_data$treatment_type)
+
+# Plot mosaic plot for pathologic_stage and treatment_type with y-axis labels on the right side
+mosaic(~pathologic_stage + treatment_type, data = analysis_data, shade = TRUE, labeling_args = list(rot_labels = c(90, 0)), gp = gpar(mar = c(5, 5, 5, 5)), split_vertical = TRUE)
+
+```
+
+```{r}
+#| echo: false
+#| eval: true
+#| warning: false
+#| message: false
+
 # Load required libraries

### 2024-04-20 paper/paper.qmd
playing around with some new figures for results section

--- 2024-04-18 paper/paper.qmd
@@
       axis.title.y = element_blank() # Remove y-axis title
-    )
+    ) + guides(color = guide_legend(title.position = "top", title.vjust = 0.5))  
+    # Adjust legend title position
+
 )
@@
 Our results are summarized in @.
+
+```{r}
+##| echo: false
+#| eval: true
+#| warning: false
+#| message: false
+
+# install.packages
+# install.packages("reshape2")
+# install.packages("vcd")
+
+# Load required libraries
+library(ggplot2)
+library(arrow)
+library(dplyr)
+library(showtext)
+library(reshape2)
+library(vcd)
+library(patchwork)
+
+### read in data ###
+analysis_data <- read_parquet(file = here::here("data/analysis_data/lung_cancer_analysis_data.parquet"))
+
+# Add the desired font (e.g., Patrick Hand)
+font_add_google("Patrick Hand", "patrick")
+
+# Set the font for ggplot
+showtext_auto()
+
+# Create a scatter plot with trend lines - pathologic stage vs. time vs. treatment type
+ggplot(data = analysis_data, aes(x = days_to_death, y = pathologic_stage, color = treatment_type)) +
+  geom_point(size = 2.5, alpha = 0.6, shape = 19) +  
+  # Adjusted point size and transparency
+  geom_smooth(method = "lm", se = FALSE, aes(group = treatment_type), alpha = 0.3) +  
+  # Adjust trend line transparency
+  scale_color_manual(values = c("Pharmaceutical Therapy, NOS" = "blue", "Radiation Therapy, NOS" = "green")) +
+  labs(x = "Days to Death", y = "Pathologic Stage", color = "Treatment Type") +
+  scale_x_continuous(breaks = seq(0, max(analysis_data$days_to_death),
+                                  by = 500))+  # Set breaks on x-axis
+  theme_minimal() +
+  theme(text = element_text(family = "patrick", size = 14), 
+        # Set font to Patrick Hand and increase the size
+      legend.position = "bottom", legend.text = element_text(size = 12)) +  
+  # Adjust legend position and text size
+  guides(color = guide_legend(title.position = "top", title.vjust = 0.5))  
+  # Adjust legend title position
+
+# Create a scatter plot with trend lines - presence of distant metastasis vs. time vs. treatment type
+ggplot(data = analysis_data, aes(x = days_to_death, y = presence_of_distant_metastasis, color = treatment_type)) +
+  geom_point(size = 2.5, alpha = 0.6, shape = 19) +  
+  # Adjusted point size and transparency
+  geom_smooth(method = "lm", se = FALSE, aes(group = treatment_type), alpha = 0.3) +  
+  # Adjust trend line transparency
+  scale_color_manual(values = c("Pharmaceutical Therapy, NOS" = "blue", "Radiation Therapy, NOS" = "green")) +
+  labs(x = "Days to Death", y = "Presence of Distant Metastasis", color = "Treatment Type") +
+  scale_x_continuous(breaks = seq(0, max(analysis_data$days_to_death),
+                                  by = 500)) +  # Set breaks on x-axis
+  theme_minimal() +
+  theme(text = element_text(family = "patrick", size = 14), 
+        # Set font to Patrick Hand
+      legend.position = "bottom", legend.text = element_text(size = 12)) +  
+  # Adjust legend position and text size
+  guides(color = guide_legend(title.position = "top", title.vjust = 0.5))  
+  # Adjust legend title position
+
+# Create a scatter plot with trend lines - lymph node involvement vs. time vs. treatment type
+ggplot(data = analysis_data, aes(x = days_to_death, y = lymph_node_involvement, color = treatment_type)) +
+  geom_point(size = 2.5, alpha = 0.6, shape = 19) +  
+  # Adjusted point size and transparency
+  geom_smooth(method = "lm", se = FALSE, aes(group = treatment_type), alpha = 0.3) +  
+  # Adjust trend line transparency
+  scale_color_manual(values = c("Pharmaceutical Therapy, NOS" = "blue", "Radiation Therapy, NOS" = "green")) +
+  labs(x = "Days to Death", y = "Lymph Node Involvement", color = "Treatment Type") +
+  scale_x_continuous(breaks = seq(0, max(analysis_data$days_to_death),
+                                  by = 500)) +  # Set breaks on x-axis
+  theme_minimal() +
+  theme(text = element_text(family = "patrick", size = 14), # Set font to Patrick Hand
+      legend.position = "bottom", legend.text = element_text(size = 12)) +  
+  # Adjust legend position and text size
+  guides(color = guide_legend(title.position = "top", title.vjust = 0.5))  
+  # Adjust legend title position
+
+# Create a scatter plot with trend lines - tumor size vs. time vs. treatment type
+ggplot(data = analysis_data, aes(x = days_to_death, y = tumor_size, color = treatment_type)) +
+  geom_point(size = 2.5, alpha = 0.6, shape = 19) +  
+  # Adjusted point size and transparency
+  geom_smooth(method = "lm", se = FALSE, aes(group = treatment_type), alpha = 0.3) +  
+  # Adjust trend line transparency
+  scale_color_manual(values = c("Pharmaceutical Therapy, NOS" = "blue", "Radiation Therapy, NOS" = "green")) +
+  labs(x = "Days to Death", y = "Tumor Size", color = "Treatment Type") +
+  scale_x_continuous(breaks = seq(0, max(analysis_data$days_to_death),
+                                  by = 500)) +  # Set breaks on x-axis
+  theme_minimal() +
+  theme(text = element_text(family = "patrick", size = 14), # Set font to Patrick Hand
+      legend.position = "bottom", legend.text = element_text(size = 12)) +  
+  # Adjust legend position and text size
+  guides(color = guide_legend(title.position = "top", title.vjust = 0.5))  
+  # Adjust legend title position
+
+```
+
+```{r}
+#| echo: false
+#| eval: true
+#| warning: false
+#| message: false
+
+# install.packages
+# install.packages("reshape2")
+# install.packages("vcd")
+
+# Load required libraries
+library(ggplot2)
+library(arrow)
+library(dplyr)
+library(reshape2)
+library(vcd)
+
+### read in data ###
+analysis_data <- read_parquet(file = here::here("data/analysis_data/lung_cancer_analysis_data.parquet"))
+
+# Create a ggplot2 plot for pathologic_stage and treatment_type with adjusted margins, legend, and horizontal y-axis labels
+ggplot(analysis_data, aes(x = pathologic_stage, fill = treatment_type)) +
+  geom_bar(position = "fill") +
+  theme(axis.text.x = element_blank(), axis.text.y = element_text(angle = 0, hjust = 1)) +
+  scale_fill_manual(name = "Treatment Type", values = c("Pharmaceutical Therapy, NOS" = "blue", "Radiation Therapy, NOS" = "red"))
+
+```
 
@@
 
-```{r}
-#| echo: false
-#| eval: true
-#| label: tbl-modelresults
-#| tbl-cap: "Explanatory models of flight time based on wing width and wing length"
-#| warning: false
-
-```
-
 # Discussion {#sec-discussion}

### 2024-04-21 paper/paper.qmd
fixed captions

--- 2024-04-18 paper/paper.qmd
@@
 Our results are summarized in @.
+
+```{r}
+#| warning: false
+#| echo: false
+#| message: false
+#| label: fig-two
+#| fig-title: Pathologic Stage
+#| fig-cap: TO DO
+
+# install.packages
+# install.packages("reshape2")
+# install.packages("vcd")
+# install.packages("cowplot")
+
+# Load required libraries
+library(ggplot2)
+library(arrow)
+library(dplyr)
+library(showtext)
+library(reshape2)
+library(vcd)
+library(patchwork)
+library(cowplot)
+
+### read in data ###
+analysis_data <- read_parquet(file = here::here("data/analysis_data/lung_cancer_analysis_data.parquet"))
+
+# Add the desired font (e.g., Patrick Hand)
+font_add_google("Patrick Hand", "patrick")
+
+# Set the font for ggplot
+showtext_auto()
+
+# Create a scatter plot with trend lines - pathologic stage vs. time vs. treatment type
+plot1 <- ggplot(data = analysis_data, aes(x = days_to_death, y = pathologic_stage, color = treatment_type)) +
+  geom_point(size = 2.5, alpha = 0.6, shape = 19) +
+  # Adjusted point size and transparency
+  geom_smooth(method = "lm", se = FALSE, aes(group = treatment_type),
+              alpha = 0.3) +  # Adjust trend line transparency
+  scale_color_manual(values = c("Pharmaceutical Therapy, NOS" = "blue",
+                                "Radiation Therapy, NOS" = "green")) +
+  labs(x = "Days to Death", y = "Pathologic Stage",
+       color = "Treatment Type") +
+  scale_x_continuous(breaks = seq(0, max(analysis_data$days_to_death),
+                                  by = 500)) + # Set breaks on x-axis
+  theme_minimal() +
+  theme(text = element_text(family = "patrick", size = 14),
+        # Set font to Patrick Hand and increase the size
+        legend.position = "bottom", legend.text = element_text(size = 12)) +
+  # Adjust legend position and text size
+  guides(color = guide_legend(title.position = "top", title.vjust = 0.5)) +
+  ggtitle("Pathologic Stage vs. Time vs. Treatment Type") +
+  labs(caption = "Relationship between pathologic stage, time to death, and treatment type")
+
+# Create a scatter plot with trend lines - presence of distant metastasis vs. time vs. treatment type
+plot2 <- ggplot(data = analysis_data, aes(x = days_to_death, y = presence_of_distant_metastasis, color = treatment_type)) +
+  geom_point(size = 2.5, alpha = 0.6, shape = 19) +  
+  geom_smooth(method = "lm", se = FALSE, aes(group = treatment_type),
+              alpha = 0.3) +  
+  scale_color_manual(values = c("Pharmaceutical Therapy, NOS" = "blue",
+                                "Radiation Therapy, NOS" = "green")) +
+  labs(x = "Days to Death", y = "Presence of Distant Metastasis",
+       color = "Treatment Type") +
+  scale_x_continuous(breaks = seq(0, max(analysis_data$days_to_death),
+                                  by = 500)) +
+  theme_minimal() +
+  theme(text = element_text(family = "patrick", size = 14), 
+        legend.position = "bottom", legend.text = element_text(size = 12),
+        plot.caption = element_text(hjust = 0.5, size = 14)) +  
+  # Adjusted size here
+  guides(color = guide_legend(title.position = "top", title.vjust = 0.5)) +
+  ggtitle("Presence of Distant Metastasis vs. Time vs. Treatment Type") +
+  labs(caption = "Relationship between distant metastasis, time to death, and treatment type")  
+# Adjusted to caption
+
+# Create a scatter plot with trend lines - lymph node involvement vs. time vs. treatment type
+plot3 <- ggplot(data = analysis_data, aes(x = days_to_death,
+                                          y = lymph_node_involvement,
+                                          color = treatment_type)) +
+  geom_point(size = 2.5, alpha = 0.6, shape = 19) +  
+  geom_smooth(method = "lm", se = FALSE, aes(group = treatment_type),
+              alpha = 0.3) +  
+  scale_color_manual(values = c("Pharmaceutical Therapy, NOS" = "blue",
+                                "Radiation Therapy, NOS" = "green")) +
+  labs(x = "Days to Death", y = "Lymph Node Involvement",
+       color = "Treatment Type") +
+  scale_x_continuous(breaks = seq(0, max(analysis_data$days_to_death),
+                                  by = 500)) +
+  theme_minimal() +
+  theme(text = element_text(family = "patrick", size = 14), 
+        legend.position = "bottom", legend.text = element_text(size = 12),
+        plot.caption = element_text(hjust = 0.5, size = 14)) +  
+  # Adjusted size here
+  guides(color = guide_legend(title.position = "top", title.vjust = 0.5)) +
+  ggtitle("Lymph Node Involvement vs. Time vs. Treatment Type") +
+  labs(caption = "Relationship between lymph node involvement, time to death, and treatment type")  
+# Adjusted to caption
+
+# Create a scatter plot with trend lines - tumor size vs. time vs. treatment type
+plot4 <- ggplot(data = analysis_data, aes(x = days_to_death,
+                                          y = tumor_size,
+                                          color = treatment_type)) +
+  geom_point(size = 2.5, alpha = 0.6, shape = 19) +  
+  geom_smooth(method = "lm", se = FALSE, aes(group = treatment_type),
+              alpha = 0.3) +  
+  scale_color_manual(values = c("Pharmaceutical Therapy, NOS" = "blue",
+                                "Radiation Therapy, NOS" = "green")) +
+  labs(x = "Days to Death", y = "Tumor SIze",
+       color = "Treatment Type") +
+  scale_x_continuous(breaks = seq(0, max(analysis_data$days_to_death),
+                                  by = 500)) +
+  theme_minimal() +
+  theme(text = element_text(family = "patrick", size = 14), 
+        legend.position = "bottom", legend.text = element_text(size = 12),
+        plot.caption = element_text(hjust = 0.5, size = 14)) +  
+  # Adjusted size here
+  guides(color = guide_legend(title.position = "top", title.vjust = 0.5)) +
+  ggtitle("Tumor Size vs. Time vs. Treatment Type") +
+  labs(caption = "Relationship between tumor size, time to death, and treatment type")  
+# Adjusted to caption
+
+# Display the plots
+plot1
+plot2
+plot3
+plot4
+```
+
+```{r}
+#| echo: false
+#| eval: true
+#| warning: false
+#| message: false
+
+# install.packages
+# install.packages("reshape2")
+# install.packages("vcd")
+
+# Load required libraries
+library(ggplot2)
+library(arrow)
+library(dplyr)
+library(reshape2)
+library(vcd)
+
+### read in data ###
+analysis_data <- read_parquet(file = here::here("data/analysis_data/lung_cancer_analysis_data.parquet"))
+
+# Create a ggplot2 plot for pathologic_stage and treatment_type with adjusted margins, legend, and horizontal y-axis labels
+ggplot(analysis_data, aes(x = pathologic_stage, fill = treatment_type)) +
+  geom_bar(position = "fill") +
+  theme(axis.text.x = element_blank(), axis.text.y = element_text(angle = 0, hjust = 1)) +
+  scale_fill_manual(name = "Treatment Type", values = c("Pharmaceutical Therapy, NOS" = "blue", "Radiation Therapy, NOS" = "red"))
+
+```
 

### 2024-04-21 paper/paper.qmd
updated results section and made all graphs consistent in terms of display.

--- 2024-04-18 paper/paper.qmd
@@
 #| label: fig-one
-#| fig-cap: Distribution of survival time in patients
+#| fig-cap: The histogram shows the distribution of survival time (days_to_death) in lung cancer patients post-diagnosis.
 
@@
   geom_histogram(binwidth = 50, fill = "forestgreen", color = "black") +
-  labs(title = "Distribution of Survival Time in Lung Cancer Patients Post-Diagnosis",
+  labs(title = "Distribution of Survival Time Post-Diagnosis",
        x = "Days to Death", y = "Frequency") +
@@
 Pathologic stage, a critical determinant of lung cancer prognosis, was classified according to the TNM (tumor, lymph node, metastasis) Cancer Staging.
-As depicted in Graph 1 of @fig-two, illustrating the the distribution of pathologic stages with pathologic stage on the x-axis and percentage on the y-axis.
+As depicted in @fig-two, illustrating the the distribution of pathologic stages with pathologic stage on the x-axis and percentage on the y-axis.
 The distribution of patients varied across different stages, with the highest proportion of patients diagnosed at StageIB.
@@
 The presence of distant metastasis, indicative of cancer spread beyond the site of origin to nearby lymph nodes, significantly influences treatment decisions and patient outcomes.
-Graph 2 of @fig-two illustrates the percentage of patients with distant metastasis where the x-axis is the classification and the y-axis is the percentage of patients.
+@fig-two illustrates the percentage of patients with distant metastasis where the x-axis is the classification and the y-axis is the percentage of patients.
 The figure indicates that when patients are diagnosed, most are found in category 'M0' having no distant metastasis and thus cancer has not spread beyond the site of origin.
-It is important to note that category 'MX', where tumor could not be evaluated has the next largest proportion of patients.
-Additionally, category 'M1a' appears to show no data however this is because only 1 out of the 382 patients were classified this way.
+We decided to include the classification 'MX', where tumor could not be evaluated as it contained the next largest proportion of patients.
+If this data was omitted, there would be a large skew in the data.
+Additionally, category 'M1a' appears to show no data however this can be explained as only 1 out of the 382 patients were classified this way.
 
@@
 Lymph node involvement is another key prognostic factor in lung cancer, reflecting the extent of disease spread to regional lymph nodes.
-Graph 3 of @fig-two presents the distribution of patients based on lymph node involvement where lymph node involvement is on the x-axis and percentage is on the y-axis.
+@fig-two presents the distribution of patients based on lymph node involvement where lymph node involvement is on the x-axis and percentage is on the y-axis.
 Here, there is a clear trend that as lymph node involvement becomes more severe, the proportion of patients decline.
@@
 Tumor size is measured based on the diameter of the primary tumor in lung cancer patients.
-Graph 4 of @fig-two showcases the distribution of patients across different tumor size categories with tumor size on the x-axis and percentage of patients on the y-axis.
+@fig-two showcases the distribution of patients across different tumor size categories with tumor size on the x-axis and percentage of patients on the y-axis.
 The highest proportion of patients were categorized as 'T2' where the tumor is between 3 to 5 centimeters in diameter has grown into the inner lining of the lung, possibly leads to swelling and or collapse of the lung.
@@
 #| label: fig-two
-#| fig-title: Percentage of Lung Cancer Patients by
-#| fig-cap: Percentage of Lung Cancer Patients by Pathologic Stage, Presence of Distant Metastasis, Lymph Node Involvement, Tumor Size
+#| fig-cap: The histogram shows the percentage of lung cancer patients by pathologic stage, presence of distant metastasis, lymph node involvement, tumor size
 
@@
 showtext_auto()
+
+# Set overall title and caption
+overall_title <- "Percentage of Lung Cancer Patients by:"
 
@@
     geom_bar(stat = "identity", fill = "lightpink") +
-    labs(title = "Graph 1: Pathologic Stage",
+    labs(title = "Pathologic Stage",
          x = "",
@@
     geom_bar(stat = "identity", fill = "gold") +
-    labs(title = "Graph 2: Presence of Distant Metastasis",
+    labs(title = "Presence of Distant Metastasis",
          x = "",
@@
     geom_bar(stat = "identity", fill = "cyan") +
-    labs(title = "Graph 3: Lymph Node Involvement",
+    labs(title = "Lymph Node Involvement",
          x = "",
@@
     geom_bar(stat = "identity", fill = "purple") +
-    labs(title = "Graph 4: Tumor Size",
+    labs(title = "Tumor Size",
          x = "",
@@
       axis.title.y = element_blank() # Remove y-axis title
+    ) + guides(color = guide_legend(title.position = "top", title.vjust = 0.5))  
+    # Adjust legend title position
+
+)
+
+# Combine plots using patchwork and set overall title and caption
+plots_combined <- wrap_plots(plots, ncol = 2, heights = c(1.2, 1.2)) +
+  plot_annotation(
+    title = overall_title,
+    theme = theme(
+      text = element_text(family = "patrick", size = 14),  # Adjust caption size
+      plot.caption = element_text(family = "patrick", size = 12, hjust = 0) 
+      # Move caption to far right
     )
@@
 
-# Combine plots using patchwork and largen the dimensions
-plots_combined <- wrap_plots(plots, ncol = 2, height = 12, width = 14)
-
 # Print combined plots
 plots_combined
+
+```
+
+### Treatment Type
+Treatment type analysis examines the distribution of lung cancer patients based on the type of treatment they received.
+@fig-three illustrates this distribution, with treatment type on the x-axis and the percentage of patients on the y-axis.
+Pharmaceutical Therapy and Radiation Therapy were the two primary treatment categories observed.
+The split between usage of these two treatment types was very minimal with pharmaceutical therapy at 50.3% and radiation therapy at 49.7%.
+
+```{r}
+#| warning: false
+#| echo: false
+#| message: false
+#| label: fig-three
+#| fig-cap: The histogram shows the percentage of lung cancer patients by treamtent type.
+
+# load packages
+library(ggplot2)
+library(dplyr)
+library(showtext)
+library(arrow)
+
+### read in data ###
+analysis_data <- read_parquet(file = here::here("data/analysis_data/lung_cancer_analysis_data.parquet"))
+
+# Add the desired font (e.g., Patrick Hand)
+font_add_google("Patrick Hand", "patrick")
+
+# Set the font for ggplot
+showtext_auto()
+
+# Create bar chart for treatment type with color coding
+treatment_type_counts <- analysis_data %>%
+  group_by(treatment_type) %>%
+  summarise(count = n()) %>%
+  mutate(percentage = count / sum(count) * 100)
+
+# Define colors for pharmaceutical therapy and radiation therapy
+color_palette <- c("Pharmaceutical Therapy, NOS" = "blue", "Radiation Therapy, NOS" = "green")
+
+# Plotting with color coding and percentages
+ggplot(treatment_type_counts, aes(x = treatment_type, y = percentage,
+                                  fill = treatment_type,
+                                  label = paste0(round(percentage, 1), "%"))) +
+  geom_bar(stat = "identity") +
+  geom_text(aes(label = paste0(round(percentage, 1), "%")), family = "patrick",
+            vjust = -0.5, size = 4) +  
+  scale_fill_manual(values = color_palette) +
+  labs(x = "Treatment Type", y = "Percentage of Patients",
+       title = "Distribution of Lung Cancer Patients by Treatment Type") +
+  theme_minimal() +
+  scale_y_continuous(breaks = seq(0, max(treatment_type_counts$percentage), by = 5)) +  
+  theme(text = element_text(family = "patrick", size = 14))
 
@@
 
-Our results are summarized in @.
+Our results are summarized in @fig-four, @fig-five, @fig-six and @fig-seven.
+This analysis investigates lung cancer treatment outcomes given either radiation therapy or pharmaceutical therapy as well as pathologic stage, presence of distant metastasis, lymph node involvement and tumor size.
+
+## Treatment Type by Pathologic Stage
+This scatter plot @fig-four provides a view of how pathologic stage correlates with the time to death among lung cancer patients, differentiated by treatment type.
+Each dot represents an individual patient, color-coded to represent whether they received pharmaceutical therapy or radiation therapy.
+On the x-axis we have days to death after diagnosis and on the y-axis we have the pathologic stages.
+There are considerably fewer patients part of stages I, II, III, and IIB at diagnosis compared to the other ones.
+As the time from diagnosis to death increases along the x-axis, we observe a general trend downwards in pathologic stage, indicating disease progression over time.
+The trend lines fitted for each treatment type further highlight this progression, with radiation therapy showing a steeper decline compared to pharmaceutical therapy.
+This illustrates that upon diagnosis, patients with more advanced pathologic stages are associated with a shorter survival time.
+Additionally, this suggests that patients receiving pharmaceutical therapy tend to exhibit a slower advancement in pathologic stage over time, thus increasing their survival time more so than those undergoing radiation therapy.
 
 ```{r}
+#| warning: false
 #| echo: false
-#| eval: true
-#| warning: false
 #| message: false
-
-# install.packages
-# install.packages("reshape2")
-# install.packages("vcd")
+#| label: fig-four
+#| fig-cap: The scatterplot shows the relationship between pathologic stage, time to death, and treatment type in lung cancer patients post-diagnosis.
 
@@
 library(arrow)
-library(dplyr)
-library(reshape2)
-library(vcd)
+library(showtext)
 
@@
 
-# Create a contingency table
-contingency_table <- table(analysis_data$pathologic_stage, 
-                           analysis_data$presence_of_distant_metastasis, 
-                           analysis_data$lymph_node_involvement, 
-                           analysis_data$tumor_size, 
-                           analysis_data$treatment_type)
+# Add the desired font (e.g., Patrick Hand)
+font_add_google("Patrick Hand", "patrick")
 
-# Plot mosaic plot for pathologic_stage and treatment_type with y-axis labels on the right side
-mosaic(~pathologic_stage + treatment_type, data = analysis_data, shade = TRUE, labeling_args = list(rot_labels = c(90, 0)), gp = gpar(mar = c(5, 5, 5, 5)), split_vertical = TRUE)
+# Set the font for ggplot
+showtext_auto()
+
+# Create a scatter plot with trend lines - pathologic stage vs. time vs. treatment type
+# Create scatter plot with trend lines - pathologic stage vs. time vs. treatment type
+plot1 <- ggplot(data = analysis_data, aes(x = days_to_death, y = pathologic_stage, color = treatment_type)) +
+  geom_point(size = 2.5, alpha = 0.6, shape = 19) +  # Adjusted point size and transparency
+  geom_smooth(method = "lm", se = FALSE, aes(group = treatment_type), alpha = 0.3) +  # Adjust trend line transparency
+  scale_color_manual(values = c("Pharmaceutical Therapy, NOS" = "blue",
+                                "Radiation Therapy, NOS" = "green")) +
+  labs(x = "Days to Death", y = "Pathologic Stage", color = "Treatment Type") +
+  scale_x_continuous(breaks = seq(0, max(analysis_data$days_to_death), by = 500)) + 
+  # Set breaks on x-axis
+  theme_minimal() +
+  theme(text = element_text(family = "patrick", size = 14),  # Set font to Patrick Hand and increase the size
+        legend.position = "bottom", legend.text = element_text(size = 12),
+        plot.caption = element_text(hjust = 0, size = 14)) +  # Adjust legend position and text size
+  guides(color = guide_legend(title.position = "top", title.vjust = 0.5)) +
+  ggtitle("Pathologic Stage vs. Time vs. Treatment Type")  # Title for the plot
+
+# Display the plot
+plot1
 
@@
 
+## Treatment Type by Presence of Distant Metastasis
+In this scatter plot, @fig-five we examine the relationship between the presence of distant metastasis, time to death, and treatment type for lung cancer patients.
+Similarly to figure above, the dots on the plot represent individual patients, color-coded based on their treatment regimen.
+Days to death after diagnosis is on the x-axis and on the y-axis we have the degree of the presence of distant metastasis.
+The majority of patients are diagnosed at classification 'MO' and hence are caught before cancer is spread from the primary site of origin.
+As patients survival time increases, excluding  category 'MX', there is a decrease in the presence of distant metastasis, denoted by the downward trend in the scatter plot.
+Patients undergoing pharmaceutical therapy show a sharper decrease in distant metastasis compared to those receiving radiation therapy.
+The trend line for radiation therapy is nearly a plateau, insinuating that radiation therapy was more effective in prolonging survival.
+This suggests that pharmaceutical therapy may  correlate with a lower likelihood of survival given distant metastasis over time compared to radiation therapy.
+Additionally, patients associated with higher classification of metastasis appear to have shorter survival times.
+
 ```{r}
+#| warning: false
 #| echo: false
-#| eval: true
-#| warning: false
 #| message: false
+#| label: fig-five
+#| fig-cap: The scatterplot illustrates the relationship between distant metastasis, time to death, and treatment type in lung cancer patients post-diagnosis.
 
@@
 library(arrow)
+library(showtext)
 
@@
 
-# Scatter plot survival time vs. pathologic stage - to compare survival probabilities
-ggplot(analysis_data, aes(x = pathologic_stage, y = days_to_death)) +
-  geom_point(size = 3, color = "lightpink") +
-  geom_smooth(method = "lm", se = FALSE, color = "red") +  
-  # Add a linear trend line
-  labs(title = "Survival Time vs. Pathologic Stage Post-Diagnosis",
-       x = "Pathologic Stage",
-       y = "Survival Time (days)") +
+# Add the desired font (e.g., Patrick Hand)
+font_add_google("Patrick Hand", "patrick")
+
+# Set the font for ggplot
+showtext_auto()
+
+# Create a scatter plot with trend lines - presence of distant metastasis vs. time vs. treatment type
+plot2 <- ggplot(data = analysis_data, aes(x = days_to_death, y = presence_of_distant_metastasis, color = treatment_type)) +
+  geom_point(size = 2.5, alpha = 0.6, shape = 19) +  
+  geom_smooth(method = "lm", se = FALSE, aes(group = treatment_type),
+              alpha = 0.3) +  
+  scale_color_manual(values = c("Pharmaceutical Therapy, NOS" = "blue",
+                                "Radiation Therapy, NOS" = "green")) +
+  labs(x = "Days to Death", y = "Presence of Distant Metastasis",
+       color = "Treatment Type") +
+  scale_x_continuous(breaks = seq(0, max(analysis_data$days_to_death),
+                                  by = 500)) +
   theme_minimal() +
-  theme(axis.text.x = element_text(angle = 45, hjust = 1)) + 
-  # Rotate x-axis labels for better readability
-  theme(text = element_text(family = "patrick") 
-  # Set font to Patrick Hand
-  )
+  theme(text = element_text(family = "patrick", size = 14), 
+        legend.position = "bottom", legend.text = element_text(size = 12),
+        plot.caption = element_text(hjust = 0, size = 14)) +  
+  guides(color = guide_legend(title.position = "top", title.vjust = 0.5)) +
+  ggtitle("Presence of Distant Metastasis vs. Time vs. Treatment Type")
 
-# Line graph with trend lines 
-ggplot(analysis_data, aes(x = pathologic_stage, y = days_to_death)) +
-  geom_line(color = "blue", group = 1) +
-  geom_smooth(method = "lm", se = FALSE, color = "red") +  
-  # Add a linear trend line
-  labs(title = "Days to Death vs. Pathologic Stage for Lung Cancer Patients",
-       x = "Pathologic Stage",
-       y = "Days to Death") +
-  theme_minimal() +
-  theme(axis.text.x = element_text(angle = 45, hjust = 1))  # Rotate x-axis labels for better readability
-
-# not sure what this is 
-# Plotting
-ggplot(analysis_data, aes(x = days_to_death, fill = pathologic_stage)) +
-  geom_density(alpha = 0.5) +
-  geom_density(color = "black", size = 1.5) +
-  labs(title = "Density Plot of Days to Death by Pathologic Stage for Lung Cancer Patients",
-       x = "Days to Death",
-       y = "Density") +
-  scale_fill_manual(values = c("lightgreen", "lightgreen", "lightgreen","forestgreen", "forestgreen", "forestgreen", "darkgreen","darkgreen", "darkgreen", "red")) +
-  theme_minimal()
-
-# not sure .2
-ggplot(analysis_data, aes(x = days_to_death, fill = pathologic_stage)) +
-  geom_density(alpha = 0.5) +
-  geom_density(color = "black", size = 1.5) +
-  labs(title = "Density Plot of Days to Death by Pathologic Stage for Lung Cancer Patients",
-       x = "Days to Death",
-       y = "Density") +
-  scale_fill_manual(values = c("lightgreen", "lightgreen", "lightgreen","forestgreen", "forestgreen", "forestgreen", "darkgreen","darkgreen", "darkgreen", "red")) +
-  theme_minimal() +
-  facet_wrap(~pathologic_stage, scales = "free")
-
-# Boxplot 
-ggplot(analysis_data, aes(x = pathologic_stage, y = days_to_death)) +
-  geom_boxplot(fill = "lightblue") +
-  labs(title = "Survial Distribution Post-Diagnosis by Pathologic Stage",
-       x = "Pathologic Stage",
-       y = "Days to Death") +
-  theme_minimal() +
-  theme(axis.text.x = element_text(angle = 45, hjust = 1))  # Rotate x-axis labels for better readability
-
-######################################
-
-# Scatter plot survival time vs. Presence of Distant Metastasis
-ggplot(analysis_data, aes(x = presence_of_distant_metastasis, y = days_to_death)) +
-  geom_point(size = 3, color = "gold") +
-  geom_smooth(method = "lm", se = FALSE, color = "red") +  
-  # Add a linear trend line
-  labs(title = "Survival Time vs. Presence of Distant Metastasis Post-Diagnosis",
-       x = "Presence of Distant Metastasis",
-       y = "Survival Time (days)") +
-  theme_minimal() +
-  theme(axis.text.x = element_text(angle = 45, hjust = 1)) + 
-  # Rotate x-axis labels for better readability
-  theme(text = element_text(family = "patrick") 
-  # Set font to Patrick Hand
-  )
-
-# Scatter plot survival time vs. Lymph node involvement
-ggplot(analysis_data, aes(x = lymph_node_involvement, y = days_to_death)) +
-  geom_point(size = 3, color = "cyan") +
-  geom_smooth(method = "lm", se = FALSE, color = "red") +  
-  # Add a linear trend line
-  labs(title = "Survival Time vs. Lymph Node Involvement Post-Diagnosis",
-       x = "Lymph Node Involvement",
-       y = "Survival Time (days)") +
-  theme_minimal() +
-  theme(axis.text.x = element_text(angle = 45, hjust = 1)) + 
-  # Rotate x-axis labels for better readability
-  theme(text = element_text(family = "patrick") 
-  # Set font to Patrick Hand
-  )
-
-# Scatter plot survival time vs. Tumor Size
-ggplot(analysis_data, aes(x = tumor_size, y = days_to_death)) +
-  geom_point(size = 3, color = "purple") +
-  geom_smooth(method = "lm", se = FALSE, color = "red") +  
-  # Add a linear trend line
-  labs(title = "Survival Time vs. Tumor Size Post-Diagnosis",
-       x = "Tumor Size",
-       y = "Survival Time (days)") +
-  theme_minimal() +
-  theme(axis.text.x = element_text(angle = 45, hjust = 1)) + 
-  # Rotate x-axis labels for better readability
-  theme(text = element_text(family = "patrick") 
-  # Set font to Patrick Hand
-  )
-
-#########################
-
-# Scatter plot survival time vs. Treatment Type
-ggplot(analysis_data, aes(x = treatment_type, y = days_to_death)) +
-  geom_point(size = 3, color = "cyan") +
-  geom_smooth(method = "lm", se = FALSE, color = "red") +  
-  # Add a linear trend line
-  labs(title = "Survival Time vs. Treatment Type Post-Diagnosis",
-       x = "Treatment Type",
-       y = "Survival Time (days)") +
-  theme_minimal() +
-  theme(axis.text.x = element_text(angle = 45, hjust = 1)) + 
-  # Rotate x-axis labels for better readability
-  theme(text = element_text(family = "patrick") 
-  # Set font to Patrick Hand
-  )
-
-# Grouped bar chart for survival time vs. Treatment Type
-ggplot(analysis_data, aes(x = treatment_type, y = days_to_death, fill = treatment_type)) +
-  geom_bar(stat = "summary", fun = "mean", position = "dodge") +
-  labs(title = "Average Survival Time by Treatment Type Post-Diagnosis",
-       x = "Treatment Type",
-       y = "Average Survival Time (days)") +
-  theme_minimal() +
-  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
-  theme(text = element_text(family = "patrick")) 
-  # Set font to Patrick Hand
-
-# Horizontal grouped bar chart for survival time vs. Treatment Type
-ggplot(analysis_data, aes(y = treatment_type, x = days_to_death, fill = treatment_type)) +
-  geom_bar(stat = "summary", fun = "mean", position = "dodge", orientation = "horizontal") +
-  labs(title = "Average Survival Time by Treatment Type Post-Diagnosis",
-       y = "Treatment Type",
-       x = "Average Survival Time (days)") +
-  theme_minimal() +
-  theme(text = element_text(family = "patrick")) 
-  # Set font to Patrick Hand
-
-# Grouped dot plot for survival time vs. Treatment Type
-ggplot(analysis_data, aes(x = treatment_type, y = days_to_death, color = treatment_type)) +
-  geom_jitter(position = position_dodge(width = 0.5), size = 3, alpha = 0.7) +
-  labs(title = "Survival Time by Treatment Type Post-Diagnosis",
-       x = "Treatment Type",
-       y = "Survival Time (days)") +
-  theme_minimal() +
-  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
-  theme(text = element_text(family = "patrick")) 
-  # Set font to Patrick Hand
-
-# Grouped dot plot with flipped axes for survival time vs. Treatment Type
-ggplot(analysis_data, aes(y = treatment_type, x = days_to_death, color = treatment_type)) +
-  geom_jitter(position = position_dodge(width = 0.5), size = 3, alpha = 0.7) +
-  labs(title = "Survival Time by Treatment Type Post-Diagnosis",
-       y = "Treatment Type",
-       x = "Survival Time (days)") +
-  theme_minimal() +
-  theme(axis.text.y = element_text(angle = 0, hjust = 1)) +
-  theme(text = element_text(family = "patrick")) 
-  # Set font to Patrick Hand
-
-# Stacked bar chart for treatment method vs. Average Survival Time
-ggplot(analysis_data, aes(x = days_to_death, fill = treatment_type)) +
-  geom_bar(position = "fill") +
-  labs(title = "Distribution of Treatment Methods by Average Survival Time Post-Diagnosis",
-       x = "Average Survival Time (days)",
-       y = "Proportion of Patients") +
-  scale_fill_manual(values = c("cyan", "magenta")) + # Adjust fill colors
-  theme_minimal() +
-  theme(text = element_text(family = "patrick")) 
-  # Set font to Patrick Hand
-
-# Stacked bar chart for treatment method vs. Average Survival Time
-ggplot(analysis_data, aes(x = days_to_death, fill = treatment_type)) +
-  geom_bar(position = "fill") +
-  labs(title = "Distribution of Treatment Methods by Average Survival Time Post-Diagnosis",
-       x = "Average Survival Time (days)",
-       y = "Proportion of Patients") +
-  scale_fill_manual(values = c("cyan", "magenta")) + # Adjust fill colors
-  scale_x_continuous(expand = c(0, 0)) + # Adjust x-axis limits
-  theme_minimal() +
-  theme(text = element_text(family = "patrick")) 
-  # Set font to Patrick Hand
-
-# Stacked bar chart for treatment method vs. Average Survival Time
-ggplot(analysis_data, aes(x = days_to_death, fill = treatment_type)) +
-  geom_bar(position = "fill") +
-  labs(title = "Distribution of Treatment Methods by Average Survival Time Post-Diagnosis",
-       x = "Average Survival Time (days)",
-       y = "Proportion of Patients") +
-  scale_fill_manual(values = c("cyan", "magenta")) + # Adjust fill colors
-  scale_x_continuous(expand = c(0, 0), limits = c(0, max(analysis_data$days_to_death))) + # Adjust x-axis limits
-  theme_minimal() +
-  theme(text = element_text(family = "patrick")) 
-  # Set font to Patrick Hand
-
-# Boxplot 
-ggplot(analysis_data, aes(x = treatment_type, y = days_to_death)) +
-  geom_boxplot(fill = "lightblue") +
-  labs(title = "Survial Distribution Post-Diagnosis by Treatment Type",
-       x = "Treatment Type",
-       y = "Days to Death") +
-  theme_minimal() +
-  theme(axis.text.x = element_text(angle = 45, hjust = 1))  # Rotate x-axis labels for better readability
-
+# Display the plot
+plot2
 ```
+## Treatment Type by Lymph Node Involvement 
+This scatter plot  @fig-six explores the relationship between lymph node involvement, time to death, and treatment type among cancer patients.
+Like above, each dot symbolizes an individual patient, with colors indicating their treatment type; pharmaceutical therapy or radiation therapy.
+Days to death is on the x-axis and lymph node involvement classifications are on the y-axis.
+At diagnosis, most patients are classified under either NO, N1 or N2.
+As survival time increases, there is an observable decline in lymph node involvement, reflected by the downward trend in the scatter plot.
+The trend lines further illuminate this pattern, with patients undergoing pharmaceutical therapy exhibiting a slightly steeper decline in lymph node involvement over time compared to those receiving radiation therapy, however this is very minimal and insignificant difference between the two treatment types.
+This suggests that pharmaceutical therapy and radiation therapy are both equally associated with increasing survival time based on lymph node involvement.
+Moreover, patients associated with higher classification of involvement with lymph nodes appear to have shorter survival times.
 
 ```{r}
+#| warning: false
 #| echo: false
-#| eval: true
-#| label: tbl-modelresults
-#| tbl-cap: "Explanatory models of flight time based on wing width and wing length"
+#| message: false
+#| label: fig-six
+#| fig-caption: The scatterplot demonstrates the relationship between lymph node involvement, time to death, and treatment type in lung cancer patients post-diagnosis.
+
+# Load required libraries
+library(ggplot2)
+library(arrow)
+library(showtext)
+
+### read in data ###
+analysis_data <- read_parquet(file = here::here("data/analysis_data/lung_cancer_analysis_data.parquet"))
+
+# Add the desired font (e.g., Patrick Hand)
+font_add_google("Patrick Hand", "patrick")
+
+# Set the font for ggplot
+showtext_auto()
+
+# Create a scatter plot with trend lines - lymph node involvement vs. time vs. treatment type
+plot3 <- ggplot(data = analysis_data, aes(x = days_to_death,
+                                          y = lymph_node_involvement,
+                                          color = treatment_type)) +
+  geom_point(size = 2.5, alpha = 0.6, shape = 19) +  
+  geom_smooth(method = "lm", se = FALSE, aes(group = treatment_type),
+              alpha = 0.3) +  
+  scale_color_manual(values = c("Pharmaceutical Therapy, NOS" = "blue",
+                                "Radiation Therapy, NOS" = "green")) +
+  labs(x = "Days to Death", y = "Lymph Node Involvement",
+       color = "Treatment Type") +
+  scale_x_continuous(breaks = seq(0, max(analysis_data$days_to_death),
+                                  by = 500)) +
+  theme_minimal() +
+  theme(text = element_text(family = "patrick", size = 14), 
+        legend.position = "bottom", legend.text = element_text(size = 12),
+        plot.caption = element_text(hjust = 0, size = 14)) +  
+  guides(color = guide_legend(title.position = "top", title.vjust = 0.5)) +
+  ggtitle("Lymph Node Involvement vs. Time vs. Treatment Type")
+
+# Display the plot
+plot3
+```
+## Treatment Type by Tumor Size
+In this scatter plot, @fig-seven we investigate the relationship between tumor size, time to death, and treatment type for cancer patients.
+Each dot on the plot corresponds to an individual patient, categorized by their treatment regimen whether it is pharmaceutical therapy or radiation therapy.
+The x-axis represents the time from diagnosis until death and the y-axis is comprised of the tumor sizes.
+At diagnosis most patients are classified with a T2 tumour.
+As survival time increases, there is a noticeable reduction in tumor size, depicted by the downwards trajectory in the scatter plot.
+The trend lines further accentuate this trend, with patients undergoing radiation therapy demonstrating a more pronounced decrease in tumor size over time compared to those receiving pharmaceutical therapy.
+This suggests that patients associated with larger tumors tend to have shorter survival times.
+This suggests that pharmaceutical therapy might be associated with a slower tumor growth rate and hence be more efficient in treating lung cancer patients in terms of tumor size.
+
+```{r}
 #| warning: false
+#| echo: false
+#| message: false
+#| label: fig-seven
+#| fig-caption: The scatterplot portrays the relationship between tumor size, time to death, and treatment type in lung cancer patients post-diagnosis.
+
+# Load required libraries
+library(ggplot2)
+library(arrow)
+library(showtext)
+
+### read in data ###
+analysis_data <- read_parquet(file = here::here("data/analysis_data/lung_cancer_analysis_data.parquet"))
+
+# Add the desired font (e.g., Patrick Hand)
+font_add_google("Patrick Hand", "patrick")
+
+# Set the font for ggplot
+showtext_auto()
+
+# Create a scatter plot with trend lines - tumor size vs. time vs. treatment type
+plot4 <- ggplot(data = analysis_data, aes(x = days_to_death,
+                                          y = tumor_size,
+                                          color = treatment_type)) +
+  geom_point(size = 2.5, alpha = 0.6, shape = 19) +  
+  geom_smooth(method = "lm", se = FALSE, aes(group = treatment_type),
+              alpha = 0.3) +  
+  scale_color_manual(values = c("Pharmaceutical Therapy, NOS" = "blue",
+                                "Radiation Therapy, NOS" = "green")) +
+  labs(x = "Days to Death", y = "Tumor Size",
+       color = "Treatment Type") +
+  scale_x_continuous(breaks = seq(0, max(analysis_data$days_to_death),
+                                  by = 500)) +
+  theme_minimal() +
+  theme(text = element_text(family = "patrick", size = 14), 
+        legend.position = "bottom", legend.text = element_text(size = 12),
+        plot.caption = element_text(hjust = 0, size = 14)) +  
+  guides(color = guide_legend(title.position = "top", title.vjust = 0.5)) +
+  ggtitle("Tumor Size vs. Time vs. Treatment Type")
+
+# Display the plot
+plot4
 

### 2024-04-22 paper/paper.qmd
updated title, updated discussion + conclusion, added appendix.

@@
 title: "Survival Compass"
-subtitle: "Statistical Insights into Lung Cancer Patients Journey Post Diagnosis"
+subtitle: "Understanding the Impact of Pathologic Stage and Treatment Modalities on Lung Cancer Patients' Survival Post-Diagnosis."
 author: 
@@
 The trend lines fitted for each treatment type further highlight this progression, with radiation therapy showing a steeper decline compared to pharmaceutical therapy.
-This illustrates that upon diagnosis, patients with more advanced pathologic stages are associated with a shorter survival time.
-Additionally, this suggests that patients receiving pharmaceutical therapy tend to exhibit a slower advancement in pathologic stage over time, thus increasing their survival time more so than those undergoing radiation therapy.
 
@@
 Patients undergoing pharmaceutical therapy show a sharper decrease in distant metastasis compared to those receiving radiation therapy.
-The trend line for radiation therapy is nearly a plateau, insinuating that radiation therapy was more effective in prolonging survival.
-This suggests that pharmaceutical therapy may  correlate with a lower likelihood of survival given distant metastasis over time compared to radiation therapy.
-Additionally, patients associated with higher classification of metastasis appear to have shorter survival times.
 
@@
 The trend lines further illuminate this pattern, with patients undergoing pharmaceutical therapy exhibiting a slightly steeper decline in lymph node involvement over time compared to those receiving radiation therapy, however this is very minimal and insignificant difference between the two treatment types.
-This suggests that pharmaceutical therapy and radiation therapy are both equally associated with increasing survival time based on lymph node involvement.
-Moreover, patients associated with higher classification of involvement with lymph nodes appear to have shorter survival times.
 
@@
 The trend lines further accentuate this trend, with patients undergoing radiation therapy demonstrating a more pronounced decrease in tumor size over time compared to those receiving pharmaceutical therapy.
-This suggests that patients associated with larger tumors tend to have shorter survival times.
-This suggests that pharmaceutical therapy might be associated with a slower tumor growth rate and hence be more efficient in treating lung cancer patients in terms of tumor size.
 
@@
 # Discussion {#sec-discussion}
+Lung cancer presents significant challenges in oncology, with survival rates influenced by various factors.
+This study aimed to elucidate the interplay between pathologic stage, treatment modalities, and survival outcomes in lung cancer patients post-diagnosis.
 
-## First discussion point {#sec-first-point}
+## Survival Dynamics {#sec-first-point}
+Eldridge explains that the severity of lung cancer is staged by pathologic stage which is determined by all three presence of distant metastasis, lymph node involvement and tumor size [@Eldrige].
+Previous research found that patients with larger tumors and more extensive lymph node involvement are associated with a decrease in survival [@Zhang].
+Additionally, greater presence of distant metastasis and thus advanced spread to other organs is shown to lower a patients survival [@Markman].
+Our analysis revealed compelling associations between various clinical factors and survival times in lung cancer patients.
+Upon diagnosis, lung cancer patients with more advanced pathologic stages (@fig-four), classification of presence of distant metastasis (@fig-five), higher greater involvement of lymph nodes (@fig-six) and larger tumors (@fig-seven) were all shown to be associated with a shorter survival time.
 
-## Second discussion point
+## Treatment Efficacy {#sec-second-point}
+An intriguing finding from our study was the differential impact of treatment modalities on survival outcomes.
+It was observed that @fig-four which looked at pathologic stage, demonstrates that patients receiving pharmaceutical therapy tend to exhibit a slower progression in pathologic stage over time, thus linked to having greater survival time compared to those undergoing radiation therapy.
+As pathologic stage is determined based on presence of distant metastasis, lymph node involvement and tumor size, this means that the main finding is that pharmaceutical therapy is the more effective treatment.
+In looking at these variables separately however, we see that not all the variables indicate that pharmaceutical therapy is superior.
 
-## Third discussion point
+Results looking at tumor size @fig-seven, pharmaceutical therapy is associated with slower tumor growth rate and thus pharmaceutical therapy is more potent at treating lung cancer patients, aligning with the results of pathologic stage @fig-four.
+On the other hand, the plateaued trend line for radiation therapy compared to the sharper decline of pharmaceutical therapy in @fig-five of presence of distant metastasis, this portrays radiation therapy was more effective in prolonging survival.
+Finally, with lymph node involvement @fig-six, pharmaceutical therapy and radiation therapy are both equally associated with increasing survival time based on lymph node involvement.
+These nuanced insights underscore the need for tailored treatment strategies based on individual patient characteristics and disease profiles.
 
-## Weaknesses and next steps
+## Limitations
+Despite the insights gained from our study, several limitations warrant acknowledgment.
+As mentioned earlier, the data set included 981 patients and after removing patients with incomplete data, we were left with 382 patients.
+This is a much smaller sample and thus there is a chance of bias and impact validity of our findings.
+Another area of weakness could be that values for 'MX' presence of distant metastasis and 'NX" lymph node involvement which indicate cases that could not be assed, these were left in the data set as to not decrease the sample size anymore.
+It likely that had these been taken out, the results would be different.
+We made the executive decision to include these values as reasons for inability to be assessed could be because the spread was too small to be detected.
+Additionally, our study did not account for certain variables such as performance status, smoking status, and genetic mutations, which may influence treatment responses and survival outcomes.
+Another limitation is that this study treatment types were pharmaceutical and radiation treatment individually.
+In previous studies, pharmaceutical treatment is often used in combination with radiation therapy to treat lung cancer as its shown better results in destroying cancer cells.
+In some patients, pharmaceutical therapy ensures the tumor remains small in size and the radiation destroys the cancer.
+
+## Future Directions
+The dataset we used also offered information regarding demographics such as race, gender and year of birth.
+It would have been interesting to explore whether the treatment type and success would have been affected by these demographics.
+For example, radiation oncologists are looking into gender-dependent radiation therapies and are finding that female patients are more curative however they portray more side effects.
+Future research efforts should aim to address these gaps and explore the role of novel treatment modalities, biomarkers, and personalized medicine approaches in improving lung cancer prognosis.
+
+# Conclusion {#sec-conclusion}
+In conclusion, our study provides valuable insights into the complex factors influencing survival outcomes in lung cancer patients post-diagnosis.
+By elucidating the relationships between pathologic stage, treatment modalities, and survival times, we aim to inform clinical practice and guide future research endeavors.
+Although we established a general trend for those more likely to survive longer post-diagnosis, it's important to note that no two lung cancer patients are the same due to gene mutations present in the tumor and thus survival will vary person to person @Eldridge.
+Additionally, treatments are constantly improving and changing.
+Ultimately, our collective efforts must continue to focus on advancing early detection methods, refining treatment strategies, and enhancing the quality of care for lung cancer patients worldwide.
+
+“words and phrases to avoid”: enduring; landscape or evolving landscape; intricate; meticulous approach; nuanced; robust; pivotal; crafting; crucial; layered perspective.
+doing extensive/exhaustive statistical analysis/statistical endeavors on comprehensive/substantial data
 
@@
 
-# Additional data details
+### Pathologic Stage ###
+* stage is determined by tumor size, number of tumors found and where
+  the cancer has spread
+Stage I: cancer is found early, is in one lung, has not spread to LN. Localized
+Stage II: found early but has spread inside or just ouside of one lung
+  and may have spread to nearby LN on same side of the body. Regional spread.
+Stage III: cancer found in one lung and has spread to the chest and LN
+  further further away from the lungs. Regional spread.
+Stage IV: advanced cancer and has spread to the lining or fluid around 
+  the heart or lungs, or has spread to distant parts of the body. 
+  Includes having tumors in both lungs, no matter the size of the tumors.
 
-# Model details {#sec-model-details}
-we compare the posterior with the prior. This shows...
+### Tumor size ###
+TX: size cannot be assessed - too small for testing
+T1: <3cm diameter, only in lungs 
+T1a: <1cm, have not reached tissues surrounding lung/main branches of the
+  airways
+T1b: 1-2cm, have not reached tissues surrounding lung/main branches of the
+  airways
+T2: 3-5cm, grown into main bronchus/inner lining of lungs or caused lung 
+  collapse or swelling. 
+T2a: 3-4cm, grown into main airways and/or tissue around the lungs
+T2b: 4-5cm, grown into main airways and/or tissue around the lungs
+T3: 5-7cm, growth beyond primary site, into chest wall/ nearby site or
+  other tumors in the same lobe of the lung
+T4: >7cm, extensive tumor growth, invading nearby structures/organs. Growth
+  into center of chest, to the diaphragm, heart or its major blood vessels,
+  windpipe, esophagus, spine or other tumors in same lobe of the lung
 
-```{r}
-#| eval: true
-#| echo: false
-#| message: false
-#| warning: false
-#| label: fig-ppcheckandposteriorvsprior
-#| layout-ncol: 2
-#| fig-cap: "Examining how the model fits, and is affected by, the data"
-#| fig-subcap: ["Posterior prediction check", "Comparing the posterior with the prior"]
+### Presence of distant metastasis ###
+MX: not evaluated 
+M0: no distant metastasis, cancer not spread beyond site of origin/ nearby LN.
+  Localized.
+M1a: presence of metastasis to distant sites within the same lung/pleura.
+  Regional.
+M1b: metastasis to distant sites beyong lung and pleura, ex. brain, liver,
+  bones. Distant.
 
-```
+### Lymph Node (LN) involvement ###
+NX: LN cannot be assessed
+N0: no regional LN metastasis, cancer not spread to LN
+N1: involvement of regional LN
+N2: more extensive LN involvement
 
-## Diagnostics
-Is this needed?
+### Treatment type ###
+Pharmaceutical therapy 
+Radiation Therapy (RT)
 
-```{r}
-#| echo: false
-#| eval: true
-#| message: false
-#| warning: false
-#| label: fig-stanareyouokay
-#| fig-cap: "Checking the convergence of the MCMC algorithm"
-#| fig-subcap: ["Trace plot", "Rhat plot"]
-#| layout-ncol: 2
-
-```
+(AJCC, 2017)
+Stage I = T1 to T2a, N0, M0
+Stage II = T1 to T3, N0 to N1, M0
+Stage III = T1 to T4, N0 to N3, M0
+Stage IV = Any T, Any N, M1s
 

### 2024-04-22 paper/paper.qmd
Last run through

@@
 abstract: "This study investigates the impact of pathologic stage and treatment modalities on lung cancer survival post-diagnosis.
-Analysis of patient data reveals significant correlation between pathologic stage, seeking treatment and survival outcomes.
-Notably, patients at advanced stages with metastases in distant sites beyond the lung, extensive lymph node involvement and tumors with extensive growth, invading nearby structures demonstrate lower survival rates.
+Analysis of patient data reveals significant correlation between the variables that make up pathologic stage, treatment type and survival outcomes.
+Notably, patients who demonstrate lower surivival rates are those with advanced spread, extensive lymph node involvement and tumors with extensive growth into nearby structures.
+Additionally pharmaceutical therapy is shown to be more effective in prolonging patient survial times than radiation therapy.
 These findings underscore the critical importance of early detection, tailored treatment strategies and ongoing research efforts to enhance lung cancer survival rates globally."
@@
 
-Clinging to life amidst the shadows of lung cancer, where every breath becomes a battleground.
-Survival becomes not just a statistic but an interplay between several individual characteristics.
-We explore the hidden keys to defying the odds and emerging victorious against one of the deadliest adversaries of our time.
-Lung cancer is the leading cause of cancer-related deaths in the world [@Park].
-It is a disease that develops in the lining of the airways in lung tissues.
-Non-small cell lung cancer (NSCLC) is the most common type, accounting for  80-85% of all lung cancers according to the American Cancer society [@Markman].
+Surviving lung cancer is a daunting challenge, where each breath marks a battle against a formidable adversary.
+As the leading cause of cancer-related deaths worldwide (Park et al.), lung cancer casts a shadow over millions of lives, demanding a nuanced understanding of its complexities.
+In this study, we explore the interplay between individual characteristics and survival outcomes, aiming to uncover the hidden determinants of victory against this relentless disease.
+Lung cancer manifests primarily in the airway linings of lung tissues, with non-small cell lung cancer (NSCLC) representing the predominant subtype according to the American Cancer Society [@Markman].
 Staging is important for prognosis and making treatment decisions.
 Common treatments include surgery, radiation therapy and chemotherapy [@Kai].
-pathologic stage is determined by presence of nearby metastasis, lymph node involvement as well as tumor spread and size [@Markman].
-This paper investigates the relationship between lung cancer patients' survival and pathologic stage.
+Staging relies on tumor size, lymph node involvement, and the presence of distant metastasis [@Markman].
+This paper investigates the relationship between lung cancer patients' survival, pathologic stage and treatment type.
 The estimand is the median survival time in days post-diagnosis.
-We also look at whether patients decided to have treatment and if so, which method; radiation therapy or chemotherapy.
-Through analysis of a dataset made up of  981 patients in Sydney, Australia, we offer insight into the prognostic markers.
+Through analysis of a data set made up of 382 patients in Sydney, Australia, we offer insight into the prognostic markers.
 
 Tumor size is often the main determinant of stage and treatment.
-As tumor categories increase, the tumor expands, invading nearby structures [@Zhang].
-A study involving 52,287 patients diagnosed between the years 1998 and 2003 found tumor size to be an independent prognostic factor in estimating overall survival.
-The authors found that patients presenting with larger tumors predicted a worse prognosis and thus are associated with a decrease in survival.
-There is a similar relationship between extensive lymph node involvement and patient survival [@Zhang].
-Initial spread of cancer cells are localized, then become regional, involving nearby lymph nodes and the most severe cases comprises expansion to other organs such as the brain, liver and bones [@Markman].
-A study looked at five year survival rates based on the severity of spread.
-62.8% of patients with localized spread, 34.8% of patients with regional and 8% of patients with distant, advanced spread were found to survive for 5 years post diagnosis.
-More than half of these lung cancer patients have advanced spread to other organs when diagnosed [@Markman].
-Overall, it is found that patients with no regional lymph node metastases, and smaller tumors are easier to be treated and thus are associated with improved survival rates [@Zhang].
+As tumor categories increase, the tumor expands, invading nearby structures.
+A study involving 52,287 patients diagnosed between the years 1998 and 2003 found tumor size to be an independent prognostic factor in estimating overall survival and that larger tumors predicted a worse prognosis and  decreased survival [@Zhang].
+Similarly, extensive lymph node involvement  and presence of distant metastasis signifies disease progression and is associated with reduced survival rates [@Park].
+Initial spread of cancer cells are localized, then become regional, involving nearby lymph nodes and the most severe cases comprises expansion to other organs such as the brain, liver and bones.
+As cancer cells metastasize to distant sites, survival prospects diminish drastically [@Markman].
+Additionally, patients with no regional lymph node metastases, and smaller tumors are easier to be treated and thus are associated with improved survival rates [@Zhang].
+Lymph node involvement is a crucial factor in guiding treatment options.
+The authors of a study also noted that disease burden and anatomical location of the lymph nodes involved may influence the patients survival [@Kai].
 
-Presence of metastatic LN is one of the most important determinants of prognosis of NSCLC cases [@Kai].
-In the early stage, cancer has not spread to lymph nodes.
-As severity increases, lymph node metastasis sequentially spreads to more distant lymph nodes such as mediastinal and there is severe lymph node involvement [@Park].
-Lymph node involvement, also termed lymph node ratio, is a crucial factor in guiding treatment options [@Kai].
-A study made up of 97 patients with a mean age of 63 who have undergone surgery between the years 2009 and 2015 in Korea find that increased lymph node involvement is associated with a more advanced disease status and hence affiliated with prognosis [@Park].
-Another study looked at 11,341 NSCLC patients between the years 2004 to 2015, from 18 geographically diverse populations, covering approximately 28% of the population of the United States.
-These patients were treatment naive and underwent surgical resection of the tumor.
-Although 5757 patients died, the rest showed great results, with a median survival of 22 months [@Kai].
-The authors found that patients with low lymph node involvement lead to higher survival compared to patients with high lymph node ratios.
-A regression analysis revealed that lymph node ratio is an independent and significant predictor of patient survival.
-The authors also observed that disease burden and anatomical location of the lymph nodes involved may influence the patients survival [@Kai].
-
-After tumor size, LN involvement and presence of distant metastasis are categorized, the pathologic stage of the cancer is then determined [@Eldridge].
-The most valuable prognostic factor in non-small cell lung cancer is the pathologic stage [@Park] .Stage is determined by tumor size, number of tumors  and where the cancer has spread.
-Stage 1 is localized spread, stage 2 and 3 is regional spread while stage 4 is distant spread of the tumor [@Eldridge].
-Cancer stage was determined using the seventh American Joint Committee on Cancer staging system (AJCC) [@Park].
-A study done in Australia including 2119 lung cancer patients illustrated those with stage IV disease, the most advanced stage,  showed shorter survival than those at lower stages [@Denton].
+Following tumor size, LN involvement and presence of distant metastasis categorization, the pathologic stage of the cancer is then determined, the most valuable prognostic facor in non-small cell lung cancer [@Eldridge].
+Stages ranging from localized to distant spread dictate treatment strategies and prognosis, with curative options diminishing as the disease advances (see @sec-appendix for staging).
+Despite advancements in treatment modalities such as radiation therapy and chemotherapy, stage IV disease remains particularly challenging, emphasizing the importance of early detection and intervention [@Denton].
 The earlier the cancer is found, that is the lower the pathologic stage, the greater the likelihood curative radiation therapy is an effective treatment [@Eldridge].
-However, there is minimal literature looking at post-diagnosis survival rates based on pathologic stage and method of treatment.
-The extent of this disease illustrates the importance of living a healthy lifestyle, undergoing regular screening and development of improved treatment methods.
-Over the past decade, there has been great improvement of lymph node assessment in cancer patients [@Kai].
-Experts hope survival rates continue to improve with  new therapies and treatment approaches [@Markman].
 
 Radiation therapy is a local treatment, targeting the tumor directly, damaging the DNA within cancer cells with the aim of shrinking the tumor.
-The success rate of radiation therapy treatment is dependent on the location, stage of the cancer as well as individual factors.
-It is the primary treatment for early stage NSCLC and palliative treatment.
-Chemotherapy is a systemic therapy utilizing drugs to kill and inhibit cancer cell growth and is often the primary treatment for stage 3 and 4 cancers.
-NSCLC patients with signs of lymph node metastases have shown great benefit in survival when treated with chemotherapy [@Kai].
-A study looking at patients with pathologic stage 3 and 4 NSCLC showed at those who received chemotherapy survived for an average of 10.5 months whereas those that received RT only survived for 3.7 months [@Eldridge].
-Patients with early pathologic stages such as 1, 2 and 3 undergo curative treatment.
-Stage 4, the advanced stage however, is treated as palliative treatment in that the cancer is far too advanced and thus all that can be done is to minimize symptoms and try to improve quality of life.
-Stage 3b and stage 4 tumors inoperable.
-Overall, the choice between treatments is made based on assessment of the patient's condition, pathologic stage with tradeoff between providing effective treatment and optimizing quality of life.
-[@Eldridge].
+Pharmaceutical therapy on the other hand is a systemic therapy utilizing drugs to kill and inhibit cancer cell growth.
+While radiation therapy offers promise in early-stage NSCLC and palliative care, chemotherapy demonstrates efficacy, particularly in advanced stages with lymph node involvement [@Kai].
+However, treatment decisions hinge on a delicate balance between therapeutic efficacy and quality of life considerations, particularly in advanced stages where palliative care predominates [@Eldridge].
+
+There is minimal literature looking at post-diagnosis survival rates based on the variables that make up pathologic stage and as well as looking at the effectiveness of the method of treatment.
+The extent of this disease illustrates the importance of living a healthy lifestyle, undergoing regular screening and development of improved treatment methods.
+Over the past decade, there has been great improvement of lymph node assessment in cancer patients [@Kai] and experts hope survival rates continue to improve with new therapies as well as treatment approaches [@Markman].
 
@@
 Moreover, @sec-results offers visual depictions of the study's outcomes.
-Finally, in @sec-discussion, we summarize the primary findings, propose avenues for enhancement, and identify potential areas for future research.
+In @sec-discussion, we summarize the primary findings, propose avenues for enhancement, and identify potential areas for future research.
+Finally in @sec-appendix, we include a detailed definition the staging of each variable.
 
@@
 The dataset comprised several key features relevant to lung cancer prognosis, including pathologic stage, presence of distant metastasis, lymph node involvement, tumor size, and treatment type.
+
+### Days to Death
+
+The main feature analyzed in this study is the duration between the date of lung cancer diagnosis and the date of death, referred to as "days to death." This metric serves as a key indicator of patient survival and provides valuable insights into the disease trajectory and prognosis.
+By examining the distribution of survival times among lung cancer patients post-diagnosis, we aim to characterize the temporal patterns of disease progression and assess the impact of various clinical factors on survival outcomes.
+Understanding the time course from diagnosis to death is crucial for guiding treatment decisions, predicting patient outcomes, and identifying opportunities for intervention to improve survival rates.
+Through comprehensive analysis of days to death data, we seek to elucidate the factors influencing patient survival in lung cancer and contribute to the refinement of prognostic models for clinical practice.
+
+@fig-one illustrates the survival curve for lung cancer patients post-diagnosis where frequency is on the y-axis and survival time in days is on the x-axis.
+There is a clear trend of a decrease in frequency of survival as time increases.
+Mortality is most abundant 500 days after diagnosis, that is equivalent to about a year and four months.
+After about 1500 days, just over four years, the trend line plateus.
+
+@fig-one, the distribution of survival time post-diagnosis suggests that there is a high mortality rate shortly after diagnosis, thus most lung cancer patients die within one year of diagnosis (500 days).
+There is a steady decline in death and hence incline in survial up until four years (1500 days).
+The plateu after this indicates that there is a subgroup of patients who survive beyond four years.
+@sec-results explores which patients had longer survival.
 
@@
 ```
-
-### Days to Death
-
-The main feature analyzed in this study is the duration between the date of lung cancer diagnosis and the date of death, referred to as "days to death." This metric serves as a key indicator of patient survival and provides valuable insights into the disease trajectory and prognosis.
-By examining the distribution of survival times among lung cancer patients post-diagnosis, we aim to characterize the temporal patterns of disease progression and assess the impact of various clinical factors on survival outcomes.
-Understanding the time course from diagnosis to death is crucial for guiding treatment decisions, predicting patient outcomes, and identifying opportunities for intervention to improve survival rates.
-Through comprehensive analysis of days to death data, we seek to elucidate the factors influencing patient survival in lung cancer and contribute to the refinement of prognostic models for clinical practice.
-
-@fig-one illustrates the survival curve for lung cancer patients post-diagnosis where frequency is on the y-axis and survival time in days is on the x-axis.
-There is a clear trend of a decrease in frequency of survival as time increases.
-Mortality is most abundant 500 days after diagnosis, that is equivalent to about a year and four months.
-After about 1500 days, just over four years, the trend line plateus.
-
-@fig-one, the distribution of survival time post-diagnosis suggests that there is a high mortality rate shortly after diagnosis, thus most lung cancer patients die within one year of diagnosis (500 days).
-There is a steady decline in death and hence incline in survial up until four years (1500 days).
-The plateu after this indicates that there is a subgroup of patients who survive beyond four years.
-@sec-results explores which patients had longer survival.
 
@@
                                   label = paste0(round(percentage, 1), "%"))) +
-  geom_bar(stat = "identity") +
+  geom_bar(stat = "identity", width = 0.7) +  # Increased bin width
   geom_text(aes(label = paste0(round(percentage, 1), "%")), family = "patrick",
@@
   theme_minimal() +
-  scale_y_continuous(breaks = seq(0, max(treatment_type_counts$percentage), by = 5)) +  
-  theme(text = element_text(family = "patrick", size = 14))
+  scale_y_continuous(limits = c(0, 55), breaks = seq(0, 55, by = 5)) +  # Adjusted y-axis range
+  theme(text = element_text(family = "patrick", size = 14),
+        axis.text.x = element_blank())  # Remove x-axis tick mark labels
 
 ```
+
+\newpage
 
@@
   theme_minimal() +
-  theme(text = element_text(family = "patrick", size = 14),  # Set font to Patrick Hand and increase the size
-        legend.position = "bottom", legend.text = element_text(size = 12),
+  theme(text = element_text(family = "patrick", size = 13),  # Set font to Patrick Hand and increase the size
+        legend.position = "bottom", legend.text = element_text(size = 11),
         plot.caption = element_text(hjust = 0, size = 14)) +  # Adjust legend position and text size
@@
 ```
+
+\newpage
 
@@
   theme_minimal() +
-  theme(text = element_text(family = "patrick", size = 14), 
-        legend.position = "bottom", legend.text = element_text(size = 12),
+  theme(text = element_text(family = "patrick", size = 13), 
+        legend.position = "bottom", legend.text = element_text(size = 11),
         plot.caption = element_text(hjust = 0, size = 14)) +  
@@
 ```
+
+\newpage
+
 ## Treatment Type by Lymph Node Involvement 
@@
 #| label: fig-six
-#| fig-caption: The scatterplot demonstrates the relationship between lymph node involvement, time to death, and treatment type in lung cancer patients post-diagnosis.
+#| fig-cap: The scatterplot demonstrates the relationship between lymph node involvement, time to death, and treatment type in lung cancer patients post-diagnosis.
 
@@
 ```
+
+\newpage
+
 ## Treatment Type by Tumor Size
@@
 #| label: fig-seven
-#| fig-caption: The scatterplot portrays the relationship between tumor size, time to death, and treatment type in lung cancer patients post-diagnosis.
+#| fig-cap: "The scatterplot portrays the relationship between tumor size, time to death, and treatment type in lung cancer patients post-diagnosis."
 
@@
 
+\newpage
+
 # Discussion {#sec-discussion}
@@
 ## Survival Dynamics {#sec-first-point}
-Eldridge explains that the severity of lung cancer is staged by pathologic stage which is determined by all three presence of distant metastasis, lymph node involvement and tumor size [@Eldrige].
+Eldridge explains that the severity of lung cancer is staged by pathologic stage which is determined by all three presence of distant metastasis, lymph node involvement and tumor size [@Eldridge].
 Previous research found that patients with larger tumors and more extensive lymph node involvement are associated with a decrease in survival [@Zhang].
@@
 ## Future Directions
-The dataset we used also offered information regarding demographics such as race, gender and year of birth.
+The data set we used also offered information regarding demographics such as race, gender and year of birth.
 It would have been interesting to explore whether the treatment type and success would have been affected by these demographics.
@@
 By elucidating the relationships between pathologic stage, treatment modalities, and survival times, we aim to inform clinical practice and guide future research endeavors.
-Although we established a general trend for those more likely to survive longer post-diagnosis, it's important to note that no two lung cancer patients are the same due to gene mutations present in the tumor and thus survival will vary person to person @Eldridge.
+Although we established a general trend for those more likely to survive longer post-diagnosis, it's important to note that no two lung cancer patients are the same due to gene mutations present in the tumor and thus survival will vary person to person [@Eldridge].
 Additionally, treatments are constantly improving and changing.
 Ultimately, our collective efforts must continue to focus on advancing early detection methods, refining treatment strategies, and enhancing the quality of care for lung cancer patients worldwide.
-
-“words and phrases to avoid”: enduring; landscape or evolving landscape; intricate; meticulous approach; nuanced; robust; pivotal; crafting; crucial; layered perspective.
-doing extensive/exhaustive statistical analysis/statistical endeavors on comprehensive/substantial data
 
@@
 ### Pathologic Stage ###
-* stage is determined by tumor size, number of tumors found and where
-  the cancer has spread
+Note: Stage is determined by tumor size, number of tumors found and where
+  the cancer has spread. Includes having tumors in both lungs, no matter the size of the tumors.
+
 Stage I: cancer is found early, is in one lung, has not spread to LN. Localized
-Stage II: found early but has spread inside or just ouside of one lung
+
+Stage II: found early but has spread inside or just outside of one lung
   and may have spread to nearby LN on same side of the body. Regional spread.
+
 Stage III: cancer found in one lung and has spread to the chest and LN
   further further away from the lungs. Regional spread.
+
 Stage IV: advanced cancer and has spread to the lining or fluid around 
   the heart or lungs, or has spread to distant parts of the body.
-  Includes having tumors in both lungs, no matter the size of the tumors.
 
@@
 TX: size cannot be assessed - too small for testing
+
 T1: <3cm diameter, only in lungs
+
 T1a: <1cm, have not reached tissues surrounding lung/main branches of the
   airways
+
 T1b: 1-2cm, have not reached tissues surrounding lung/main branches of the
   airways
+
 T2: 3-5cm, grown into main bronchus/inner lining of lungs or caused lung 
   collapse or swelling.
+
 T2a: 3-4cm, grown into main airways and/or tissue around the lungs
+
 T2b: 4-5cm, grown into main airways and/or tissue around the lungs
+
 T3: 5-7cm, growth beyond primary site, into chest wall/ nearby site or
   other tumors in the same lobe of the lung
+
 T4: >7cm, extensive tumor growth, invading nearby structures/organs. Growth
@@
 MX: not evaluated
-M0: no distant metastasis, cancer not spread beyond site of origin/ nearby LN.
-  Localized.
-M1a: presence of metastasis to distant sites within the same lung/pleura.
-  Regional.
-M1b: metastasis to distant sites beyong lung and pleura, ex. brain, liver,
-  bones. Distant.
+
+M0: no distant metastasis, cancer not spread beyond site of origin/ nearby LN. Localized.
+
+M1a: presence of metastasis to distant sites within the same lung/pleura. Regional.
+
+M1b: metastasis to distant sites beyong lung and pleura, ex. brain, liver, bones. Distant.
 
@@
 NX: LN cannot be assessed
+
 N0: no regional LN metastasis, cancer not spread to LN
+
 N1: involvement of regional LN
+
 N2: more extensive LN involvement
@@
 Pharmaceutical therapy
-Radiation Therapy (RT)
 
-(AJCC, 2017)
-Stage I = T1 to T2a, N0, M0
-Stage II = T1 to T3, N0 to N1, M0
-Stage III = T1 to T4, N0 to N3, M0
-Stage IV = Any T, Any N, M1s
+Radiation Therapy
 

### 2024-04-28 paper/paper.qmd

@@
 Analysis of patient data reveals significant correlation between the variables that make up pathologic stage, treatment type and survival outcomes.
-Notably, patients who demonstrate lower surivival rates are those with advanced spread, extensive lymph node involvement and tumors with extensive growth into nearby structures.
-Additionally pharmaceutical therapy is shown to be more effective in prolonging patient survial times than radiation therapy.
+Notably, patients who demonstrate lower survival rates are those with advanced spread, extensive lymph node involvement and tumors with extensive growth into nearby structures.
+Additionally pharmaceutical therapy is shown to be more effective in prolonging patient survival times than radiation therapy.
 These findings underscore the critical importance of early detection, tailored treatment strategies and ongoing research efforts to enhance lung cancer survival rates globally."
@@
 
-Following tumor size, LN involvement and presence of distant metastasis categorization, the pathologic stage of the cancer is then determined, the most valuable prognostic facor in non-small cell lung cancer [@Eldridge].
+Following tumor size, LN involvement and presence of distant metastasis categorization, the pathologic stage of the cancer is then determined, the most valuable prognostic factor in non-small cell lung cancer [@Eldridge].
 Stages ranging from localized to distant spread dictate treatment strategies and prognosis, with curative options diminishing as the disease advances (see @sec-appendix for staging).
@@
 This left us with a sample of 382 lung cancer patients.
-Additionally, we converted the days to death column to numeric.Tests where included to ensure accuracy, reliability and validity of the datset for subsequent analysis and interpretation.
+Additionally, we converted the days to death column to numeric.
+Tests where included to ensure accuracy, reliability and validity of the dataset for subsequent analysis and interpretation.
 
@@
 Mortality is most abundant 500 days after diagnosis, that is equivalent to about a year and four months.
-After about 1500 days, just over four years, the trend line plateus.
+After about 1500 days, just over four years, the trend line plateaus.
 
 @fig-one, the distribution of survival time post-diagnosis suggests that there is a high mortality rate shortly after diagnosis, thus most lung cancer patients die within one year of diagnosis (500 days).
-There is a steady decline in death and hence incline in survial up until four years (1500 days).
-The plateu after this indicates that there is a subgroup of patients who survive beyond four years.
+There is a steady decline in death and hence incline in survival up until four years (1500 days).
+The plateau after this indicates that there is a subgroup of patients who survive beyond four years.
 @sec-results explores which patients had longer survival.
@@
 Pathologic stage, a critical determinant of lung cancer prognosis, was classified according to the TNM (tumor, lymph node, metastasis) Cancer Staging.
-As depicted in @fig-two, illustrating the the distribution of pathologic stages with pathologic stage on the x-axis and percentage on the y-axis.
-The distribution of patients varied across different stages, with the highest proportion of patients diagnosed at StageIB.
+As depicted in @fig-two, illustrating the distribution of pathologic stages with pathologic stage on the x-axis and percentage on the y-axis.
+The distribution of patients varied across different stages, with the highest proportion of patients diagnosed at Stage IB.
 At this stage, the cancer is still found early however the tumor has regional spread to nearby lymph nodes.
@@
 #| label: fig-three
-#| fig-cap: The histogram shows the percentage of lung cancer patients by treamtent type.
+#| fig-cap: The histogram shows the percentage of lung cancer patients by treatment type.
 
@@
 In this section, we aim to predict the survival outcomes of lung cancer patients post-diagnosis with a linear regression model framework.
-We consider several predictors including pathologic stage, lymph node involvement, presence of distant metastasis, tumor size, and treatment type.We specify the model and subsequently justify its appropriateness for our analysis.
+We consider several predictors including pathologic stage, lymph node involvement, presence of distant metastasis, tumor size, and treatment type.
+We specify the model and subsequently justify its appropriateness for our analysis.
 
@@
 
-Linear regression models are most appropritate in predicting continuous outcomes.
+Linear regression models are most appropriate in predicting continuous outcomes.
 As survival time is continuous, this model allows us to quantify the relationships between these predictors and survival outcomes, providing valuable insights into the factors influencing the prognosis of lung cancer patients.
@@
 
-Out variable of interest is survival time in lung cancer patient afte they have been diagnosed
+Our variable of interest is survival time in lung cancer patient after they have been diagnosed
 
@@
 Stage III: cancer found in one lung and has spread to the chest and LN
-  further further away from the lungs. Regional spread.
+  further away from the lungs. Regional spread.
 
@@
 
-M1b: metastasis to distant sites beyong lung and pleura, ex. brain, liver, bones. Distant.
+M1b: metastasis to distant sites beyond lung and pleura, ex. brain, liver, bones. Distant.
 
